### 2010-07-03 paper.tex
Just getting started; drawn kD-tree, title, authors, etc.

\documentclass[prd,preprint]{revtex4}

\usepackage{amsmath}
\usepackage{hyperref}

\begin{document}
\title{An Efficient Interpolation Technique for Jump Proposals in
  Reversible-Jump Markov Chain Monte Carlo Calculations}

\date{\today}

\author{Will M. Farr}
\email{w-farr@northwestern.edu}

\author{Ilya Mandel}
\email{ilyamandel@chgk.info}

\affiliation{Northwestern University Center for Interdisciplinary
  Research and Education in Astrophysics}

\begin{abstract}
  We do something.
\end{abstract}

\maketitle

\end{document}

### 2010-12-06 paper.tex
Bibliography begun.
Outline with Ilya.

@@
 \usepackage{hyperref}
+
+\newcommand{\vtheta}{\vec{\theta}}
+
+\bibliographystyle{h-physrev}
 
@@
 
+\affiliation{Northwestern University Center for Interdisciplinary
+  Research and Education in Astrophysics}
+
 \author{Ilya Mandel}
@@
 
-\affiliation{Northwestern University Center for Interdisciplinary
-  Research and Education in Astrophysics}
+\affiliation{MIT Kavli Institute}
 
 \begin{abstract}
-  We do something.
+  Reversible-jump Markov chain monte carlo (RJMCMC) is an extremely
+  powerful technique for performing Bayesian model selection, but it
+  suffers from a fundamental difficulty: it requires jumps between
+  model parameter spaces, but cannot retain a memory of the favored
+  locations in more than one parameter space at a time.  Thus, a naive
+  jump between parameter spaces is unlikely to be accepted in the MCMC
+  algorithm and convergence is correspondingly slow.  Here we
+  demonstrate an interpolation technique that uses samples from
+  single-model MCMCs to propose inter-model jumps from an
+  approximation to the single-model posterior of the target parameter
+  space.  The interpolation technique, based on a kD-tree
+  datastructure, is adaptive and efficient in arbitrary dimensions.
+  We show that our technique leads to dramatically improved convergene
+  over naive jumps in an RJMCMC, and compare it to other proposals in
+  the literature to improve the convergence of RJMCMCs.  We also
+  discuss the use of the same interpolation technique in two other
+  contexts: as a convergence test for a single-model MCMC and as a way
+  to construct efficient ``global'' proposal distributions for
+  single-model MCMCs without prior knowledge of the structure of the
+  posterior distribution.
 \end{abstract}
@@
 
+\section{Introduction}
+
+\section{Reversible Jump MCMC}
+
+Reversible jump MCMC (RJMCMC) \cite{Green1995} is a technique for
+Bayesian model comparison.  Consider a set of models indexed by an
+integer, $i$: $\{M_i | i = 1, 2, \ldots \}$.  Each model has some
+continuous parameters, $\vtheta_i$; given the model and its
+parameters, we can make a prediction about the likelihood of observing
+some experimental data, $d$: $L(d|\vtheta_i, M_i)$.  When some data
+have been observed in an experiment, Bayes' rule gives us a way to
+compute the posterior probability distribution for the model
+parameters implied by the data:
+\begin{equation}
+  p(M_i, \vtheta_i | d) = \frac{L(d|\vtheta_i, M_i) p(\vtheta_i|M_i) p(M_i)}{p(d)},
+\end{equation}
+where $p(M_i, \vtheta_i |d)$ is the posterior distribution for the
+model parameters $\vtheta_i$ implied by the data in the context of
+model $M_i$, $p(\vtheta_i|M_i)$ and $p(M_i)$ are the prior
+probabilities of the model parameters and the model itself that
+represent our beliefs before accumulating any of the data $d$, and
+$p(d)$ is an overall normalizing constant that ensures that $p(M_i,
+\vtheta_i|d)$ is properly normalized as a probability distribution on
+the $M_i$ and $\vtheta_i$.  This implies that 
+\begin{equation}
+  p(d) = \sum_i \int_{V_i} d\vtheta_i p(M_i, \vtheta_i|d) \equiv
+  \sum_i p(M_i|d),
+\end{equation}
+where $V_i$ is the parameter space volume in model $M_i$, and we have
+defined the integrated model posterior, 
+\begin{equation}
+  p(M_i|d) \equiv \int_{V_i} d\vtheta_i p(M_i, \vtheta_i | d),
+\end{equation}
+which represents the total posterior probability in the model $M_i$
+parameter space.
+
+\section{kD Trees and Interpolation}
+
+\section{RJMCMC Efficiency}
+
+\section{Examples}
+
+\section{Other Uses}
+
+\section{Conclusion}
+
+\nocite{Littenberg2009}
+
+\bibliography{paper}
+
 \end{document}

### 2010-12-08 paper.tex
.

@@
 
-\section{Examples}
+\section{Example}
 
-\section{Other Uses}
+BH mass paper [still not sure how to present this without repetition...  ]
 
-\section{Conclusion}
+\section{Conclusion/Discussion}
+
+...
+
+\subsection{Other Uses}
+The efficient interpolation of a PDF via a kD-tree described in this paper can be extremely useful in other contexts, beyond a reversible-jump MCMC.
+We suggest two examples below: generating efficient jump proposal distributions and convergence tests?
+
+...
+
+Mention nested sampling?
 

### 2010-12-09 paper.tex
Working on Section 2...

--- 2010-07-03 paper.tex
@@
 \usepackage{hyperref}
+\usepackage{color}
+
+\newcommand{\vtheta}{\vec{\theta}}
+
+\newcommand{\be}{\begin{equation}}
+\newcommand{\ee}{\end{equation}}
+\newcommand{\bel}[1]{\begin{equation}\label{#1}}
+\newcommand{\ba}{\begin{eqnarray}}
+\newcommand{\ea}{\end{eqnarray}}
+\newcommand{\bal}[1]{\begin{eqnarray}\label{#1}}
+
+\def\ilya{\textcolor{red}}
+\def\will{\textcolor{blue}}
+
+\bibliographystyle{h-physrev}
 
@@
 
-\affiliation{Northwestern University Center for Interdisciplinary
-  Research and Education in Astrophysics}
+\affiliation{MIT Kavli Institute; and University of Birmingham, UK}
 
 \begin{abstract}
-  We do something.
+\ilya{Add physics/astro motivation}
+  Reversible-jump Markov chain monte carlo (RJMCMC) is an extremely
+  powerful technique for performing Bayesian model selection, but it
+  suffers from a fundamental difficulty: it requires jumps between
+  model parameter spaces, but cannot retain a memory of the favored
+  locations in more than one parameter space at a time.  Thus, a naive
+  jump between parameter spaces is unlikely to be accepted in the MCMC
+  algorithm and convergence is correspondingly slow.  Here we
+  demonstrate an interpolation technique that uses samples from
+  single-model MCMCs to propose inter-model jumps from an
+  approximation to the single-model posterior of the target parameter
+  space.  The interpolation technique, based on a kD-tree
+  data structure, is adaptive and efficient in arbitrary dimensions.
+  We show that our technique leads to dramatically improved convergence
+  over naive jumps in an RJMCMC, and compare it to other proposals in
+  the literature to improve the convergence of RJMCMCs.  We also
+  discuss the use of the same interpolation technique in two other
+  contexts: as a convergence test for a single-model MCMC and as a way
+  to construct efficient ``global'' proposal distributions for
+  single-model MCMCs without prior knowledge of the structure of the
+  posterior distribution.
 \end{abstract}
@@
 
+\section{Introduction}
+
+\section{Reversible Jump MCMC}
+
+Reversible jump MCMC (RJMCMC) \cite{Green1995} is a technique for Bayesian model comparison.
+Below, we give a very brief introduction to Bayesian analysis; describe a standard MCMC; and introduce RJMCMC.
+
+\subsection{Bayesian analysis}
+
+Consider an observed data set $d$ and a set of competing models for the data, indexed by an integer $i$: $\{M_i | i = 1, 2, \ldots \}$.
+Each model has some continuous parameters, $\vtheta_i$; given the model and its parameters, we can make a prediction about the likelihood of observing
+the experimental data: $L(d|\vtheta_i, M_i)$.  Within the framework of each model, Bayes' rule gives us a way to compute the posterior probability distribution function (PDF) for the model
+parameters implied by the data:
+\begin{equation}
+  p(\vtheta_i | d, M_i) = \frac{L(d|\vtheta_i, M_i) p(\vtheta_i|M_i)}{p(d|M_i)},
+\end{equation}
+where $p(M_i, \vtheta_i |d)$ is the posterior distribution for the
+model parameters $\vtheta_i$ implied by the data in the context of
+model $M_i$, $p(\vtheta_i|M_i)$ is the prior
+probability of the model parameters that 
+represents our beliefs before accumulating any of the data $d$, and
+$p(d)$ is an overall normalizing constant that ensures that $p(\vtheta_i|d,M_i)$ is properly normalized as a probability distribution on
+the $\vtheta_i$.  This implies that the evidence is equal to
+\bel{evidence}
+  p(d|M_i) = \int_{V_i} d\vtheta_i L(d|\vtheta_i, M_i) p(\vtheta_i|M_i),
+\end{equation}
+where $V_i$ is the parameter space volume in model $M_i$.  We are interested in the evidence for model $M_i$ given the data, $p(M_i|d)$, which is again given by Bayes' rule as
+\begin{equation}
+p(M_i|d) = \frac{p(d|M_i) p(M_i)}{p(d)},
+\end{equation},
+where $p(M_i)$ is our a priori belief in model $M_i$ and $p(d)$ is a normalizing constant, 
+\be
+p(d)=\sum_i p(d|M_i) p(M_i).
+\ee
+
+When selecting among alternative models, we are interested in finding the model with the highest evidence $p(M_i|d)$.
+However, attempts to directly compute the evidence by performing the integration in Eq.~(\ref{bel}) are generally very difficult in a multi-dimensional, multi-modal parameter space when the likelihood has to be evaluated numerically.
+In particular, a grid-based integral quickly becomes computationally unfeasible as the dimensionality of $\vtheta$ exceeds a few.
+The parameter space must typically be explored in a stochastic manner before the evidence integral can be computed.
+Although several stochastic parameter-exploration techniques are focused directly on evidence computation (e.g., nested sampling and MultiNest \ilya {Add references}), we frequently want to compute the posterior PDFs within each model along with the evidences for the various models, and the Markov chain Monte Carlo is one of the most widely used tools for this purpose.
+
+\subsection{MCMC}
+
+MCMC methods produce a Markov chain (or sequence) of parameter
+samples, $\{ \vtheta^(i) \, | \, i = 1, \ldots \}$, such that a particular
+parameter set, $\vtheta$, appears in the sequence with a frequency
+equal to its probability according to a posterior, $p(\vtheta|d)$.  A
+Markov chain has the property that the transition probability from one
+element to the next, $p(\vtheta_i \to \vtheta_{i+1})$, depends only on
+the value of $\vtheta_i$, not on any previous values in the chain.
+
+One way to produce a sequence of MCMC samples is via the following
+algorithm, first proposed by \citet{Metropolis1953} and used widely in
+the physical sciences thereafter:
+\begin{enumerate}
+  \item Begin with the current sample, $\vtheta_i$.
+  \item Propose a new sample, $\vtheta_p$, by drawing randomly from a
+    ``jump proposal distribution'' with probability $Q(\vtheta_i \to
+    \vtheta_p)$.  Note that $Q(\vtheta_i \to \vtheta_p)$ can depend on
+    the current parameters, $\vtheta_i$, and any other ``constant''
+    data, but cannot examine the history of the chain beyond the most
+    recent point.  This is necessary to preserve the Markovian
+    property of the chain.
+  \item Compute the ``acceptance'' probability,
+    \begin{equation}
+      \label{eq:paccept}
+      p_{\textnormal{accept}} \equiv
+      \frac{p(\vtheta_p|d)}{p(\vtheta_i|d)} \frac{Q(\vtheta_p \to
+        \vtheta_i)}{Q(\vtheta_i \to \vtheta_p)}
+    \end{equation}
+  \item With probability $\min(1,p_{\textnormal{accept}})$ ``accept''
+    the proposed $\vtheta_p$, setting $\vtheta_{i+1} = \vtheta_p$;
+    otherwise set $\vtheta_{i+1} = \vtheta_i$.
+\end{enumerate}
+This algorithm is more likely to accept a proposed jump when it
+increases the posterior (the first factor in Equation
+\eqref{eq:paccept}) and when it is to a location in parameter space
+from which it is easy to return (the second factor in Equation
+\eqref{eq:paccept}); the combination of these influences in Equation
+\eqref{eq:paccept} ensures that the equilibrium distribution of the
+chain is $p(\vtheta|d)$.  As $i \to \infty$ the samples $\vtheta_i$
+are distributed according to $p(\vtheta|d)$.
+
+In practice the number of samples required before the chain
+appropriately samples $p(\vtheta|d)$ depends strongly on the jump
+proposal distribution; proposal distributions that often propose jumps
+toward or within regions of large $p(\vtheta|d)$ can be very
+efficient, while poor proposal distributions can require prohibitively
+large numbers of samples before convergence.
+
+There is no foolproof test for the convergence of a chain.  In this
+work, we test the convergence of our chains by comparing the
+statistics calculated from the entire chain to statistics calculated
+from only the first half of the chain; when the chain has converged,
+the two calculations agree.  This is a necessary, but not sufficient,
+condition for convergence.
+
+We begin the chain at an arbitrary point in parameter space; this is
+equivalent to taking a finite section of an infinite chain that begins
+with the chosen point.  Every point in parameter space occurs in an
+infinite chain, and no section of the chain is better than any other,
+so a sufficiently long, but finite, section of the infinite chain
+chosen in this manner can be representative of the statistics of the
+chain as a whole.  However, because consecutive samples in a chain are
+correlated with each other, the beginning of our finite chain has a
+``memory'' of the starting point; we discard enough points at the
+beginning of the finite chain that we can be confident it does not
+retain a memory of the arbitrary starting point.  The points discarded
+in this way are commonly called ``burn-in'' points.
+
+\subsection{RJMCMC}
+
+Consider the problem of model selection among a set of models, and the
+``super-model'' that encompasses all the models under consideration.
+The parameter space of the super-model consists of a discrete
+parameter that identifies the choice of model, $M_i$, and the
+continuous parameters appropriate for this model, $\vtheta_i$.  We
+denote a point in the super-model parameter space by $\{M_i,
+\vtheta_i\}$; each such point is a statement that, e.g., ``the
+underlying mass distribution for black holes in the galaxy is a
+Gaussian, with parameters $\mu$ and $\sigma$,'' or ``the underlying
+mass distribution for black holes in the galaxy is a triple-bin
+histogram with parameters $w_1$, $w_2$, $w_3$, and $w_4$,'' or ....
+To compare models, we are interested in the quantity (see Equation
+\eqref{eq:model-posterior-def})
+\begin{equation}
+  p(M_i|d) = \int d\vtheta_i p(M_i, \vtheta_i|d).
+\end{equation}
+If we perform an MCMC in the super-model parameter space, then we
+obtain a chain of samples $\{M_i, \vtheta_i \, | \, i = 1, \ldots\}$
+distributed in parameter space with density $p(M_i,\vtheta_i|d)
+d\vtheta_i$ and we can estimate the integral as
+\begin{equation}
+    p(M_i|d) = \int d\vtheta_i p(M_i, \vtheta_i|d) \approx \frac{N_i}{N},
+\end{equation}
+where $N_i$ is the number of samples in the chain lying in the
+parameter space of model $M_i$ and $N$ is the total number of samples
+in the chain.  The fraction of samples lying in the parameter space of
+model $M_i$ gives the probability of that model relative to the other
+models under consideration.
+
+To perform the MCMC in the super-model parameter space, we must
+propose jumps not only between points in a particular model's
+parameter space, but also between the parameter spaces of different
+models.  For this MCMC to be efficient, proposed jumps into a model
+from another should favor regions with large posterior; when the
+posterior is highly-peaked in a small region of parameter space,
+proposed jumps outside this region are unlikely to be accepted, and
+the reversible-jump MCMC samples will require a very long chain to
+properly sample the ``super-model'' posterior.
+
+We can exploit the information we have from single-model MCMCs to
+generate efficient jump proposal distributions for our reversible jump
+MCMC.  We would like to propose jumps that roughly follow the
+distribution of samples in the single-model MCMCs.
+
+\section{kD Trees and Interpolation}
+
+Mention example of bricks from \cite{Littenberg2009}.
+
+\section{RJMCMC Efficiency}
+
+\section{Example}
+
+BH mass paper [still not sure how to present this without repetition...  ]
+
+\section{Conclusion/Discussion}
+
+...
+
+\subsection{Other Uses}
+The efficient interpolation of a PDF via a kD-tree described in this paper can be extremely useful in other contexts, beyond a reversible-jump MCMC.
+We suggest two examples below: generating efficient jump proposal distributions and convergence tests?
+
+...
+
+Mention nested sampling?
+
+\nocite{Littenberg2009}
+
+\bibliography{paper}
+
 \end{document}

### 2010-12-09 paper.tex
Done with 2.1 and 2.2.

@@
 
-MCMC methods produce a Markov chain (or sequence) of parameter
-samples, $\{ \vtheta^(i) \, | \, i = 1, \ldots \}$, such that a particular
-parameter set, $\vtheta$, appears in the sequence with a frequency
-equal to its probability according to a posterior, $p(\vtheta|d)$.  A
-Markov chain has the property that the transition probability from one
-element to the next, $p(\vtheta_i \to \vtheta_{i+1})$, depends only on
-the value of $\vtheta_i$, not on any previous values in the chain.
+A Markov chain Monte Carlo produces a set of samples $\{ \vtheta^(i) \, | \, i = 1, \ldots \}$ from the model parameter space that are sampled according to the posterior, meaning that, in the limit that the chain length tends to infinity, the relative frequency with which a given set of parameter appears in the chain is proportional to the desired posterior, $p(\vtheta|d,M)$.
+Therefore, the output of an MCMC can be directly interpreted as the posterior PDF over the full parameter space, while PDFs for individual parameters can be obtained through trivially marginalizing over the uninteresting parameters, simply by considering the distribution of the parameter of interest in the MCMC samples.
 
-One way to produce a sequence of MCMC samples is via the following
-algorithm, first proposed by \citet{Metropolis1953} and used widely in
-the physical sciences thereafter:
+A Markov chain has the property that the probability distribution of the next state can depend only on the current state, not on the past history:
+\be
+p(\vtheta^(i+1))=\int_{V} d\vtheta^{i} p(\vtheta^{i} \to \vtheta^{i+1}) p(\vtheta^{i+1}),
+\ee
+where $p(\vtheta^{i} \to \vtheta^{i+1})$ depends only on $\vtheta^{i}$ and $\vtheta^{i+1}$. 
+An additional requirement for an MCMC arises from the fact that the desired distribution is the equilibrium distribution.
+In other words, if we assume that state $(i)$ of the chain is sampled from the desired PDF, $p(\vtheta^{i})=p(\vtheta|d,M)$ , then the next state $(i+1)$ must be sampled from the PDF as well, so that $p(\vtheta^{i+1})=p(\vtheta|d,M)$.
+
+One way to produce such a sequence of samples is via the Metropolis-Hastings algorithm, first proposed by Metropolis et al.~in 1953 \cite{Metropolis:1953}, and later generalized by Hastings:
 \begin{enumerate}
-  \item Begin with the current sample, $\vtheta_i$.
-  \item Propose a new sample, $\vtheta_p$, by drawing randomly from a
-    ``jump proposal distribution'' with probability $Q(\vtheta_i \to
-    \vtheta_p)$.  Note that $Q(\vtheta_i \to \vtheta_p)$ can depend on
-    the current parameters, $\vtheta_i$, and any other ``constant''
-    data, but cannot examine the history of the chain beyond the most
-    recent point.  This is necessary to preserve the Markovian
-    property of the chain.
-  \item Compute the ``acceptance'' probability,
-    \begin{equation}
-      \label{eq:paccept}
-      p_{\textnormal{accept}} \equiv
-      \frac{p(\vtheta_p|d)}{p(\vtheta_i|d)} \frac{Q(\vtheta_p \to
-        \vtheta_i)}{Q(\vtheta_i \to \vtheta_p)}
-    \end{equation}
-  \item With probability $\min(1,p_{\textnormal{accept}})$ ``accept''
-    the proposed $\vtheta_p$, setting $\vtheta_{i+1} = \vtheta_p$;
-    otherwise set $\vtheta_{i+1} = \vtheta_i$.
-\end{enumerate}
-This algorithm is more likely to accept a proposed jump when it
-increases the posterior (the first factor in Equation
-\eqref{eq:paccept}) and when it is to a location in parameter space
-from which it is easy to return (the second factor in Equation
-\eqref{eq:paccept}); the combination of these influences in Equation
-\eqref{eq:paccept} ensures that the equilibrium distribution of the
-chain is $p(\vtheta|d)$.  As $i \to \infty$ the samples $\vtheta_i$
-are distributed according to $p(\vtheta|d)$.
+  \item Given a current state $\vtheta^(i)$, propose the next state $\vtheta^p$ by drawing from a jump proposal distribution with probability $Q(\vtheta^(i) \to \vtheta^p)$.  
+  \item Compute the probability of accepting the proposed jump as
+\be
+p_{\textnormal{accept}}  \equiv \min\Bigl(1,  frac{p(\vtheta^p|d)}{p(\vtheta^(i)|d)} \frac{Q(\vtheta^p \to
+        \vtheta^(i))}{Q(\vtheta^(i) \to \vtheta^p)} \Bigr).
+\ee
+\item Pick a uniform random number $\alpha \in [0,1]$.
+If $\alpha<  p_{\textnormal{accept}}$, accept the proposed jump, setting $\vtheta^(i+1)=\vtheta^p$.
+Otherwise, reject the jump, and remain at the same location in parameter space for the next step, $\vtheta^(i+1)=\vtheta^(i)$.
+ \end{itemize}
 
-In practice the number of samples required before the chain
-appropriately samples $p(\vtheta|d)$ depends strongly on the jump
-proposal distribution; proposal distributions that often propose jumps
-toward or within regions of large $p(\vtheta|d)$ can be very
-efficient, while poor proposal distributions can require prohibitively
-large numbers of samples before convergence.
+This jump proposal distribution $Q(\vtheta^(i) \to \vtheta^p)$ can depend on the parameters of the current state $\vtheta^(i)$, but not on the past history.
+It must also allow any state within the allowed prior volume to be reachable by the MCMC.
+Beyond these conditions, there is obviously a great deal of freedom in choosing the jump proposal distribution.
+This is the most important choice in the MCMC, as it determines the sampling efficiency of the algorithm, i.e., the necessary length of the chain before it converges to the posterior PDF.
+Creating an efficient jump proposal distribution requires an understanding of the structure of the parameter space which may not be available until the PDFs are found, creating a Catch-22; one possibility for resolving this infinite loop is described in Section \ref{sec:examples}.
 
-There is no foolproof test for the convergence of a chain.  In this
-work, we test the convergence of our chains by comparing the
-statistics calculated from the entire chain to statistics calculated
-from only the first half of the chain; when the chain has converged,
-the two calculations agree.  This is a necessary, but not sufficient,
-condition for convergence.
+It should be noted that although an MCMC who jump acceptance criterium obeys detailed balance (as the Metropolis-Hastings algorithm does) must eventually converge to the desired distribution, there is no way to guarantee convergence in a fixed number of steps or to test whether a chain has converged in a foolproof manner.
+For example, MCMC chains can get stuck on local maxima, and if the autocorrelation length of the chain represents a substantial fraction of the total number of samples, the effective sample size may be too small to accurately represent the relative sizes of the modes in the PDF (however, see Section \ref{sec:examples} for one intriguing suggestion for remedying this issue).
 
-We begin the chain at an arbitrary point in parameter space; this is
-equivalent to taking a finite section of an infinite chain that begins
-with the chosen point.  Every point in parameter space occurs in an
-infinite chain, and no section of the chain is better than any other,
-so a sufficiently long, but finite, section of the infinite chain
-chosen in this manner can be representative of the statistics of the
-chain as a whole.  However, because consecutive samples in a chain are
-correlated with each other, the beginning of our finite chain has a
-``memory'' of the starting point; we discard enough points at the
-beginning of the finite chain that we can be confident it does not
-retain a memory of the arbitrary starting point.  The points discarded
-in this way are commonly called ``burn-in'' points.
+Finally, we note that, in practice, the randomly chosen initial starting point of the MCMC may be in a particularly unlikely location in the parameter space.
+Because jumps are frequently local, we will generally want to ignore the first few autocorrelation lengths worth of points in a finite-size chain to avoid biases in the recovered posterior PDF due to the choice of the initial location.
+The points thus discarded are referred to as "burn-in" points.
 
 \subsection{RJMCMC}
+
+The samples produced during a Markov Chain Monte Carlo can be used to directly perform a Monte Carlo evidence integral.
+This results in a harmonic mean estimator for the evidence, which suffers from a large variance and a possible bias.
+Additional techniques for the direct integration of evidence are described in
+
+An alternative approach to model selection  models
 
@@
 
-\section{Example}
+\section{Examples and other uses} \label{sec:examples}
 
 BH mass paper [still not sure how to present this without repetition...  ]
+
+GW parameter estimation
 

### 2010-12-09 paper.tex
Finished Section 2.

@@
 
-\def\ilya{\textcolor{red}}
-\def\will{\textcolor{blue}}
+\newcommand{\ilya}[1]{{\color{red} \bf Ilya: #1}}
+\newcommand{\will}[1]{{\color{blue} \bf Will: #1}}
 
@@
 
-Reversible jump MCMC (RJMCMC) \cite{Green1995} is a technique for Bayesian model comparison.
+Reversible jump Markov chain Monte Carlo (RJMCMC) \cite{Green1995} is a technique for Bayesian model comparison.
 Below, we give a very brief introduction to Bayesian analysis; describe a standard MCMC; and introduce RJMCMC.
@@
 p(M_i|d) = \frac{p(d|M_i) p(M_i)}{p(d)},
-\end{equation},
+\end{equation}
 where $p(M_i)$ is our a priori belief in model $M_i$ and $p(d)$ is a normalizing constant, 
@@
 
-A Markov chain Monte Carlo produces a set of samples $\{ \vtheta^(i) \, | \, i = 1, \ldots \}$ from the model parameter space that are sampled according to the posterior, meaning that, in the limit that the chain length tends to infinity, the relative frequency with which a given set of parameter appears in the chain is proportional to the desired posterior, $p(\vtheta|d,M)$.
+A Markov chain Monte Carlo produces a set of samples $\{ \vtheta^{(i)} \, | \, i = 1, \ldots \}$ from the model parameter space that are sampled according to the posterior, meaning that, in the limit that the chain length tends to infinity, the relative frequency with which a given set of parameter appears in the chain is proportional to the desired posterior, $p(\vtheta|d,M)$.
 Therefore, the output of an MCMC can be directly interpreted as the posterior PDF over the full parameter space, while PDFs for individual parameters can be obtained through trivially marginalizing over the uninteresting parameters, simply by considering the distribution of the parameter of interest in the MCMC samples.
@@
 \be
-p(\vtheta^(i+1))=\int_{V} d\vtheta^{i} p(\vtheta^{i} \to \vtheta^{i+1}) p(\vtheta^{i+1}),
+p(\vtheta^(i+1))=\int_{V} d\vtheta^{i} p(\vtheta^{(i)} \to \vtheta^{(i+1)}) p(\vtheta^{(i+1)}),
 \ee
-where $p(\vtheta^{i} \to \vtheta^{i+1})$ depends only on $\vtheta^{i}$ and $\vtheta^{i+1}$. 
+where $p(\vtheta^{(i)} \to \vtheta^{(i+1}))$ depends only on $\vtheta^{(i)}$ and $\vtheta^{(i+1)}$. 
 An additional requirement for an MCMC arises from the fact that the desired distribution is the equilibrium distribution.
-In other words, if we assume that state $(i)$ of the chain is sampled from the desired PDF, $p(\vtheta^{i})=p(\vtheta|d,M)$ , then the next state $(i+1)$ must be sampled from the PDF as well, so that $p(\vtheta^{i+1})=p(\vtheta|d,M)$.
+In other words, if we assume that state $(i)$ of the chain is sampled from the desired PDF, $p(\vtheta^{(i)})=p(\vtheta|d,M)$ , then the next state $(i+1)$ must be sampled from the PDF as well, so that $p(\vtheta^{(i+1)})=p(\vtheta|d,M)$.
 
@@
 \be
-p_{\textnormal{accept}}  \equiv \min\Bigl(1,  frac{p(\vtheta^p|d)}{p(\vtheta^(i)|d)} \frac{Q(\vtheta^p \to
-        \vtheta^(i))}{Q(\vtheta^(i) \to \vtheta^p)} \Bigr).
+p_{\textnormal{accept}}  \equiv \min\Bigl(1,  frac{p(\vtheta^p|d)}{p(\vtheta^{(i)}|d)} \frac{Q(\vtheta^p \to
+        \vtheta^{(i)})}{Q(\vtheta^{(i)} \to \vtheta^p)} \Bigr).
 \ee
 \item Pick a uniform random number $\alpha \in [0,1]$.
-If $\alpha<  p_{\textnormal{accept}}$, accept the proposed jump, setting $\vtheta^(i+1)=\vtheta^p$.
-Otherwise, reject the jump, and remain at the same location in parameter space for the next step, $\vtheta^(i+1)=\vtheta^(i)$.
- \end{itemize}
+If $\alpha<  p_{\textnormal{accept}}$, accept the proposed jump, setting $\vtheta^{(i+1)}=\vtheta^p$.
+Otherwise, reject the jump, and remain at the same location in parameter space for the next step, $\vtheta^{(i+1})=\vtheta^{(i)}$.
+ \end{enumerate}
 
-This jump proposal distribution $Q(\vtheta^(i) \to \vtheta^p)$ can depend on the parameters of the current state $\vtheta^(i)$, but not on the past history.
+This jump proposal distribution $Q(\vtheta^{(i)} \to \vtheta^p)$ can depend on the parameters of the current state $\vtheta^{(i)}$, but not on the past history.
 It must also allow any state within the allowed prior volume to be reachable by the MCMC.
@@
 
-The samples produced during a Markov Chain Monte Carlo can be used to directly perform a Monte Carlo evidence integral.
+The samples produced during by an MCMC algorithm can be used to directly perform a Monte Carlo evidence integral.
 This results in a harmonic mean estimator for the evidence, which suffers from a large variance and a possible bias.
-Additional techniques for the direct integration of evidence are described in
+Additional techniques for the direct integration of evidence are described in \cite{Weinberg}, but also suffer from a variety of shortcomings.
+\ilya{Will, please expand/modify this appropriately.}  An alternative approach to model selection among a set of models is based on performing an MCMC in a ``super-model'' that encompasses all of the models under consideration; this is known the the Reversible Jump Markov chain Monte Carlo (MCMC).
 
-An alternative approach to model selection  models
+The parameter space of the super-model in an RJMCMC consists of a discrete parameter that identifies the model, $M_i$, and a set of continuous parameters appropriate for that model, $\vtheta_i$.
+Thus, each sample consists of a model identifier and a location within the parameter space of that model, $\{M_i, \vtheta_i\}$.
+We perform the MCMC in the ``super-model'' parameter space just like a regular MCMC; in general, we can both propose jumps to different parameters within a model (intramodel jumps) and jumps to a different model with different parameters (intermodel jumps).
+The resulting chain samples from the posterior $p(M_i,\vtheta_i|d)$.
+As in a usual MCMC, the PDF on the model as a parameter, with other parameters ignored, is obtained by trivially marginalizing over the remaining parameters, i.e., the evidence for a given model is just
+\be
+    p(M_i|d) = \int d\vtheta_i p(M_i, \vtheta_i|d) \approx \frac{N_i}{N},
+ \ee
+ where $N_i$ is the number of RJMCMC samples listing the $i$'th model and $N$ is the total chain length.
+Thus, the probability of a particular model relative to other models under consideration is given by the fraction of RJMCMC samples lying in the parameter space of that model.
 
-Consider the problem of model selection among a set of models, and the
-``super-model'' that encompasses all the models under consideration.
-The parameter space of the super-model consists of a discrete
-parameter that identifies the choice of model, $M_i$, and the
-continuous parameters appropriate for this model, $\vtheta_i$.  We
-denote a point in the super-model parameter space by $\{M_i,
-\vtheta_i\}$; each such point is a statement that, e.g., ``the
-underlying mass distribution for black holes in the galaxy is a
-Gaussian, with parameters $\mu$ and $\sigma$,'' or ``the underlying
-mass distribution for black holes in the galaxy is a triple-bin
-histogram with parameters $w_1$, $w_2$, $w_3$, and $w_4$,'' or ....
-To compare models, we are interested in the quantity (see Equation
-\eqref{eq:model-posterior-def})
-\begin{equation}
-  p(M_i|d) = \int d\vtheta_i p(M_i, \vtheta_i|d).
-\end{equation}
-If we perform an MCMC in the super-model parameter space, then we
-obtain a chain of samples $\{M_i, \vtheta_i \, | \, i = 1, \ldots\}$
-distributed in parameter space with density $p(M_i,\vtheta_i|d)
-d\vtheta_i$ and we can estimate the integral as
-\begin{equation}
-    p(M_i|d) = \int d\vtheta_i p(M_i, \vtheta_i|d) \approx \frac{N_i}{N},
-\end{equation}
-where $N_i$ is the number of samples in the chain lying in the
-parameter space of model $M_i$ and $N$ is the total number of samples
-in the chain.  The fraction of samples lying in the parameter space of
-model $M_i$ gives the probability of that model relative to the other
-models under consideration.
+The main difficulty of achieving an efficient RJMCMC is finding a good jump proposal distribution for intermodel jumps.
+In order to have relatively high acceptance ratios for intermodel jumps, which is necessary for efficient mixing between models, jumps should be preferentially proposed into regions with a high posterior.
+However, because the algorithm is Markovian, it has no past memory, so a jump proposed into a model from outside it can not access information from earlier in the chain which may identify a posterior peak.
 
-To perform the MCMC in the super-model parameter space, we must
-propose jumps not only between points in a particular model's
-parameter space, but also between the parameter spaces of different
-models.  For this MCMC to be efficient, proposed jumps into a model
-from another should favor regions with large posterior; when the
-posterior is highly-peaked in a small region of parameter space,
-proposed jumps outside this region are unlikely to be accepted, and
-the reversible-jump MCMC samples will require a very long chain to
-properly sample the ``super-model'' posterior.
-
-We can exploit the information we have from single-model MCMCs to
-generate efficient jump proposal distributions for our reversible jump
-MCMC.  We would like to propose jumps that roughly follow the
-distribution of samples in the single-model MCMCs.
+The most promising solution to this problem is to identify a good jump proposal distribution in advance, by exploiting information from single-model MCMCs to generate efficient jump proposal distributions for our reversible jump MCMC.
+(Single-model MCMCs can take small local jumps within their model, meaning that they are much less likely than an RJMCMC to lose a high-posterior mode once it has been located.)  The ideal jump proposal distribution for the parameters within a model would consist of the posterior PDF for those parameters, $p(\vtheta_i|M_i,d)$, and single-model MCMCs already represent samples from these posterior PDFs.
+However, they are discrete, and so must be interpolated so that any location in the ``super-model'' parameter space can be reached.
+The novel strategy we propose for efficiently interpolating a discretely sampled PDF is described in the next section.
 
@@
 
-Mention example of bricks from \cite{Littenberg2009}.
+\ilya{Mention example of bricks from \cite{Littenberg2009}.}
 
@@
 
-\section{Conclusion/Discussion}
-
-...
-
-\subsection{Other Uses}
 The efficient interpolation of a PDF via a kD-tree described in this paper can be extremely useful in other contexts, beyond a reversible-jump MCMC.
@@
 
-...
+\ilya{Mention nested sampling?}
 
-Mention nested sampling?
+\section{Conclusion/Discussion}
 

### 2010-12-09 paper.tex
Added section 5.

@@
 
-\subsection{MCMC}
+\subsection{MCMC} \label{sec:mcmc}
 
@@
 
-BH mass paper [still not sure how to present this without repetition...  ]
+RJMCMC with efficient PDF interpolation via kD trees can be used in a large variety of problems in physics and astronomy, whenever Bayesian analysis is used to perform model selection.
+Below, we describe several scenarios in which we have successfully applied this technique.
+Moreover, PDF interpolation with a kD tree can be extremely useful in other contexts, beyond a reversible-jump MCMC.
+We suggest two examples below: generating efficient jump proposal distributions in a single-model MCMC and convergence tests.
 
-GW parameter estimation
+We have successfully employed the RJMCMC technique described above when evaluating several alternative models for the distribution of the masses of black holes in X-ray binaries \cite{Farr:2010}.
+We performed a Bayesian analysis of the mass distribution of stellar-mass black holes using the observed masses of 15 low-mass X-ray binary systems undergoing Roche lobe overflow and five high-mass, wind-fed X-ray binary systems.
+We considered ten different mass distribution models: Gaussian, double Gaussian, power law, exponential decay, log normal, and 1-, 2-, 3-, 4-, and 5-bin histograms.
+Each model was described by between two and six parameters.
+Model selection using RJMCMC with kD trees allowed us to determine that the mass distribution of the low-mass systems is  best fit by a power-law, while the distribution of the combined sample is best fit by the exponential model.
+We were thus able to conclude that the low-mass subsample is not consistent with being drawn from the distribution of the combined population \cite{Farr:2010}.
 
-The efficient interpolation of a PDF via a kD-tree described in this paper can be extremely useful in other contexts, beyond a reversible-jump MCMC.
-We suggest two examples below: generating efficient jump proposal distributions and convergence tests?
+Gravitational-wave astronomy provides the setting for another ongoing study.
+LIGO and Virgo gravitational-wave detectors have recently completed taking data, which is currently being analyzed.
+Interesting triggers from the detection pipeline that  searches for compact binary coalescences are followed up with Bayesian parameter estimation tools.
+In general, several possible models for the data are considered: it can contain pure noise, or noise superimposed on a gravitational-wave signal that may or may not include significant spins in the binary components.
+The models can include a large number of parameters (up to 15 for the model with two spinning components), and RJMCMC with interpolation via kD trees is again successful at computing the model evidences.
+\ilya{Add references throughout this paragraph.}
+
+As discussed in Section \ref{sec:mcmc}, finding an efficient jump proposal distribution can be a challenge even for a single-model MCMC.
+This is particularly difficult when the parameter space is multi-modal, in which case the Markov chain can get stuck on individual high-likelihood islands in the parameter space.
+Traversals of low-likelihood oceans by a series of short jumps are extremely unlikely, and random long jumps are almost always rejected, making sampling very inefficient.
+Ideally, an analytical understanding of the model that would give a glimpse of the structure of the archipelago of islands would allow for the creation of jump proposal distributions that combine short jumps designed to explore individual islands with directed long jumps that are likely to end up on neighboring islands.
+However, it is not always possible to gain such understanding short of performing the MCMC itself.
+One solution lies in running multiple chains that explore the parameter space, perhaps getting stuck on local maxima, and then combining the points and interpolating with a kD tree to create a jump proposal distribution.
+Although the chains in the first iteration will not have converged, and the relative sizes of the modes will not be accurately known, this will ensure that all of the modes (islands) are at least included in the jump proposal distribution for a second stage, greatly improving its efficiency.
+Other modifications to this technique, which we are currently investigating, include multiple iterative stages, with the interpolated samples from each previous stage used as a jump proposal distribution for the next stage, until the posterior PDF is read off from the final stage; and the use of higher temperatures in the early-stage chains to improve sampling.
+
+The possibility of MCMC chains becoming even temporarily stuck on local maxima also affects convergence.
+In the absence of a sufficient number of jump proposals between the local maxima, the MCMC may not accurately represent their relative sizes.
+For example, this can happen with antipodal sky location degeneracies for networks of three ground-based gravitational-wave interferometers or for low-frequency gravitational wave signals in the proposed space-borne instrument LISA \cite{LISA}.
+Although PDFs produced by MCMC chains will generally show both locations on the sky, the chain length may not be sufficiently long in practice to include enough jumps between the two high-likelihood antipodal locations.
+We propose that the convergence of such PDFs can be tested by interpolating them with kD trees as described above and using them as jump proposal distributions in a follow-up MCMC.
+This follow-up MCMC would make large global jump proposals rather than small local ones, meaning that it could rapidly accumulate sufficient points to determine whether the original run spent correct fractional amounts of time in the various modes.
 

### 2010-12-09 paper.tex
Fixes to 2.1

@@
 \end{equation}
-where $p(M_i, \vtheta_i |d)$ is the posterior distribution for the
+where $p(\vtheta_i |d, M_i)$ is the posterior distribution for the
 model parameters $\vtheta_i$ implied by the data in the context of
@@
 represents our beliefs before accumulating any of the data $d$, and
-$p(d)$ is an overall normalizing constant that ensures that $p(\vtheta_i|d,M_i)$ is properly normalized as a probability distribution on
+$p(d|M_i)$, called the evidence, is an overall normalizing constant that ensures that $p(\vtheta_i|d,M_i)$ is properly normalized as a probability distribution on
 the $\vtheta_i$.  This implies that the evidence is equal to
@@
 \be
-p(\vtheta^(i+1))=\int_{V} d\vtheta^{i} p(\vtheta^{(i)} \to \vtheta^{(i+1)}) p(\vtheta^{(i+1)}),
+p(\vtheta^{(i+1)})=\int_{V} d\vtheta^{i} p(\vtheta^{(i)} \to \vtheta^{(i+1)}) p(\vtheta^{(i+1)}),
 \ee
-where $p(\vtheta^{(i)} \to \vtheta^{(i+1}))$ depends only on $\vtheta^{(i)}$ and $\vtheta^{(i+1)}$. 
+where $p(\vtheta^{(i)} \to \vtheta^{(i+1)})$ depends only on $\vtheta^{(i)}$ and $\vtheta^{(i+1)}$. 
 An additional requirement for an MCMC arises from the fact that the desired distribution is the equilibrium distribution.
@@
 \begin{enumerate}
-  \item Given a current state $\vtheta^(i)$, propose the next state $\vtheta^p$ by drawing from a jump proposal distribution with probability $Q(\vtheta^(i) \to \vtheta^p)$.  
+  \item Given a current state $\vtheta^{(i)}$, propose the next state $\vtheta^p$ by drawing from a jump proposal distribution with probability $Q(\vtheta^{(i)} \to \vtheta^p)$.  
   \item Compute the probability of accepting the proposed jump as

### 2010-12-09 paper.tex
Fixed Eq. 2 reference.

@@
 When selecting among alternative models, we are interested in finding the model with the highest evidence $p(M_i|d)$.
-However, attempts to directly compute the evidence by performing the integration in Eq.~(\ref{bel}) are generally very difficult in a multi-dimensional, multi-modal parameter space when the likelihood has to be evaluated numerically.
+However, attempts to directly compute the evidence by performing the integration in Eq.~(\ref{evidence}) are generally very difficult in a multi-dimensional, multi-modal parameter space when the likelihood has to be evaluated numerically.
 In particular, a grid-based integral quickly becomes computationally unfeasible as the dimensionality of $\vtheta$ exceeds a few.

### 2010-12-09 paper.tex
Section 2 fixes.
Edits to Ilya's Section V, re-formatting.

@@
 
-We have successfully employed the RJMCMC technique described above when evaluating several alternative models for the distribution of the masses of black holes in X-ray binaries \cite{Farr:2010}.
-We performed a Bayesian analysis of the mass distribution of stellar-mass black holes using the observed masses of 15 low-mass X-ray binary systems undergoing Roche lobe overflow and five high-mass, wind-fed X-ray binary systems.
-We considered ten different mass distribution models: Gaussian, double Gaussian, power law, exponential decay, log normal, and 1-, 2-, 3-, 4-, and 5-bin histograms.
-Each model was described by between two and six parameters.
-Model selection using RJMCMC with kD trees allowed us to determine that the mass distribution of the low-mass systems is  best fit by a power-law, while the distribution of the combined sample is best fit by the exponential model.
-We were thus able to conclude that the low-mass subsample is not consistent with being drawn from the distribution of the combined population \cite{Farr:2010}.
+We have successfully employed the RJMCMC technique described above
+when evaluating several alternative models for the distribution of the
+masses of black holes in X-ray binaries \cite{Farr:2010}.  We
+performed a Bayesian analysis of the mass distribution of stellar-mass
+black holes using the observed masses of 15 low-mass X-ray binary
+systems undergoing Roche lobe overflow and five high-mass, wind-fed
+X-ray binary systems.  We considered ten different mass distribution
+models: Gaussian, double Gaussian, power law, exponential decay, log
+normal, and 1-, 2-, 3-, 4-, and 5-bin histograms.  Each model was
+described by between two and six parameters.  Model selection using
+RJMCMC with kD trees allowed us to determine that the mass
+distribution of the low-mass systems is best fit by a power-law, while
+the distribution of the combined sample is best fit by the exponential
+model.  Based on the model selection, we were able to determine which
+models would provide the best information about astrophysically
+relevant parameters, such as the minimum black hole mass.  We were
+also able to conclude that the low-mass subsample is not consistent
+with being drawn from the distribution of the combined population
+\cite{Farr:2010}.
 
-Gravitational-wave astronomy provides the setting for another ongoing study.
-LIGO and Virgo gravitational-wave detectors have recently completed taking data, which is currently being analyzed.
-Interesting triggers from the detection pipeline that  searches for compact binary coalescences are followed up with Bayesian parameter estimation tools.
-In general, several possible models for the data are considered: it can contain pure noise, or noise superimposed on a gravitational-wave signal that may or may not include significant spins in the binary components.
-The models can include a large number of parameters (up to 15 for the model with two spinning components), and RJMCMC with interpolation via kD trees is again successful at computing the model evidences.
-\ilya{Add references throughout this paragraph.}
+Gravitational-wave astronomy provides the setting for another ongoing
+study.  Interesting triggers from the LIGO and Virgo
+gravitational-wave detector pipeline that searches for compact binary
+coalescences are followed up with Bayesian parameter estimation tools.
+In general, several possible models for the data are considered,
+including a pure noise model, noise superimposed on a non-spinning
+gravitational-wave signal, or noise superimposed on a gravitational
+wave signal that includes significant spin or spins in the binary
+components.  The models can include a large number of parameters (up
+to 15 for the model with two spinning components), and RJMCMC with
+interpolation via kD trees is again successful at computing the model
+evidences. \ilya{Add references throughout this paragraph.}
 
@@
 
-The possibility of MCMC chains becoming even temporarily stuck on local maxima also affects convergence.
-In the absence of a sufficient number of jump proposals between the local maxima, the MCMC may not accurately represent their relative sizes.
-For example, this can happen with antipodal sky location degeneracies for networks of three ground-based gravitational-wave interferometers or for low-frequency gravitational wave signals in the proposed space-borne instrument LISA \cite{LISA}.
-Although PDFs produced by MCMC chains will generally show both locations on the sky, the chain length may not be sufficiently long in practice to include enough jumps between the two high-likelihood antipodal locations.
-We propose that the convergence of such PDFs can be tested by interpolating them with kD trees as described above and using them as jump proposal distributions in a follow-up MCMC.
-This follow-up MCMC would make large global jump proposals rather than small local ones, meaning that it could rapidly accumulate sufficient points to determine whether the original run spent correct fractional amounts of time in the various modes.
-
-\ilya{Mention nested sampling?}
+The possibility of MCMC chains becoming even temporarily stuck on
+local maxima also affects convergence.  In the absence of a sufficient
+number of jump proposals between the local maxima, the MCMC may not
+accurately represent their relative sizes.  For example, this can
+happen with antipodal sky location degeneracies for networks of three
+ground-based gravitational-wave interferometers or for low-frequency
+gravitational wave signals in the proposed space-borne instrument LISA
+\cite{LISA}.  Although PDFs produced by MCMC chains will generally
+show both locations on the sky, the chain length may not be
+sufficiently long in practice to include enough jumps between the two
+high-likelihood antipodal locations.  In general, the relative weight
+of two separated peaks is determined with a fractional error that
+scales as $1/\sqrt{N_{\textnormal{transitions}}}$, where
+$N_{\textnormal{transitions}}$ is the number of transitions between
+the peaks.  By increasing the number of proposed transitions, the
+convergence of such PDFs can be enhanced using the kD tree
+interpolated proposal.  We can test the convergence of an MCMC with
+local jumps by running a followup MCMC with a jump proposal
+distribution from an interpolation of the first MCMC's PDF.  This
+follow-up MCMC would make large global jump proposals rather than
+small local ones, meaning that it could rapidly accumulate sufficient
+points to determine whether the original run spent correct fractional
+amounts of time in the various modes.
 

### 2010-12-14 paper.tex
Edits to Bayesian Analysis section; formatting.
Changed \begin{equation} to \be, etc.  Fixed some language (model
evidence --> model posterior probability, etc).  Added Weinberg to
references.

@@
 \begin{abstract}
-\ilya{Add physics/astro motivation}
-  Reversible-jump Markov chain monte carlo (RJMCMC) is an extremely
-  powerful technique for performing Bayesian model selection, but it
-  suffers from a fundamental difficulty: it requires jumps between
-  model parameter spaces, but cannot retain a memory of the favored
-  locations in more than one parameter space at a time.  Thus, a naive
-  jump between parameter spaces is unlikely to be accepted in the MCMC
-  algorithm and convergence is correspondingly slow.  Here we
-  demonstrate an interpolation technique that uses samples from
-  single-model MCMCs to propose inter-model jumps from an
-  approximation to the single-model posterior of the target parameter
-  space.  The interpolation technique, based on a kD-tree
-  data structure, is adaptive and efficient in arbitrary dimensions.
-  We show that our technique leads to dramatically improved convergence
-  over naive jumps in an RJMCMC, and compare it to other proposals in
-  the literature to improve the convergence of RJMCMCs.  We also
-  discuss the use of the same interpolation technique in two other
-  contexts: as a convergence test for a single-model MCMC and as a way
-  to construct efficient ``global'' proposal distributions for
-  single-model MCMCs without prior knowledge of the structure of the
-  posterior distribution.
+  \ilya{Add physics/astro motivation} Reversible-jump Markov chain
+  monte carlo (RJMCMC) is an extremely powerful technique for
+  performing Bayesian model selection, but it suffers from a
+  fundamental difficulty: it requires jumps between model parameter
+  spaces, but cannot retain a memory of the favored locations in more
+  than one parameter space at a time.  Thus, a naive jump between
+  parameter spaces is unlikely to be accepted in the MCMC algorithm
+  and convergence is correspondingly slow.  Here we demonstrate an
+  interpolation technique that uses samples from single-model MCMCs to
+  propose inter-model jumps from an approximation to the single-model
+  posterior of the target parameter space.  The interpolation
+  technique, based on a kD-tree data structure, is adaptive and
+  efficient in arbitrary dimensions.  We show that our technique leads
+  to dramatically improved convergence over naive jumps in an RJMCMC,
+  and compare it to other proposals in the literature to improve the
+  convergence of RJMCMCs.  We also discuss the use of the same
+  interpolation technique in two other contexts: as a convergence test
+  for a single-model MCMC and as a way to construct efficient
+  ``global'' proposal distributions for single-model MCMCs without
+  prior knowledge of the structure of the posterior distribution.
 \end{abstract}
@@
 \section{Introduction}
+
+\will{Mention Weinberg direct integration (Ref.~\cite{Weinberg2009}).}
 
@@
 
-Consider an observed data set $d$ and a set of competing models for the data, indexed by an integer $i$: $\{M_i | i = 1, 2, \ldots \}$.
-Each model has some continuous parameters, $\vtheta_i$; given the model and its parameters, we can make a prediction about the likelihood of observing
-the experimental data: $L(d|\vtheta_i, M_i)$.  Within the framework of each model, Bayes' rule gives us a way to compute the posterior probability distribution function (PDF) for the model
-parameters implied by the data:
-\begin{equation}
+Consider an observed data set $d$ and a set of competing models for
+the data, indexed by an integer $i$: $\{M_i | i = 1, 2, \ldots \}$.
+Each model has some continuous parameters, $\vtheta_i$; given the
+model and its parameters, we can make a prediction about the
+likelihood of observing the experimental data: $L(d|\vtheta_i, M_i)$.
+Within the framework of each model, Bayes' rule gives us a way to
+compute the posterior probability distribution function (PDF) for the
+model parameters implied by the data:
+\be
   p(\vtheta_i | d, M_i) = \frac{L(d|\vtheta_i, M_i) p(\vtheta_i|M_i)}{p(d|M_i)},
-\end{equation}
+\ee
 where $p(\vtheta_i |d, M_i)$ is the posterior distribution for the
 model parameters $\vtheta_i$ implied by the data in the context of
-model $M_i$, $p(\vtheta_i|M_i)$ is the prior
-probability of the model parameters that 
-represents our beliefs before accumulating any of the data $d$, and
-$p(d|M_i)$, called the evidence, is an overall normalizing constant that ensures that $p(\vtheta_i|d,M_i)$ is properly normalized as a probability distribution on
-the $\vtheta_i$.  This implies that the evidence is equal to
+model $M_i$, $p(\vtheta_i|M_i)$ is the prior probability of the model
+parameters that represents our beliefs before accumulating any of the
+data $d$, and $p(d|M_i)$, called the evidence, is an overall
+normalizing constant that ensures that $p(\vtheta_i|d,M_i)$ is
+properly normalized as a probability distribution on the $\vtheta_i$.
+This implies that the evidence is equal to
 \bel{evidence}
   p(d|M_i) = \int_{V_i} d\vtheta_i L(d|\vtheta_i, M_i) p(\vtheta_i|M_i),
-\end{equation}
-where $V_i$ is the parameter space volume in model $M_i$.  We are interested in the evidence for model $M_i$ given the data, $p(M_i|d)$, which is again given by Bayes' rule as
-\begin{equation}
+\ee
+where $V_i$ is the parameter space volume in model $M_i$.  For model
+comparison, we are interested in the posterior probability of a
+particular model, $M_i$, given the data, $p(M_i|d)$.  Using Bayes'
+rule, we see that this involves the evidence (Eq.~\eqref{evidence}):
+\be
 p(M_i|d) = \frac{p(d|M_i) p(M_i)}{p(d)},
-\end{equation}
-where $p(M_i)$ is our a priori belief in model $M_i$ and $p(d)$ is a normalizing constant, 
+\ee
+where $p(M_i)$ is our a priori belief in model $M_i$ and $p(d)$ is a
+normalizing constant,
 \be
@@
 
-When selecting among alternative models, we are interested in finding the model with the highest evidence $p(M_i|d)$.
-However, attempts to directly compute the evidence by performing the integration in Eq.~(\ref{evidence}) are generally very difficult in a multi-dimensional, multi-modal parameter space when the likelihood has to be evaluated numerically.
-In particular, a grid-based integral quickly becomes computationally unfeasible as the dimensionality of $\vtheta$ exceeds a few.
-The parameter space must typically be explored in a stochastic manner before the evidence integral can be computed.
-Although several stochastic parameter-exploration techniques are focused directly on evidence computation (e.g., nested sampling and MultiNest \ilya {Add references}), we frequently want to compute the posterior PDFs within each model along with the evidences for the various models, and the Markov chain Monte Carlo is one of the most widely used tools for this purpose.
+When selecting among alternative models, we are interested in finding
+the model with the highest posterior probability $p(M_i|d)$.  However,
+attempts to directly compute the evidence by performing the
+integration in Eq.~\eqref{evidence} are generally very difficult in a
+multi-dimensional, multi-modal parameter space when the likelihood has
+to be evaluated numerically.  In particular, a grid-based integral
+quickly becomes computationally unfeasible as the dimensionality of
+$\vtheta$ exceeds a few.  The parameter space must typically be
+explored in a stochastic manner before the evidence integral can be
+computed.  There are several stochastic parameter-exploration
+techniques focused directly on evidence computation (e.g., nested
+sampling and MultiNest \ilya {Add references}).  However, we
+frequently want to compute the posterior PDFs within each model along
+with the evidences for the various models.  One of the most common
+techniques for computing posterior PDFs in the context of a model is
+the Markov Chain Monte Carlo, which we now describe.
 
@@
 \be
-p_{\textnormal{accept}}  \equiv \min\Bigl(1,  frac{p(\vtheta^p|d)}{p(\vtheta^{(i)}|d)} \frac{Q(\vtheta^p \to
+p_{\textnormal{accept}}  \equiv \min\Bigl(1,  \frac{p(\vtheta^p|d)}{p(\vtheta^{(i)}|d)} \frac{Q(\vtheta^p \to
         \vtheta^{(i)})}{Q(\vtheta^{(i)} \to \vtheta^p)} \Bigr).
@@
 If $\alpha<  p_{\textnormal{accept}}$, accept the proposed jump, setting $\vtheta^{(i+1)}=\vtheta^p$.
-Otherwise, reject the jump, and remain at the same location in parameter space for the next step, $\vtheta^{(i+1})=\vtheta^{(i)}$.
+Otherwise, reject the jump, and remain at the same location in parameter space for the next step, $\vtheta^{(i+1)}=\vtheta^{(i)}$.
  \end{enumerate}

### 2010-12-14 paper.tex
Edits to RJMCMC section, more references.

@@
 
+\will{Here $i$ should become $j$ (or some other index), because we
+  have used $i$ already to denote the model.}
+
 A Markov chain Monte Carlo produces a set of samples $\{ \vtheta^{(i)} \, | \, i = 1, \ldots \}$ from the model parameter space that are sampled according to the posterior, meaning that, in the limit that the chain length tends to infinity, the relative frequency with which a given set of parameter appears in the chain is proportional to the desired posterior, $p(\vtheta|d,M)$.
@@
 
-The samples produced during by an MCMC algorithm can be used to directly perform a Monte Carlo evidence integral.
-This results in a harmonic mean estimator for the evidence, which suffers from a large variance and a possible bias.
-Additional techniques for the direct integration of evidence are described in \cite{Weinberg}, but also suffer from a variety of shortcomings.
-\ilya{Will, please expand/modify this appropriately.}  An alternative approach to model selection among a set of models is based on performing an MCMC in a ``super-model'' that encompasses all of the models under consideration; this is known the the Reversible Jump Markov chain Monte Carlo (MCMC).
+The samples produced during by an MCMC algorithm can be used to
+directly perform a Monte Carlo evidence integral.  This results in a
+harmonic mean estimator for the evidence, which suffers from a large
+variance and a possible bias\will{Ilya: Can you add citations here for
+  the suckiness of HM?}.  Additional techniques for the direct
+integration of evidence, also based on a kD tree decomposition of the
+parameter space (see Sec.~\ref{sec:kDTree}), are described in
+\cite{Weinberg2009}.  These techniques are promising, but in some
+cases suffer \cite{Farr2010} from large variance and bias as well.  An
+alternative approach to model selection among a set of models is based
+on performing an MCMC in a ``super-model'' that encompasses all of the
+models under consideration; this is known the the Reversible Jump
+Markov chain Monte Carlo (RJMCMC).
 
-The parameter space of the super-model in an RJMCMC consists of a discrete parameter that identifies the model, $M_i$, and a set of continuous parameters appropriate for that model, $\vtheta_i$.
-Thus, each sample consists of a model identifier and a location within the parameter space of that model, $\{M_i, \vtheta_i\}$.
-We perform the MCMC in the ``super-model'' parameter space just like a regular MCMC; in general, we can both propose jumps to different parameters within a model (intramodel jumps) and jumps to a different model with different parameters (intermodel jumps).
-The resulting chain samples from the posterior $p(M_i,\vtheta_i|d)$.
-As in a usual MCMC, the PDF on the model as a parameter, with other parameters ignored, is obtained by trivially marginalizing over the remaining parameters, i.e., the evidence for a given model is just
+The parameter space of the super-model in an RJMCMC consists of a
+discrete parameter that identifies the model, $M_i$, and a set of
+continuous parameters appropriate for that model, $\vtheta_i$.  Thus,
+each sample consists of a model identifier and a location within the
+parameter space of that model, $\{M_i, \vtheta_i\}$.  We perform the
+MCMC in the ``super-model'' parameter space just like a regular MCMC;
+in general, we can both propose jumps to different parameters within a
+model (intramodel jumps) and jumps to a different model with different
+parameters (intermodel jumps).  The resulting chain samples from the
+posterior $p(M_i,\vtheta_i|d)$.  As in a usual MCMC, the PDF on the
+model as a parameter, with other parameters ignored, is obtained by
+marginalizing over the remaining parameters, i.e., the evidence for a
+given model is just
+%
 \be
@@
 \ee
- where $N_i$ is the number of RJMCMC samples listing the $i$'th model and $N$ is the total chain length.
-Thus, the probability of a particular model relative to other models under consideration is given by the fraction of RJMCMC samples lying in the parameter space of that model.
+%
+where $N_i$ is the number of RJMCMC samples listing the $i$'th model
+and $N$ is the total chain length.  Thus, the probability of a
+particular model relative to other models under consideration is given
+by the fraction of RJMCMC samples lying in the parameter space of that
+model.
 
-The main difficulty of achieving an efficient RJMCMC is finding a good jump proposal distribution for intermodel jumps.
-In order to have relatively high acceptance ratios for intermodel jumps, which is necessary for efficient mixing between models, jumps should be preferentially proposed into regions with a high posterior.
-However, because the algorithm is Markovian, it has no past memory, so a jump proposed into a model from outside it can not access information from earlier in the chain which may identify a posterior peak.
+The main difficulty of achieving an efficient RJMCMC is finding a good
+jump proposal distribution for intermodel jumps.  In order to have
+relatively high acceptance ratios for intermodel jumps, which is
+necessary for efficient mixing between models, jumps should be
+preferentially proposed into regions with a high posterior.  However,
+because the algorithm is Markovian, it has no past memory, so a jump
+proposed into a model from outside can not access information from
+earlier in the chain which may identify a posterior peak.
 
-The most promising solution to this problem is to identify a good jump proposal distribution in advance, by exploiting information from single-model MCMCs to generate efficient jump proposal distributions for our reversible jump MCMC.
-(Single-model MCMCs can take small local jumps within their model, meaning that they are much less likely than an RJMCMC to lose a high-posterior mode once it has been located.)  The ideal jump proposal distribution for the parameters within a model would consist of the posterior PDF for those parameters, $p(\vtheta_i|M_i,d)$, and single-model MCMCs already represent samples from these posterior PDFs.
-However, they are discrete, and so must be interpolated so that any location in the ``super-model'' parameter space can be reached.
-The novel strategy we propose for efficiently interpolating a discretely sampled PDF is described in the next section.
+The way to solve this problem is to identify a good jump proposal
+distribution in advance, by exploiting information from single-model
+MCMCs to generate efficient jump proposal distributions for our
+reversible jump MCMC.  (Single-model MCMCs can take small local jumps
+within their model, meaning that they are much less likely than an
+RJMCMC to lose a high-posterior mode once it has been located.)  The
+ideal jump proposal distribution for the parameters within a model
+would consist of the posterior PDF for those parameters,
+$p(\vtheta_i|M_i,d)$, and single-model MCMCs already represent samples
+from these posterior PDFs.  However, the samples are discrete, and a
+jump proposal must be continuous.  Therefore, the output of each
+single-model MCMC must be interpolated to construct the desired jump
+proposal.  The novel strategy we propose for efficiently interpolating
+a discretely sampled PDF is described in the next section.
 
 \section{kD Trees and Interpolation}
+\label{sec:kDTree}
+
+From the set of points output by a single-model MCMC,
+$\{\vtheta_i^{(j)} | j = 1, 2, \ldots \}$, all lying in the $M_i$
+parameter space, we would like to construct a set of ``neighborhoods''
+associated with each of the sample points.  The neighborhoods should
+be non-overlapping and fill parameter space.
 

### 2010-12-14 paper.tex
Added acceptRate plots, Makefile, and Matlab script for plot.

@@
 \usepackage{color}
+\usepackage{graphicx}
 
@@
 associated with each of the sample points.  The neighborhoods should
-be non-overlapping and fill parameter space.
+be non-overlapping and fill parameter space.  To draw a proposed jump
+from an approximation to the model PDF, we select a point uniformly
+from the $\vtheta^{(j)}_i$, find its associated neighborhood, and then
+draw the proposed jump uniformly from the neighborhood.  Since the
+MCMC points are distributed according to the posterior PDF for $M_i$,
+this procedure produces proposed jumps that are approximately
+distributed according to the posterior PDF.  In fact, the proposed
+jumps are drawn from a piecewise-constant (i.e.\ constant on each
+neighborhood) interpolation of the PDF.
 
-\ilya{Mention example of bricks from \cite{Littenberg2009}.}
+There are various techniques that could be used to construct the set
+of neighborhoods associated with each point.
+Ref.~\cite{Littenberg2009} decomposed parameter space into
+constant-volume ``bricks'' whose size was set by the typical size of
+the peaks of the PDF.  Each point was associated with the brick that
+contained it.  Unfortunately, this process does not guarantee that
+every brick contains at least one point, which is required for the
+jump proposal to cover the entire model parameter space.  The solution
+to this problem was to add a uniform distribution to the proposal PDF,
+in the form of one ``fictitious'' point per brick.
 
 \section{RJMCMC Efficiency}
+
+\begin{figure}
+  \begin{center}
+    \includegraphics[width=0.8\textwidth]{acceptRate}
+  \end{center}
+  \caption{\label{fig:acceptRate} The inter-model acceptance rate as a
+    function of boxing number for the RJMCMC described in the text.}
+\end{figure}
 

### 2010-12-14 makePlots.m

@@ -1,0 +1,8 @@
+%%%%%%%%%%%
+% Acceptance Rate Plot
+data=importdata('acceptRate.dat');
+plot(data(:,1), data(:,3)/1000000.0, 'k');
+xlabel('N_{boxing}');
+ylabel('f_{accept}');
+axis([0 10000 0 inf]);
+print -deps 'acceptRate.eps'

### 2010-12-14 paper.tex
Minor edits, Makefile update.
Added 1,2,3 description of kD-tree algorithm.

@@
 
-From the set of points output by a single-model MCMC,
-$\{\vtheta_i^{(j)} | j = 1, 2, \ldots \}$, all lying in the $M_i$
-parameter space, we would like to construct a set of ``neighborhoods''
-associated with each of the sample points.  The neighborhoods should
-be non-overlapping and fill parameter space.  To draw a proposed jump
-from an approximation to the model PDF, we select a point uniformly
-from the $\vtheta^{(j)}_i$, find its associated neighborhood, and then
-draw the proposed jump uniformly from the neighborhood.  Since the
-MCMC points are distributed according to the posterior PDF for $M_i$,
+The problem of drawing a proposed jump from an interpolation of
+single-model MCMC points can be thought of as the problem of assigning
+a local ``neighborhood'' to each point in the chain of MCMC samples.
+These neighborhoods should be non-overlapping and fill parameter
+space.  To draw a proposed jump, we select a point uniformly from the
+MCMC samples, find its associated neighborhood, and then draw the
+proposed jump uniformly from the neighborhood.  Since the MCMC points
+are distributed according to the posterior PDF for the single model,
 this procedure produces proposed jumps that are approximately
@@
 jumps are drawn from a piecewise-constant (i.e.\ constant on each
-neighborhood) interpolation of the PDF.
+neighborhood) interpolation of the PDF.  There are various techniques
+that could be used to construct the set of neighborhoods associated
+with each point.
 
-There are various techniques that could be used to construct the set
-of neighborhoods associated with each point.
 Ref.~\cite{Littenberg2009} decomposed parameter space into
@@
 to this problem was to add a uniform distribution to the proposal PDF,
-in the form of one ``fictitious'' point per brick.
+in the form of one ``fictitious'' point per brick.  The extra uniform
+distribution from the fictituous points represents a significant cost,
+since inter-model jumps proposed from the uniform distribution are
+unlikely to be accepted.  Additionally, the bricks in this algorithm
+do not adapt to the local structure of the PDF.  One must either use
+small bricks to capture the local structure of the PDF, placing many
+bricks in regions without MCMC samples, or use large bricks, missing
+the local strucutre of the PDF in exchange for fewer empty bricks.
+
+An alternate technique for producing adaptive neighborhoods would be
+to use the Voronoi regions associated with each MCMC sample.  The
+Voronoi region associated with a sample contains all the parameter
+space points that are closer to that sample than any other sample.
+The Voronoi region decomposition into neighborhoods is, in a sense,
+maximally adaptive, in contrast to the approach of
+Ref.~\cite{Littenberg2009}, which is minimally adaptive.
+Unfortunately, defining the Voronoi regions requires a metric on
+parameter space, which may be difficult or impossible to define.  A
+larger difficulty with the Voronoi approach is that the computational
+cost for computing the Voronoi regions for $N$ points increases
+rapidly with dimension, $d$, as $N^X$ \will{Find X in the literature;
+  add references.  I think it's $X = d/2$}.
+
+We propose to use a decomposition of parameter space into
+neighborhoods based on a datastructure called a kD-tree
+\will{References.}.  The decomposition is more adaptive than the cells
+of Ref.~\cite{Littenberg2009}, and more efficient in high dimension
+than the Voronoi decomposition.
+
+A kD-tree is a binary, space-partitioning tree.  To partition a set of
+points into a kD-tree, begin by placing them in a rectangular box that
+contains all of parameter space.  Then proceed recursively:
+\begin{enumerate}
+\item If the given box contains exactly one point, stop; this is a
+  leaf of the tree.  Otherwise:
+\item Choose a dimension along which to divide the points.  Divide the
+  points in half along this dimension (or nearly in half, if the
+  number of points is odd), forming two sub-boxes.  The ``left''
+  sub-box contains the half (or nearly half) of the points that have
+  small coordinates along the chosen dimension; the ``right'' sub-box
+  contains the half (or nearly half) of the points that have large
+  coordinates along the chosen dimension.
+\item Return to step 1 with each of the sub-boxes.
+\end{enumerate}
 

### 2010-12-14 paper.tex
More text on kD-trees and finding the neighborhood associated with a point.

@@
 \newcommand{\will}[1]{{\color{blue} \bf Will: #1}}
+
+\newcommand{\order}[1]{\mathcal{O}\left( #1 \right)}
 
@@
 
-\will{Mention Weinberg direct integration (Ref.~\cite{Weinberg2009}).}
+\will{Mention Weinberg direct integration (Ref.~\cite{Weinberg2009}).
+  Also \cite{Ascasibar2005} in the cosmological context (not really
+  kD-tree, but close).}
 
@@
   \item Compute the probability of accepting the proposed jump as
-\be
+\bel{eq:p-accept}
 p_{\textnormal{accept}}  \equiv \min\Bigl(1,  \frac{p(\vtheta^p|d)}{p(\vtheta^{(i)}|d)} \frac{Q(\vtheta^p \to
@@
   coordinates along the chosen dimension.
-\item Return to step 1 with each of the sub-boxes.
+\item Return to Step 1 with each of the sub-boxes, storing the
+  resulting trees as sub-trees of the current box.
+\end{enumerate}
+The key algorithmic step in the production of a kD-tree is finding the
+median point along a given dimension in order to divide the points in
+half in Step 2.  For $n$ points, this can be accomplished in
+$\order{n}$ time (see, e.g., Ref.~\cite{Press2007}).  If there are $N$
+points in total, there are $\order{\log N}$ levels in the tree; at
+each level, $\order{N}$ points must be processed once in the
+median-finding algorithm.  Tree construction thus costs $\order{N \log
+  N}$ in time, and the tree consumes $\order{N \log N}$ space.  Box
+boundaries for a kD-tree constructed around a point set that is
+normally distributed around the origin in two dimensions are shown in
+Figure \ref{fig:kD-tree}.
+
+\begin{figure}
+  \begin{center}
+    \includegraphics[width=0.8\textwidth]{kdtree.png}
+  \end{center}
+  \caption{\label{fig:kD-tree} The neighborhoods from a kD-tree
+    constructed around a set of points that are normally distributed
+    about the origin in two dimensions.  As the points become denser
+    around the origin, the typical neighborhood gets smaller.  The
+    interpolated PDF within a box of volume $V_i$ is $1/(N V_i)$,
+    where $N$ is the total number of points (and boxes).}
+\end{figure}
+
+In order to use the kD-tree interpolation in an MCMC, we must be able
+to quickly find the neighborhood associated with a given point.  This
+can be accomplished in $\order{\log N}$ time and constant space with
+the following algorithm.  Given the point, $\vtheta_i$---which could
+either be a point in the chain of MCMC samples if we are trying to
+propose a jump, or a previously proposed jump if we are trying to
+compute the jump probability distribution (see
+Eq.~\ref{eq:p-accept})---and the tree, $T$:
+\begin{enumerate}
+\item If $T$ contains exactly one sample point, then its box is the
+  associated neighborhood.  Otherwise:
+\item The tree $T$ has two sub-trees.  If the point $\vtheta_i$ is
+  contained in the ``left'' sub-tree, then return to Step 1,
+  considering this sub-tree; otherwise return to Step 1, considering
+  the ``right'' sub-tree.
 \end{enumerate}
@@
 when evaluating several alternative models for the distribution of the
-masses of black holes in X-ray binaries \cite{Farr:2010}.  We
+masses of black holes in X-ray binaries \cite{Farr2010}.  We
 performed a Bayesian analysis of the mass distribution of stellar-mass
@@
 with being drawn from the distribution of the combined population
-\cite{Farr:2010}.
+\cite{Farr2010}.
 

### 2010-12-19 paper.tex
Added references in section 2, first sentence in abstract, responded to Will's section 2 comments.

@@
 
+\def\aj{Astronomical Journal}                 % Astronomical Journal
+\def\apj{Astrophysical Journal}                % Astrophysical Journal
+\def\apjl{Astrophysical Journal}             % Astrophysical Journal, Letters
+\def\pasj{PASJ}
+\def\apjs{ApJS}              % Astrophysical Journal, Supplement
+\def\mnras{MNRAS}            % Monthly Notices of the RAS
+\def\prd{Phys.~Rev.~D}       % Physical Review D
+\def\prl{Phys.~Rev.~Lett}    % Physical Review Letters
+\def\cqg{Class.~Quant.~Grav.~}%Classical and Quantum Gravity
+\def\araa{ARA\&A}             % Annual Review of Astron and Astrophys
+\def\nat{Nature}              % Nature
+\def\aap{A\&A}                % Astronomy and Astrophysics
+
 \bibliographystyle{h-physrev}
@@
 \begin{abstract}
-  \ilya{Add physics/astro motivation} Reversible-jump Markov chain
+  Selection among alternative theoretical models given an observed data set is an important challenge in many areas of physics and astronomy, ranging from gravitational-wave data analysis to measurements of black-hole masses in X-ray binaries.
+Reversible-jump Markov chain
   monte carlo (RJMCMC) is an extremely powerful technique for
@@
 techniques focused directly on evidence computation (e.g., nested
-sampling and MultiNest \ilya {Add references}).  However, we
+sampling \cite{Skilling:2006} and MultiNest \cite{Feroz:2009}).  However, we
 frequently want to compute the posterior PDFs within each model along
@@
 
-\will{Here $i$ should become $j$ (or some other index), because we
-  have used $i$ already to denote the model.}
-
-A Markov chain Monte Carlo produces a set of samples $\{ \vtheta^{(i)} \, | \, i = 1, \ldots \}$ from the model parameter space that are sampled according to the posterior, meaning that, in the limit that the chain length tends to infinity, the relative frequency with which a given set of parameter appears in the chain is proportional to the desired posterior, $p(\vtheta|d,M)$.
+A Markov chain Monte Carlo \cite{Gilks:1996} produces a set of samples $\{ \vtheta^{(j)} \, | \, i = 1, \ldots \}$ from the model parameter space that are sampled according to the posterior, meaning that, in the limit that the chain length tends to infinity, the relative frequency with which a given set of parameter appears in the chain is proportional to the desired posterior, $p(\vtheta|d,M)$.
 Therefore, the output of an MCMC can be directly interpreted as the posterior PDF over the full parameter space, while PDFs for individual parameters can be obtained through trivially marginalizing over the uninteresting parameters, simply by considering the distribution of the parameter of interest in the MCMC samples.
@@
 \be
-p(\vtheta^{(i+1)})=\int_{V} d\vtheta^{i} p(\vtheta^{(i)} \to \vtheta^{(i+1)}) p(\vtheta^{(i+1)}),
+p(\vtheta^{(j+1)})=\int_{V} d\vtheta^{(j)} p(\vtheta^{(j)} \to \vtheta^{(j+1)}) p(\vtheta^{(j)}),
 \ee
-where $p(\vtheta^{(i)} \to \vtheta^{(i+1)})$ depends only on $\vtheta^{(i)}$ and $\vtheta^{(i+1)}$. 
+where the jump probability $p(\vtheta^{(j)} \to \vtheta^{(j+1)})$ depends only on $\vtheta^{(j)}$ and $\vtheta^{(j+1)}$. 
 An additional requirement for an MCMC arises from the fact that the desired distribution is the equilibrium distribution.
-In other words, if we assume that state $(i)$ of the chain is sampled from the desired PDF, $p(\vtheta^{(i)})=p(\vtheta|d,M)$ , then the next state $(i+1)$ must be sampled from the PDF as well, so that $p(\vtheta^{(i+1)})=p(\vtheta|d,M)$.
+In other words, if we assume that state $(j)$ of the chain is sampled from the desired PDF, $p(\vtheta^{(j)})=p(\vtheta|d,M)$ , then the next state $(j+1)$ must be sampled from the PDF as well, so that $p(\vtheta^{(j+1)})=p(\vtheta|d,M)$.
 
@@
 \begin{enumerate}
-  \item Given a current state $\vtheta^{(i)}$, propose the next state $\vtheta^p$ by drawing from a jump proposal distribution with probability $Q(\vtheta^{(i)} \to \vtheta^p)$.  
+  \item Given a current state $\vtheta^{(j)}$, propose the next state $\vtheta^p$ by drawing from a jump proposal distribution with probability $Q(\vtheta^{(j)} \to \vtheta^p)$.  
   \item Compute the probability of accepting the proposed jump as
 \bel{eq:p-accept}
-p_{\textnormal{accept}}  \equiv \min\Bigl(1,  \frac{p(\vtheta^p|d)}{p(\vtheta^{(i)}|d)} \frac{Q(\vtheta^p \to
-        \vtheta^{(i)})}{Q(\vtheta^{(i)} \to \vtheta^p)} \Bigr).
+p_{\textnormal{accept}}  \equiv \min\Bigl(1,  \frac{p(\vtheta^p|d)}{p(\vtheta^{(j)}|d)} \frac{Q(\vtheta^p \to
+        \vtheta^{(j)})}{Q(\vtheta^{(j)} \to \vtheta^p)} \Bigr).
 \ee
 \item Pick a uniform random number $\alpha \in [0,1]$.
-If $\alpha<  p_{\textnormal{accept}}$, accept the proposed jump, setting $\vtheta^{(i+1)}=\vtheta^p$.
-Otherwise, reject the jump, and remain at the same location in parameter space for the next step, $\vtheta^{(i+1)}=\vtheta^{(i)}$.
+If $\alpha<  p_{\textnormal{accept}}$, accept the proposed jump, setting $\vtheta^{(j+1)}=\vtheta^p$.
+Otherwise, reject the jump, and remain at the same location in parameter space for the next step, $\vtheta^{(j+1)}=\vtheta^{(j)}$.
  \end{enumerate}
 
-This jump proposal distribution $Q(\vtheta^{(i)} \to \vtheta^p)$ can depend on the parameters of the current state $\vtheta^{(i)}$, but not on the past history.
+This jump proposal distribution $Q(\vtheta^{(j)} \to \vtheta^p)$ can depend on the parameters of the current state $\vtheta^{(j)}$, but not on the past history.
 It must also allow any state within the allowed prior volume to be reachable by the MCMC.
@@
 harmonic mean estimator for the evidence, which suffers from a large
-variance and a possible bias\will{Ilya: Can you add citations here for
-  the suckiness of HM?}.  Additional techniques for the direct
-integration of evidence, also based on a kD tree decomposition of the
+variance and a possible bias \cite{NewtonRaftery:1994,Chib:1995,vanHaasteren:2009}.  Additional techniques for the direct integration of evidence, also based on a kD tree decomposition of the
 parameter space (see Sec.~\ref{sec:kDTree}), are described in

### 2010-12-19 paper.tex
Added references in Section 5.

@@
 Gravitational-wave astronomy provides the setting for another ongoing
-study.  Interesting triggers from the LIGO and Virgo
+study.  Interesting triggers from the LIGO \cite{InitLIGO} and Virgo \cite{Virgo}
 gravitational-wave detector pipeline that searches for compact binary
-coalescences are followed up with Bayesian parameter estimation tools.
+coalescences are followed up with Bayesian parameter estimation tools (see, e.g., \cite{Rover:2006ni, vanderSluys:2008b, VeitchVecchio:2009}).
 In general, several possible models for the data are considered,
@@
 components.  The models can include a large number of parameters (up
-to 15 for the model with two spinning components), and RJMCMC with
+to 15 for the model with two spinning components \cite{Raymond:2010}), and RJMCMC with
 interpolation via kD trees is again successful at computing the model
-evidences. \ilya{Add references throughout this paragraph.}
+evidences.
 

### 2010-12-19 paper.tex
Added comments on Sec. 3.

@@
 
+%\ilya{I think this section could just be phrased in terms of interpolation rather than jump proposal distributions; i.e., "the problem of finding the local density" instead of "the problem of drawing a proposed jump", etc.}
+
 The problem of drawing a proposed jump from an interpolation of
@@
 These neighborhoods should be non-overlapping and fill parameter
-space.  To draw a proposed jump, we select a point uniformly from the
+space.
+\ilya{I don't think this is generic enough: for example, Littenberg and Cornish also constructed an interpolation, but one in which multiple points could be in the same "neighborhood", while other neighborhoods could be empty.
+I.e., the algorithm below is one way of solving this problem, but not an alternative statement of it.}
+To draw a proposed jump, we select a point uniformly from the
 MCMC samples, find its associated neighborhood, and then draw the
@@
 this procedure produces proposed jumps that are approximately
-distributed according to the posterior PDF.  In fact, the proposed
+distributed according to the posterior PDF.
+\ilya{This last statement is presumably only true if the size of the neighborhood associated with each point is roughly the inverse of the local point density.
+I think it's worthwhile to point that out that your kD trees allow you to just that in a computationally efficient manner.}  In fact, the proposed
 jumps are drawn from a piecewise-constant (i.e.\ constant on each
@@
 median-finding algorithm.  Tree construction thus costs $\order{N \log
-  N}$ in time, and the tree consumes $\order{N \log N}$ space.  Box
+  N}$ in time, and the tree consumes $\order{N \log N}$ space.  As an illustration, box
 boundaries for a kD-tree constructed around a point set that is
@@
 can be accomplished in $\order{\log N}$ time and constant space with
-the following algorithm.  Given the point, $\vtheta_i$---which could
+the following algorithm.  Given the point, $\vtheta_i$ --- which could
 either be a point in the chain of MCMC samples if we are trying to
@@
 compute the jump probability distribution (see
-Eq.~\ref{eq:p-accept})---and the tree, $T$:
+Eq.~\ref{eq:p-accept}) --- and the tree, $T$:
 \begin{enumerate}

### 2011-01-06 paper.tex
Capitalized Monte Carlo in abstract.

@@
 Reversible-jump Markov chain
-  monte carlo (RJMCMC) is an extremely powerful technique for
+  Monte Carlo (RJMCMC) is an extremely powerful technique for
   performing Bayesian model selection, but it suffers from a

### 2011-01-26 paper.tex
Minor typos and tex formatting changes.

@@
 \begin{abstract}
-  Selection among alternative theoretical models given an observed data set is an important challenge in many areas of physics and astronomy, ranging from gravitational-wave data analysis to measurements of black-hole masses in X-ray binaries.
-Reversible-jump Markov chain
-  Monte Carlo (RJMCMC) is an extremely powerful technique for
-  performing Bayesian model selection, but it suffers from a
-  fundamental difficulty: it requires jumps between model parameter
-  spaces, but cannot retain a memory of the favored locations in more
-  than one parameter space at a time.  Thus, a naive jump between
-  parameter spaces is unlikely to be accepted in the MCMC algorithm
-  and convergence is correspondingly slow.  Here we demonstrate an
-  interpolation technique that uses samples from single-model MCMCs to
-  propose inter-model jumps from an approximation to the single-model
-  posterior of the target parameter space.  The interpolation
-  technique, based on a kD-tree data structure, is adaptive and
-  efficient in arbitrary dimensions.  We show that our technique leads
-  to dramatically improved convergence over naive jumps in an RJMCMC,
-  and compare it to other proposals in the literature to improve the
-  convergence of RJMCMCs.  We also discuss the use of the same
-  interpolation technique in two other contexts: as a convergence test
-  for a single-model MCMC and as a way to construct efficient
-  ``global'' proposal distributions for single-model MCMCs without
-  prior knowledge of the structure of the posterior distribution.
+  Selection among alternative theoretical models given an observed
+  data set is an important challenge in many areas of physics and
+  astronomy, ranging from gravitational-wave data analysis to
+  measurements of black-hole masses in X-ray binaries.
+  Reversible-jump Markov chain Monte Carlo (RJMCMC) is an extremely
+  powerful technique for performing Bayesian model selection, but it
+  suffers from a fundamental difficulty: it requires jumps between
+  model parameter spaces, but cannot retain a memory of the favored
+  locations in more than one parameter space at a time.  Thus, a naive
+  jump between parameter spaces is unlikely to be accepted in the MCMC
+  algorithm and convergence is correspondingly slow.  Here we
+  demonstrate an interpolation technique that uses samples from
+  single-model MCMCs to propose inter-model jumps from an
+  approximation to the single-model posterior of the target parameter
+  space.  The interpolation technique, based on a kD-tree data
+  structure, is adaptive and efficient in arbitrary dimensions.  We
+  show that our technique leads to dramatically improved convergence
+  over naive jumps in an RJMCMC, and compare it to other proposals in
+  the literature to improve the convergence of RJMCMCs.  We also
+  discuss the use of the same interpolation technique in two other
+  contexts: as a convergence test for a single-model MCMC and as a way
+  to construct efficient ``global'' proposal distributions for
+  single-model MCMCs without prior knowledge of the structure of the
+  posterior distribution.
 \end{abstract}
@@
 
-It should be noted that although an MCMC who jump acceptance criterium obeys detailed balance (as the Metropolis-Hastings algorithm does) must eventually converge to the desired distribution, there is no way to guarantee convergence in a fixed number of steps or to test whether a chain has converged in a foolproof manner.
-For example, MCMC chains can get stuck on local maxima, and if the autocorrelation length of the chain represents a substantial fraction of the total number of samples, the effective sample size may be too small to accurately represent the relative sizes of the modes in the PDF (however, see Section \ref{sec:examples} for one intriguing suggestion for remedying this issue).
+It should be noted that although an MCMC whose jump acceptance
+criterium obeys detailed balance (as the Metropolis-Hastings algorithm
+does) must eventually converge to the desired distribution, there is
+no way to guarantee convergence in a fixed number of steps or to test
+whether a chain has converged in a foolproof manner.  For example,
+MCMC chains can get stuck on local maxima, producing an apparently
+well-converged sampling of the PDF in the vicinity of the maximum; or,
+if the chain visits a sequence of local maxima, moving rarely between
+maxima, the autocorrelation length of the chain may represent a
+substantial fraction of the total number of samples, resulting in an
+effective sample size that is too small to accurately represent the
+relative sizes of the modes in the PDF (however, see Section
+\ref{sec:examples} for one intriguing suggestion for remedying this
+issue).
 
-Finally, we note that, in practice, the randomly chosen initial starting point of the MCMC may be in a particularly unlikely location in the parameter space.
-Because jumps are frequently local, we will generally want to ignore the first few autocorrelation lengths worth of points in a finite-size chain to avoid biases in the recovered posterior PDF due to the choice of the initial location.
-The points thus discarded are referred to as "burn-in" points.
+Finally, we note that, in practice, the randomly chosen initial
+starting point of the MCMC may be in a particularly unlikely location
+in the parameter space.  Because jumps are frequently local, we will
+generally want to ignore the first few autocorrelation lengths worth
+of points in a finite-size chain to avoid biases in the recovered
+posterior PDF due to the choice of the initial location.  The points
+thus discarded are referred to as "burn-in" points.
 
@@
 harmonic mean estimator for the evidence, which suffers from a large
-variance and a possible bias \cite{NewtonRaftery:1994,Chib:1995,vanHaasteren:2009}.  Additional techniques for the direct integration of evidence, also based on a kD tree decomposition of the
-parameter space (see Sec.~\ref{sec:kDTree}), are described in
-\cite{Weinberg2009}.  These techniques are promising, but in some
-cases suffer \cite{Farr2010} from large variance and bias as well.  An
-alternative approach to model selection among a set of models is based
-on performing an MCMC in a ``super-model'' that encompasses all of the
-models under consideration; this is known the the Reversible Jump
-Markov chain Monte Carlo (RJMCMC).
+variance and a possible bias
+\cite{NewtonRaftery:1994,Chib:1995,vanHaasteren:2009}.  Additional
+techniques for the direct integration of evidence, also based on a kD
+tree decomposition of the parameter space (see Sec.~\ref{sec:kDTree}),
+are described in \cite{Weinberg2009}.  These techniques are promising,
+but in some cases suffer \cite{Farr2010} from large variance and bias
+as well.  An alternative approach to model selection among a set of
+models is based on performing an MCMC in a ``super-model'' that
+encompasses all of the models under consideration; this is known the
+the Reversible Jump Markov chain Monte Carlo (RJMCMC).
 

### 2011-01-26 paper.tex
New references for kD trees, some additions and minor edits.

@@
 a local ``neighborhood'' to each point in the chain of MCMC samples.
-These neighborhoods should be non-overlapping and fill parameter
-space.
-\ilya{I don't think this is generic enough: for example, Littenberg and Cornish also constructed an interpolation, but one in which multiple points could be in the same "neighborhood", while other neighborhoods could be empty.
-I.e., the algorithm below is one way of solving this problem, but not an alternative statement of it.}
-To draw a proposed jump, we select a point uniformly from the
+We choose these neighborhoods to be non-overlapping and fill parameter
+space.  To draw a proposed jump, we select a point uniformly from the
 MCMC samples, find its associated neighborhood, and then draw the
@@
 this procedure produces proposed jumps that are approximately
-distributed according to the posterior PDF.
-\ilya{This last statement is presumably only true if the size of the neighborhood associated with each point is roughly the inverse of the local point density.
-I think it's worthwhile to point that out that your kD trees allow you to just that in a computationally efficient manner.}  In fact, the proposed
-jumps are drawn from a piecewise-constant (i.e.\ constant on each
-neighborhood) interpolation of the PDF.  There are various techniques
-that could be used to construct the set of neighborhoods associated
-with each point.
+distributed according to the posterior PDF.  In other words, the size
+of a neighborhood is inversely proportional to the local point
+density.  The proposed jumps are drawn from a piecewise-constant
+(i.e.\ constant on each neighborhood) interpolation of the PDF.  There
+are various techniques that could be used to construct the set of
+neighborhoods associated with each point.
 
@@
 larger difficulty with the Voronoi approach is that the computational
-cost for computing the Voronoi regions for $N$ points increases
-rapidly with dimension, $d$, as $N^X$ \will{Find X in the literature;
-  add references.  I think it's $X = d/2$}.
+cost for computing the Voronoi regions increases rapidly with
+dimension.  % Apparently the exact computational cost is not known in
+            % arbitrary dimension.
 
 We propose to use a decomposition of parameter space into
-neighborhoods based on a datastructure called a kD-tree
-\will{References.}.  The decomposition is more adaptive than the cells
-of Ref.~\cite{Littenberg2009}, and more efficient in high dimension
-than the Voronoi decomposition.
+neighborhoods based on a datastructure called a kD-tree (see, e.g.\
+\cite{Berg2008} or \cite{Gaede1998}).  The decomposition is more
+adaptive than the cells of Ref.~\cite{Littenberg2009}, and more
+efficient in high dimension than the Voronoi decomposition.
 
@@
 median-finding algorithm.  Tree construction thus costs $\order{N \log
-  N}$ in time, and the tree consumes $\order{N \log N}$ space.  As an illustration, box
-boundaries for a kD-tree constructed around a point set that is
-normally distributed around the origin in two dimensions are shown in
-Figure \ref{fig:kD-tree}.
+  N}$ in time, and the tree consumes $\order{N \log N}$ space.  As an
+illustration, box boundaries for a kD-tree constructed around a point
+set that is normally distributed around the origin in two dimensions
+are shown in Figure \ref{fig:kD-tree}.
 
@@
     interpolated PDF within a box of volume $V_i$ is $1/(N V_i)$,
-    where $N$ is the total number of points (and boxes).}
+    where $N$ is the total number of points (which is also the number
+    of boxes).}
 \end{figure}
 
-In order to use the kD-tree interpolation in an MCMC, we must be able
-to quickly find the neighborhood associated with a given point.  This
-can be accomplished in $\order{\log N}$ time and constant space with
-the following algorithm.  Given the point, $\vtheta_i$ --- which could
-either be a point in the chain of MCMC samples if we are trying to
-propose a jump, or a previously proposed jump if we are trying to
-compute the jump probability distribution (see
-Eq.~\ref{eq:p-accept}) --- and the tree, $T$:
+In order to use the kD-tree interpolation as a jump proposal in an
+MCMC, we must be able to quickly find the neighborhood associated with
+a given point to compute the jump probability (see
+Eq.~\ref{eq:p-accept}).  This can be accomplished in $\order{\log N}$
+time and constant space with the following algorithm, which is a
+standard binary tree search.  Given the point, $\vtheta_i$ and the
+tree, $T$:
 \begin{enumerate}

### 2011-03-16 paper.tex
Partially complete edit, pushing for Ilya.

@@
-\documentclass[prd,preprint]{revtex4}
+\documentclass[aps,prd,reprint]{revtex4-1}
 
@@
 
-\bibliographystyle{h-physrev}
+\bibliographystyle{apsrev4-1}
 
@@
   data set is an important challenge in many areas of physics and
-  astronomy, ranging from gravitational-wave data analysis to
-  measurements of black-hole masses in X-ray binaries.
+  astronomy.
+  %ranging from gravitational-wave data analysis to
+  %measurements of black-hole masses in X-ray binaries.
   Reversible-jump Markov chain Monte Carlo (RJMCMC) is an extremely
@@
 
-\will{Mention Weinberg direct integration (Ref.~\cite{Weinberg2009}).
-  Also \cite{Ascasibar2005} in the cosmological context (not really
-  kD-tree, but close).}
+Selection among alternative theoretical models given an observed data
+set is an important challenge in many areas of physics and astronomy.
+In a Bayesian context, model selection involves computing the evidence
+for each model given the data.  The model evidence is an integral of
+the unnormalized posterior probability distribution over the model
+parameter space, representing the ``total probability'' in that model.
+Models with larger evidence are preferred; the ratio of the evidences
+of two models is the Bayes ratio between them.  The product of the
+Bayes ratio and the ratio of prior probabilities for the model gives
+the odds ratio for the models, the ratio of Bayesian probabilities for
+the two models given the observation data.
+
+There are many ways to compute model evidences.  In low-dimensional
+parameter spaces, the unnormalized posterior probability can be
+evaluated on a grid or lattice and the integral can be performed
+directly.  For many problems or models of interest, however, the
+dimensionality of parameter space is too large to make this approach
+practical.  Nested sampling \cite{Skilling:2006,Feroz:2009} \ilya{More
+  references?} is a stochastic method for exploring the parameter
+space of a model while computing the evidence integral statistically
+\ilya{I want a better word than ``statistically'' to indicate that
+  it's doing something akin to Monte-Carlo integration (but I would
+  like to reserve Monte-Carlo for MCMC).}.
+
+\ilya{Something about how nested sampling distributes points according
+  to the prior, not the posterior, and therefore can miss modes that
+  are disfavored by prior.  Also, it does not exploit any information
+  about the posterior to make the evidence calculation more
+  efficient.}
+
+In contrast to nested sampling, where computation of evidences is the
+primary objective, Markov Chain Monte Carlo (MCMC) methods attempt to
+stochastically produce parameter samples with density proportional to
+the posterior probability distribution.  In MCMC techniques, the
+primary target is an accurate estimate of the posterior distribution.
+It is not straightforward to compute the model evidence from the MCMC
+samples.  The most direct way to estimate the evidence for a model
+from MCMC samples is to compute the harmonic-mean estimatator, but
+this estimator of the evidence suffers from infinite
+variance\cite{NewtonRaftery:1994,Chib:1995,vanHaasteren:2009}; various
+methods have been proposed to stabilize it \will{Citation}, but all
+require some assumptions about the shape of the posterior distribution
+that may not be satisfied in practice.
+
+Reference \cite{Weinberg2009} gives a method for directly computing
+the evidence integral using a kD-tree datastructure to decompose a
+parameter space into boxes containing the MCMC sample points.  The
+integral is approximated as a sum over box volumes.  This method is
+promising, but it is not clear in general what statistical and
+systematic errors it introduces and how these are affected by the
+shape of the posterior distribution from which the MCMC samples.
+
+When the goal is model selection between several known models, only
+the \emph{relative} evidence of each model is needed.  In this
+circumstance, the Reversible Jump MCMC technique first introduced in
+Reference \cite{Green1995} is one of the most reliable and accurate
+ways to compare the models.  Reversible Jump MCMC (RJMCMC), described
+more fully in Section \ref{sec:reversible-jump}, performs a standard
+MCMC in a superspace that is a direct sum of all the model parameter
+spaces.  Such an MCMC involves both intra- and inter-model jumps; the
+number of MCMC samples in each model's parameter space is proportional
+to that model's relative evidence in the suite of models being
+compared.
+
+Implemented naively, RJMCMC has a significant drawback: because the
+chain of samples must be Markovian, only the current sample is
+available to the algorithm.  Each time an RJMCMC transitions between
+models, the information about the good parameter values in the
+previous model is lost; subsequent jumps into that model must ``start
+fresh,'' and are correspondingly unlikely to be accepted, delaying
+convergence of the RJMCMC sample chain.
+
+Here we demonstrate a technique for improving the acceptance ratio of
+inter-model jumps in an RJMCMC, leading to dramatically improved
+convergence of RJMCMC sample chains.  The technique uses a kD-tree
+datastructure to construct an approximation to each model's posterior
+parameter distribution.  We propose jumps into each model drawn from
+the approximation to the posterior.  Because jumps are proposed
+preferentially to locations favored by the single-model posterior, the
+RJMCMC compares ``good'' locations in parameter space across all the
+models, and convergence is generally rapid.  The approximation to the
+posterior distribution produced by the kD-tree is a constant-in-box
+interpolation of the posterior, similar in spirit to the phase-space
+density interpolants produced from N-body positions and momenta in
+Refernce \cite{Ascasibar2005}.  The kD-tree interpolation is effective
+in arbitrary numbers of dimension, and is quite space-efficient,
+requiring $\order{N \log N}$ storage space and $\order{\log N}$ time
+to produce each proposed jump, where $N$ is the number of single-model
+MCMC samples used to construct the interpolation.  Reference
+\cite{Littenberg2009} proposed an alternate method for producing
+inter-model jumps in an RJMCMC that relies on a box decomposition of
+parameter space, but uses fixed-sized boxes; because the box size is
+fixed, the method cannot adapt to the local structure of the
+posterior, and therefore becomes asymptotically less efficient than
+the adaptive kD-tree decomposition in either acceptance ratio or
+storage space as the parameter-space dimension increases or the
+posterior becomes more peaked.  \ilya{Mention something (with
+  reference) about the 10-way RJMCMC in the black-hole mass paper?}
+
+The structure of this paper is as follows: in Section
+\ref{sec:reversible-jump} we introduce in more detail the concept of a
+Reversible Jump MCMC, and describe the fundamental difficulty with a
+naive jump proposal in an RJMCMC.  In Section \ref{sec:kDTree} we
+introduce the kD-tree datastructure used to decompose parameter space
+into boxes for the interpolation.  In Section \ref{sec:efficiency} we
+demonstrate the efficiency gains that are achieved from use of the
+interpolated jump proposal.  In Section \ref{sec:examples} we give
+examples of some other uses of the interpolated jump proposal in a
+single-model MCMC context.  Finally, in Section \ref{sec:conclusion}
+we offer a summary and some concluding remarks on the method.
 
 \section{Reversible Jump MCMC}
+\label{sec:reversible-jump}
 
-Reversible jump Markov chain Monte Carlo (RJMCMC) \cite{Green1995} is a technique for Bayesian model comparison.
-Below, we give a very brief introduction to Bayesian analysis; describe a standard MCMC; and introduce RJMCMC.
+Reversible jump Markov chain Monte Carlo (RJMCMC) \cite{Green1995} is
+a technique for Bayesian model comparison.  Below, we give a very
+brief introduction to Bayesian analysis; describe a standard MCMC; and
+introduce RJMCMC.
 
@@
 
-A Markov chain Monte Carlo \cite{Gilks:1996} produces a set of samples $\{ \vtheta^{(j)} \, | \, i = 1, \ldots \}$ from the model parameter space that are sampled according to the posterior, meaning that, in the limit that the chain length tends to infinity, the relative frequency with which a given set of parameter appears in the chain is proportional to the desired posterior, $p(\vtheta|d,M)$.
-Therefore, the output of an MCMC can be directly interpreted as the posterior PDF over the full parameter space, while PDFs for individual parameters can be obtained through trivially marginalizing over the uninteresting parameters, simply by considering the distribution of the parameter of interest in the MCMC samples.
+A Markov chain Monte Carlo \cite{Gilks:1996} produces a set of samples
+$\{ \vtheta^{(j)} \, | \, i = 1, \ldots \}$ from the model parameter
+space that are sampled according to the posterior, meaning that, in
+the limit that the chain length tends to infinity, the relative
+frequency with which a given set of parameter appears in the chain is
+proportional to the desired posterior, $p(\vtheta|d,M)$.  Therefore,
+the output of an MCMC can be directly interpreted as the posterior PDF
+over the full parameter space, while PDFs for individual parameters
+can be obtained by marginalizing over the uninteresting parameters.
 
-A Markov chain has the property that the probability distribution of the next state can depend only on the current state, not on the past history:
+A Markov chain has the property that the probability distribution of
+the next state can depend only on the current state, not on the past
+history: 
+%
 \be 
-p(\vtheta^{(j+1)})=\int_{V} d\vtheta^{(j)} p(\vtheta^{(j)} \to \vtheta^{(j+1)}) p(\vtheta^{(j)}),
-\ee
-where the jump probability $p(\vtheta^{(j)} \to \vtheta^{(j+1)})$ depends only on $\vtheta^{(j)}$ and $\vtheta^{(j+1)}$. 
-An additional requirement for an MCMC arises from the fact that the desired distribution is the equilibrium distribution.
-In other words, if we assume that state $(j)$ of the chain is sampled from the desired PDF, $p(\vtheta^{(j)})=p(\vtheta|d,M)$ , then the next state $(j+1)$ must be sampled from the PDF as well, so that $p(\vtheta^{(j+1)})=p(\vtheta|d,M)$.
+p(\vtheta^{(j+1)})=\int_{V} d\vtheta^{(j)} p(\vtheta^{(j)} \to
+\vtheta^{(j+1)}) p(\vtheta^{(j)}), \ee
+%
+where the jump probability $p(\vtheta^{(j)} \to \vtheta^{(j+1)})$
+depends only on $\vtheta^{(j)}$ and $\vtheta^{(j+1)}$.  An additional
+requirement for an MCMC arises from the fact that the desired
+distribution is the equilibrium distribution.  In other words, if we
+assume that state $(j)$ of the chain is sampled from the desired PDF,
+$p(\vtheta^{(j)})=p(\vtheta|d,M)$ , then the next state $(j+1)$ must
+be sampled from the PDF as well, so that
+$p(\vtheta^{(j+1)})=p(\vtheta|d,M)$; this condition is known as
+``detailed balance.''
 
-One way to produce such a sequence of samples is via the Metropolis-Hastings algorithm, first proposed by Metropolis et al.~in 1953 \cite{Metropolis:1953}, and later generalized by Hastings:
+One way to produce such a sequence of samples is via the
+Metropolis-Hastings algorithm, first proposed in Reference
+\cite{Metropolis:1953}, and later generalized by Hastings
+\ilya{Citation?}:
 \begin{enumerate}
-  \item Given a current state $\vtheta^{(j)}$, propose the next state $\vtheta^p$ by drawing from a jump proposal distribution with probability $Q(\vtheta^{(j)} \to \vtheta^p)$.  
+\item Given a current state $\vtheta^{(j)}$, propose the next state
+  $\vtheta^p$ by drawing from a jump proposal distribution with
+  probability $Q(\vtheta^{(j)} \to \vtheta^p)$.
 \item Compute the probability of accepting the proposed jump as
   \bel{eq:p-accept} 
-p_{\textnormal{accept}}  \equiv \min\Bigl(1,  \frac{p(\vtheta^p|d)}{p(\vtheta^{(j)}|d)} \frac{Q(\vtheta^p \to
+p_{\textnormal{accept}} \equiv \min\Bigl(1,
+\frac{p(\vtheta^p|d)}{p(\vtheta^{(j)}|d)} \frac{Q(\vtheta^p \to
   \vtheta^{(j)})}{Q(\vtheta^{(j)} \to \vtheta^p)} \Bigr).  
 \ee
-\item Pick a uniform random number $\alpha \in [0,1]$.
-If $\alpha<  p_{\textnormal{accept}}$, accept the proposed jump, setting $\vtheta^{(j+1)}=\vtheta^p$.
-Otherwise, reject the jump, and remain at the same location in parameter space for the next step, $\vtheta^{(j+1)}=\vtheta^{(j)}$.
+\item Pick a uniform random number $\alpha \in [0,1]$.  If $\alpha<
+  p_{\textnormal{accept}}$, accept the proposed jump, setting
+  $\vtheta^{(j+1)}=\vtheta^p$.  Otherwise, reject the jump, and remain
+  at the same location in parameter space for the next step,
+  $\vtheta^{(j+1)}=\vtheta^{(j)}$.
 \end{enumerate}
 
-This jump proposal distribution $Q(\vtheta^{(j)} \to \vtheta^p)$ can depend on the parameters of the current state $\vtheta^{(j)}$, but not on the past history.
-It must also allow any state within the allowed prior volume to be reachable by the MCMC.
-Beyond these conditions, there is obviously a great deal of freedom in choosing the jump proposal distribution.
-This is the most important choice in the MCMC, as it determines the sampling efficiency of the algorithm, i.e., the necessary length of the chain before it converges to the posterior PDF.
-Creating an efficient jump proposal distribution requires an understanding of the structure of the parameter space which may not be available until the PDFs are found, creating a Catch-22; one possibility for resolving this infinite loop is described in Section \ref{sec:examples}.
+This jump proposal distribution $Q(\vtheta^{(j)} \to \vtheta^p)$ can
+depend on the parameters of the current state $\vtheta^{(j)}$, but not
+on the past history.  It must also allow any state within the allowed
+prior volume to be reachable by the MCMC.  Any jump proposal that
+satisfies these properties is suitable for an MCMC.
+
+The jump proposal is the most important choice in the MCMC, as it
+determines the sampling efficiency of the algorithm, i.e., the
+necessary length of the chain before it converges to the posterior
+PDF.  Creating an efficient jump proposal distribution requires an
+understanding of the structure of the parameter space which may not be
+available until the PDFs are found, creating a Catch-22; one
+possibility for resolving this infinite loop is described in Section
+\ref{sec:examples}.
 
@@
 in the parameter space.  Because jumps are frequently local, we will
-generally want to ignore the first few autocorrelation lengths worth
-of points in a finite-size chain to avoid biases in the recovered
-posterior PDF due to the choice of the initial location.  The points
-thus discarded are referred to as "burn-in" points.
+generally want to ignore the early points in a finite-size chain to
+avoid biases in the recovered posterior PDF due to the choice of the
+initial location.  The points thus discarded are referred to as
+"burn-in" points.
 
@@
 
-The samples produced during by an MCMC algorithm can be used to
-directly perform a Monte Carlo evidence integral.  This results in a
-harmonic mean estimator for the evidence, which suffers from a large
-variance and a possible bias
-\cite{NewtonRaftery:1994,Chib:1995,vanHaasteren:2009}.  Additional
-techniques for the direct integration of evidence, also based on a kD
-tree decomposition of the parameter space (see Sec.~\ref{sec:kDTree}),
-are described in \cite{Weinberg2009}.  These techniques are promising,
-but in some cases suffer \cite{Farr2010} from large variance and bias
-as well.  An alternative approach to model selection among a set of
-models is based on performing an MCMC in a ``super-model'' that
-encompasses all of the models under consideration; this is known the
-the Reversible Jump Markov chain Monte Carlo (RJMCMC).
+The samples produced by an MCMC algorithm can be used to directly
+perform a Monte Carlo evidence integral.  This results in a harmonic
+mean estimator for the evidence, which suffers from a large variance
+and bias \cite{NewtonRaftery:1994,Chib:1995,vanHaasteren:2009}.
+Additional techniques for the direct integration of evidence, also
+based on a kD tree decomposition of the parameter space (see
+Sec.~\ref{sec:kDTree}), are described in \cite{Weinberg2009}.  These
+techniques are promising, but in some cases suffer \cite{Farr2010}
+from large variance and bias as well.  An alternative approach to
+model selection among a set of models is based on performing an MCMC
+in a ``super-model'' that encompasses all of the models under
+consideration; this is known the the Reversible Jump Markov chain
+Monte Carlo (RJMCMC).
 
@@
 model as a parameter, with other parameters ignored, is obtained by
-marginalizing over the remaining parameters, i.e., the evidence for a
-given model is just
+marginalizing over the remaining parameters.  If we assume uniform
+priors on the models, the evidence for a given model is proportional
+to the number of counts 
 %
@@
 \section{RJMCMC Efficiency}
+\label{sec:efficiency}
 
@@
 
-\section{Examples and other uses} \label{sec:examples}
+\section{Examples and other uses} 
+\label{sec:examples}
 
@@
 \section{Conclusion/Discussion}
-
-\nocite{Littenberg2009}
+\label{sec:conclusion}
 

### 2011-03-16 paper.tex
Revtex 4-1 or revtex 4.
Added references, updated intro.

@@
-\documentclass[aps,prd,reprint]{revtex4-1}
+\IfFileExists{revtex4-1.cls}{\documentclass[prd, twocolumn, lengthcheck,
+superscriptaddress, showpacs, letterpaper, nofootinbib]{revtex4-1}}{ 
+\IfFileExists{revtex4.cls}{\documentclass[prd, preprint,
+superscriptaddress, showpacs, letterpaper, nofootinbib]{revtex4}}{}
+}
 
@@
 
-\bibliographystyle{apsrev4-1}
+\IfFileExists{apsrev4-1.bst}{\bibliographystyle{apsrev4-1}}
+\IfFileExists{apsrev4.bst}{\bibliographystyle{apsrev4}}
 
@@
 the unnormalized posterior probability distribution over the model
-parameter space, representing the ``total probability'' in that model.
+parameter space, representing the probability of obtaining the data set within that model.
 Models with larger evidence are preferred; the ratio of the evidences
-of two models is the Bayes ratio between them.  The product of the
-Bayes ratio and the ratio of prior probabilities for the model gives
-the odds ratio for the models, the ratio of Bayesian probabilities for
-the two models given the observation data.
+of two models is the Bayes factor between them.  The product of the
+Bayes factor and the ratio of prior probabilities for the two models yields
+the odds ratio for the models.
+%, the ratio of Bayesian probabilities for the two models given the observation data.
 
@@
 dimensionality of parameter space is too large to make this approach
-practical.  Nested sampling \cite{Skilling:2006,Feroz:2009} \ilya{More
-  references?} is a stochastic method for exploring the parameter
-space of a model while computing the evidence integral statistically
-\ilya{I want a better word than ``statistically'' to indicate that
-  it's doing something akin to Monte-Carlo integration (but I would
-  like to reserve Monte-Carlo for MCMC).}.
+practical, and stochastic sampling must be used.
 
-\ilya{Something about how nested sampling distributes points according
-  to the prior, not the posterior, and therefore can miss modes that
-  are disfavored by prior.  Also, it does not exploit any information
-  about the posterior to make the evidence calculation more
-  efficient.}
+%\ilya{Something about how nested sampling distributes points according to the prior, not the posterior, and therefore can miss modes that are disfavored by prior.
+Also, it does not exploit any information about the posterior to make the evidence calculation more efficient.}
 
-In contrast to nested sampling, where computation of evidences is the
-primary objective, Markov Chain Monte Carlo (MCMC) methods attempt to
+Markov Chain Monte Carlo (MCMC) methods attempt to
 stochastically produce parameter samples with density proportional to
-the posterior probability distribution.  In MCMC techniques, the
-primary target is an accurate estimate of the posterior distribution.
+the posterior probability distribution.
+In MCMC techniques, the primary target is an accurate estimate of the posterior distribution.
+(We note that an alternative stochastic method for exploring a model parameter space, nested sampling \cite{Skilling:2004, Skilling:2006,Feroz:2009}, focuses on evidence computation rather than sampling the posterior probability density functions.)  \will{Will, are you happy with nested sampling in a parenthetical expression, without further comments?}
 It is not straightforward to compute the model evidence from the MCMC
 samples.  The most direct way to estimate the evidence for a model
-from MCMC samples is to compute the harmonic-mean estimatator, but
+from MCMC samples is to compute the harmonic-mean estimator, but
 this estimator of the evidence suffers from infinite
-variance\cite{NewtonRaftery:1994,Chib:1995,vanHaasteren:2009}; various
-methods have been proposed to stabilize it \will{Citation}, but all
+variance\cite{NewtonRaftery:1994,Chib:1995,vanHaasteren:2009}.  Various
+methods have been proposed to stabilize the harmonic-mean estimator \will{Citation}, but all
 require some assumptions about the shape of the posterior distribution
 that may not be satisfied in practice.
+MCMC implementations with parallel tempering \cite{EarlDeem:2005} allow for evidence computation via thermodynamic integration, but these can be computationally costly.
 
 Reference \cite{Weinberg2009} gives a method for directly computing
-the evidence integral using a kD-tree datastructure to decompose a
+the evidence integral from existing MCMC samples by using a kD-tree data structure to decompose a
 parameter space into boxes containing the MCMC sample points.  The
@@
 available to the algorithm.  Each time an RJMCMC transitions between
-models, the information about the good parameter values in the
+models, the information about the choices of parameter values in the
 previous model is lost; subsequent jumps into that model must ``start
@@
 
-Here we demonstrate a technique for improving the acceptance ratio of
+Here we introduce a technique for improving the acceptance ratio of
 inter-model jumps in an RJMCMC, leading to dramatically improved
 convergence of RJMCMC sample chains.  The technique uses a kD-tree
-datastructure to construct an approximation to each model's posterior
-parameter distribution.  We propose jumps into each model drawn from
-the approximation to the posterior.  Because jumps are proposed
+data structure to construct an approximation to each model's posterior
+parameter distribution.  We draw jump proposals into the model from this approximation to its posterior.  Because jumps are proposed
 preferentially to locations favored by the single-model posterior, the
 RJMCMC compares ``good'' locations in parameter space across all the
-models, and convergence is generally rapid.  The approximation to the
+models, and convergence is generally rapid.
+We have successfully applied this RJMCMC technique to a 10-way model selection among alternative mass distribution models for black-hole X-ray binaries \cite{Farr2010}.
+
+Reference \cite{Littenberg2009} proposed an alternate method for producing
+inter-model jumps in an RJMCMC that relies on a box decomposition of
+parameter space, but uses fixed-sized boxes.  The box size is
+fixed, so the method cannot adapt to the local structure of the
+posterior, and becomes asymptotically inefficient for high-dimensional parameter spaces or highly peaked posteriors.  Meanwhile, the approximation to the
 posterior distribution produced by the kD-tree is a constant-in-box
@@
 density interpolants produced from N-body positions and momenta in
-Refernce \cite{Ascasibar2005}.  The kD-tree interpolation is effective
-in arbitrary numbers of dimension, and is quite space-efficient,
+Reference \cite{Ascasibar2005}.  The kD-tree interpolation is effective
+in parameter spaces of arbitrary dimensionality, and is quite space-efficient,
 requiring $\order{N \log N}$ storage space and $\order{\log N}$ time
 to produce each proposed jump, where $N$ is the number of single-model
-MCMC samples used to construct the interpolation.  Reference
-\cite{Littenberg2009} proposed an alternate method for producing
-inter-model jumps in an RJMCMC that relies on a box decomposition of
-parameter space, but uses fixed-sized boxes; because the box size is
-fixed, the method cannot adapt to the local structure of the
-posterior, and therefore becomes asymptotically less efficient than
-the adaptive kD-tree decomposition in either acceptance ratio or
-storage space as the parameter-space dimension increases or the
-posterior becomes more peaked.  \ilya{Mention something (with
-  reference) about the 10-way RJMCMC in the black-hole mass paper?}
+MCMC samples used to construct the interpolation.
 
-The structure of this paper is as follows: in Section
+The structure of this paper is as follows.  In Section
 \ref{sec:reversible-jump} we introduce in more detail the concept of a
@@
 naive jump proposal in an RJMCMC.  In Section \ref{sec:kDTree} we
-introduce the kD-tree datastructure used to decompose parameter space
-into boxes for the interpolation.  In Section \ref{sec:efficiency} we
+introduce the kD-tree data structure used to decompose the parameter space
+into boxes for interpolation.  In Section \ref{sec:efficiency} we
 demonstrate the efficiency gains that are achieved from use of the
 interpolated jump proposal.  In Section \ref{sec:examples} we give
-examples of some other uses of the interpolated jump proposal in a
-single-model MCMC context.  Finally, in Section \ref{sec:conclusion}
+examples of some other uses of the interpolated jump proposal, and suggest its utility in the context of a
+single-model MCMC.  Finally, in Section \ref{sec:conclusion}
 we offer a summary and some concluding remarks on the method.
@@
 techniques for computing posterior PDFs in the context of a model is
-the Markov Chain Monte Carlo, which we now describe.
+the Markov chain Monte Carlo, which we now describe.
 
-\subsection{MCMC} \label{sec:mcmc}
+\subsection{Markov chain Monte Carlo} \label{sec:mcmc}
 
@@
 Metropolis-Hastings algorithm, first proposed in Reference
-\cite{Metropolis:1953}, and later generalized by Hastings
-\ilya{Citation?}:
+\cite{Metropolis:1953}, and later generalized by Hastings \cite{Hastings:1970}:
 \begin{enumerate}

### 2011-03-16 paper.tex
Merged Ilya's changes to intro.
First few paragraphs of IV.

@@
 assume that state $(j)$ of the chain is sampled from the desired PDF,
-$p(\vtheta^{(j)})=p(\vtheta|d,M)$ , then the next state $(j+1)$ must
+$p(\vtheta^{(j)})=p(\vtheta^{(j)}|d,M)$ , then the next state $(j+1)$ must
 be sampled from the PDF as well, so that
-$p(\vtheta^{(j+1)})=p(\vtheta|d,M)$; this condition is known as
+$p(\vtheta^{(j+1)})=p(\vtheta^{(j+1)}|d,M)$; this condition is known as
 ``detailed balance.''
@@
 depend on the parameters of the current state $\vtheta^{(j)}$, but not
-on the past history.  It must also allow any state within the allowed
-prior volume to be reachable by the MCMC.  Any jump proposal that
-satisfies these properties is suitable for an MCMC.
+on the past history.  It must also allow any state within the prior
+volume to be reachable (eventually) by the MCMC.  Any jump proposal
+that satisfies these properties is suitable for an MCMC.
 
 The jump proposal is the most important choice in the MCMC, as it
-determines the sampling efficiency of the algorithm, i.e., the
-necessary length of the chain before it converges to the posterior
-PDF.  Creating an efficient jump proposal distribution requires an
-understanding of the structure of the parameter space which may not be
-available until the PDFs are found, creating a Catch-22; one
-possibility for resolving this infinite loop is described in Section
-\ref{sec:examples}.
+determines the sampling efficiency of the algorithm, i.e., the length
+of the chain before it converges to the posterior PDF.  Creating an
+efficient jump proposal distribution requires an understanding of the
+structure of the parameter space which may not be available until the
+PDFs are found, creating a Catch-22; one possibility for resolving
+this infinite loop is described in Section \ref{sec:examples}.
 
@@
 MCMC in the ``super-model'' parameter space just like a regular MCMC;
-in general, we can both propose jumps to different parameters within a
-model (intramodel jumps) and jumps to a different model with different
-parameters (intermodel jumps).  The resulting chain samples from the
-posterior $p(M_i,\vtheta_i|d)$.  As in a usual MCMC, the PDF on the
-model as a parameter, with other parameters ignored, is obtained by
-marginalizing over the remaining parameters.  If we assume uniform
-priors on the models, the evidence for a given model is proportional
-to the number of counts 
+we propose jumps to different parameters within a model (intramodel
+jumps) and jumps to a different model with different parameters
+(intermodel jumps).  The resulting chain samples from the posterior
+$p(M_i,\vtheta_i|d)$.  As in a usual MCMC, the PDF on the model as a
+parameter, with other parameters ignored, is obtained by marginalizing
+over the remaining parameters.  The posterior probability of a model
+is proportional to the number of counts
 %
 \be
-p(M_i|d) = \int d\vtheta_i p(M_i, \vtheta_i|d) \approx \frac{N_i}{N},
+%
+p(M_i|d) = \int d\vtheta_i \frac{p(d|M_i, \vtheta_i) p(M_i, \vtheta_i)}{p(d)}
+\approx \frac{N_i}{N},
+%
 \ee
@@
 this procedure produces proposed jumps that are approximately
-distributed according to the posterior PDF.  In other words, the size
-of a neighborhood is inversely proportional to the local point
-density.  The proposed jumps are drawn from a piecewise-constant
-(i.e.\ constant on each neighborhood) interpolation of the PDF.  There
-are various techniques that could be used to construct the set of
-neighborhoods associated with each point.
+distributed according to the posterior PDF.  The size of a
+neighborhood is inversely proportional to the local point density.
+The proposed jumps are drawn from a piecewise-constant (constant on
+each neighborhood) interpolation of the PDF.  There are various
+techniques that could be used to construct the set of neighborhoods
+associated with each point.
 
@@
 in the form of one ``fictitious'' point per brick.  The extra uniform
-distribution from the fictituous points represents a significant cost,
-since inter-model jumps proposed from the uniform distribution are
-unlikely to be accepted.  Additionally, the bricks in this algorithm
-do not adapt to the local structure of the PDF.  One must either use
-small bricks to capture the local structure of the PDF, placing many
-bricks in regions without MCMC samples, or use large bricks, missing
-the local strucutre of the PDF in exchange for fewer empty bricks.
+distribution from the fictituous points can represent a significant
+cost, since inter-model jumps proposed from the uniform distribution
+are unlikely to be accepted.  Additionally, the bricks in this
+algorithm do not adapt to the local structure of the PDF.  One must
+either use small bricks to capture the local structure of the PDF,
+placing many bricks in regions without MCMC samples, or use large
+bricks, missing the local strucutre of the PDF in exchange for fewer
+empty bricks.
 
@@
 Unfortunately, defining the Voronoi regions requires a metric on
-parameter space, which may be difficult or impossible to define.  A
-larger difficulty with the Voronoi approach is that the computational
-cost for computing the Voronoi regions increases rapidly with
+parameter space, which may be difficult or impossible to define.
+Also, the computational cost for computing the Voronoi regions
+increases rapidly with
 dimension.  % Apparently the exact computational cost is not known in
@@
 
-We propose to use a decomposition of parameter space into
+Here we propose to use a decomposition of parameter space into
 neighborhoods based on a datastructure called a kD-tree (see, e.g.\
@@
   \begin{center}
-    \includegraphics[width=0.8\textwidth]{kdtree.png}
+    \includegraphics[width=0.8\columnwidth]{kdtree.png}
   \end{center}
@@
 
+In this section, we demonstrate the efficiency of the kD-interpolated
+jump proposal on a toy model comparison problem.  We draw $N = 100$
+simulated data points from a $N(0,1)$ Gaussian distribution, and then
+ask whether these data are better described by a model where they are
+Gaussian distributed with unkonwn mean, $\mu$, and standard deviation,
+$\sigma$
+%
+\be
+p(x) = \frac{1}{\sqrt{2\pi} \sigma} \exp\left( - \frac{(x-\mu)^2}{2
+    \sigma^2} \right),
+\ee
+%
+or by a model where they are Cauchy distributed with mode $\alpha$,
+and width $\beta$
+%
+\be
+p(x) = \frac{1}{\pi \beta \left( 1 + \left(\frac{x - \alpha}{\beta}\right)^2\right)}.
+\ee
+%
+We take priors on $\mu$ and $\alpha$ to be uniform in $[-1,1]$, and
+priors in $\sigma$ and $\beta$ to be uniform in $[0.5, 1.5]$.  With a
+data set of 100 points, the relative uncertainty in determining the
+parameters of the underlying distribution is approximately 10\%, so we
+expect the posterior probabilities in the $(\mu,\sigma)$ and
+$(\alpha,\beta)$ spaces to occupy roughly 1\% of the prior volume.
+
 \begin{figure}
   \begin{center}
-    \includegraphics[width=0.8\textwidth]{acceptRate}
+    \includegraphics[width=0.8\columnwidth]{acceptRate}
   \end{center}

### 2011-03-16 paper.tex
Minor edits to Section IV.

@@
 expect the posterior probabilities in the $(\mu,\sigma)$ and
-$(\alpha,\beta)$ spaces to occupy roughly 1\% of the prior volume.
+$(\alpha,\beta)$ spaces to occupy only a few percent of the prior
+volume.  The Cauchy distribution is so much broader than the Gaussian
+(it has no finite moments), that with uniform model proirs, the
+posterior probability for the Gaussian model over the Cauchy model is
+extremely large:
+%
+\be
+\frac{p(\textnormal{Gaussian} | d)}{p(\textnormal{Cauchy}|d)} > 10^9
+\ee
+%
 

### 2011-03-16 paper.tex
Minor edits to IV.

@@
 In this section, we demonstrate the efficiency of the kD-interpolated
-jump proposal on a toy model comparison problem.  We draw $N = 100$
+jump proposal on a toy model-comparison problem.  We draw $N = 100$
 simulated data points from a $N(0,1)$ Gaussian distribution, and then
 ask whether these data are better described by a model where they are
-Gaussian distributed with unkonwn mean, $\mu$, and standard deviation,
+Gaussian distributed with unknown mean $\mu$ and standard deviation
 $\sigma$
@@
 p(x) = \frac{1}{\sqrt{2\pi} \sigma} \exp\left( - \frac{(x-\mu)^2}{2
-    \sigma^2} \right),
+    \sigma^2} \right)\ ,
 \ee
 %
-or by a model where they are Cauchy distributed with mode $\alpha$,
+or by a model where they are Cauchy distributed with mode $\alpha$
 and width $\beta$
@@
 \be
-p(x) = \frac{1}{\pi \beta \left( 1 + \left(\frac{x - \alpha}{\beta}\right)^2\right)}.
+p(x) = \frac{1}{\pi \beta \left( 1 + \left(\frac{x - \alpha}{\beta}\right)^2\right)}\ .
 \ee
@@
 $(\alpha,\beta)$ spaces to occupy only a few percent of the prior
-volume.  The Cauchy distribution is so much broader than the Gaussian
-(it has no finite moments), that with uniform model proirs, the
+volume.  The Cauchy distribution is much broader than the Gaussian
+(it has no finite moments), so with uniform model priors, the
 posterior probability for the Gaussian model over the Cauchy model is
@@
 \be
-\frac{p(\textnormal{Gaussian} | d)}{p(\textnormal{Cauchy}|d)} > 10^9
+\frac{p(\textnormal{Gaussian} | d)}{p(\textnormal{Cauchy}|d)} > 10^9\ .
 \ee

### 2011-03-17 paper.tex
Finished editing III.

@@
 techniques focused directly on evidence computation (e.g., nested
-sampling \cite{Skilling:2006} and MultiNest \cite{Feroz:2009}).  However, we
-frequently want to compute the posterior PDFs within each model along
-with the evidences for the various models.  One of the most common
-techniques for computing posterior PDFs in the context of a model is
+sampling \cite{Skilling:2004,Skilling:2006} and its variant MultiNest \cite{Feroz:2009}).  Although nested sampling can be used to compute the posterior PDFs within each model along
+with the evidences for the various models, perhaps the most common
+technique for computing posterior PDFs in the context of a model is
 the Markov chain Monte Carlo, which we now describe.
@@
 initial location.  The points thus discarded are referred to as
-"burn-in" points.
+``burn-in'' points.
 
@@
 Sec.~\ref{sec:kDTree}), are described in \cite{Weinberg2009}.  These
-techniques are promising, but in some cases suffer \cite{Farr2010}
+techniques are promising, but in some cases suffer \cite{Farr2010} \will{Is this really the reference you meant?}
 from large variance and bias as well.  An alternative approach to
@@
 
-Ref.~\cite{Littenberg2009} decomposed parameter space into
+Ref.~\cite{Littenberg2009} decomposed the parameter space into
 constant-volume ``bricks'' whose size was set by the typical size of
-the peaks of the PDF.  Each point was associated with the brick that
-contained it.  Unfortunately, this process does not guarantee that
-every brick contains at least one point, which is required for the
-jump proposal to cover the entire model parameter space.  The solution
-to this problem was to add a uniform distribution to the proposal PDF,
-in the form of one ``fictitious'' point per brick.  The extra uniform
-distribution from the fictituous points can represent a significant
-cost, since inter-model jumps proposed from the uniform distribution
-are unlikely to be accepted.  Additionally, the bricks in this
+the peaks of the PDF.  Each sample was associated with the brick that
+contained it, and the probability of proposing a jump into a particular brick is thus proportional to the
+number of samples within that brick.  Additionally, an extra uniform jump proposal was added to allow for jumps into bricks that did not contain any points, so that the jump 
+proposal covers the entire model parameter space.  
+%The solution to this problem was to add a uniform distribution to the proposal PDF, in the form of one ``fictitious'' point per brick.
+The extra uniform distribution from the fictitious points can represent a significant cost, since inter-model jumps proposed from the uniform distribution are unlikely to be accepted.
+However, the bricks in this
 algorithm do not adapt to the local structure of the PDF.  One must
 either use small bricks to capture the local structure of the PDF,
-placing many bricks in regions without MCMC samples, or use large
-bricks, missing the local strucutre of the PDF in exchange for fewer
-empty bricks.
+placing many bricks in regions without MCMC samples (which can increase memory management and access costs, even for sparse matrices), or use large
+bricks, missing the local structure of the PDF in exchange for fewer
+empty bricks. \will{Are you happy with this bit of editing?}
 
@@
 Also, the computational cost for computing the Voronoi regions
-increases rapidly with
-dimension.  % Apparently the exact computational cost is not known in
-            % arbitrary dimension.
+increases rapidly with dimensionality.  
+% Apparently the exact computational cost is not known in arbitrary dimension.
 
-Here we propose to use a decomposition of parameter space into
-neighborhoods based on a datastructure called a kD-tree (see, e.g.\
+Here we propose to use a decomposition of the parameter space into
+neighborhoods based on a data structure called a kD-tree (see, e.g.\
 \cite{Berg2008} or \cite{Gaede1998}).  The decomposition is more
 adaptive than the cells of Ref.~\cite{Littenberg2009}, and more
-efficient in high dimension than the Voronoi decomposition.
+efficient in high-dimensional spaces than the Voronoi decomposition.
 

### 2011-03-17 paper.tex
Section IV finished.

@@
 An alternate technique for producing adaptive neighborhoods would be
-to use the Voronoi regions associated with each MCMC sample.  The
-Voronoi region associated with a sample contains all the parameter
-space points that are closer to that sample than any other sample.
-The Voronoi region decomposition into neighborhoods is, in a sense,
-maximally adaptive, in contrast to the approach of
+to use the Voronoi regions \cite{Voronoi1907} associated with each
+MCMC sample.  The Voronoi region associated with a sample contains all
+the parameter space points that are closer to that sample than any
+other sample.  The Voronoi region decomposition into neighborhoods is,
+in a sense, maximally adaptive, in contrast to the approach of
 Ref.~\cite{Littenberg2009}, which is minimally adaptive.
@@
 median-finding algorithm.  Tree construction thus costs $\order{N \log
-  N}$ in time, and the tree consumes $\order{N \log N}$ space.  As an
+  N}$ in time, and the tree consumes $\order{N}$ space.  As an
 illustration, box boundaries for a kD-tree constructed around a point
@@
 $(\alpha,\beta)$ spaces to occupy only a few percent of the prior
-volume.  The Cauchy distribution is much broader than the Gaussian
-(it has no finite moments), so with uniform model priors, the
-posterior probability for the Gaussian model over the Cauchy model is
-extremely large:
+volume.  The Cauchy distribution is much broader than the Gaussian (it
+has no finite moments), so with uniform model priors, the posterior
+probability for the Gaussian model over the Cauchy model is extremely
+large:
 %
 \be
-\frac{p(\textnormal{Gaussian} | d)}{p(\textnormal{Cauchy}|d)} > 10^9\ .
+\frac{p(\textnormal{Gaussian} | d)}{p(\textnormal{Cauchy}|d)} \sim 10^9.
 \ee
 %
+In order to ensure that the RJMCMC produces samples in the Cauchy
+model at all, we impose a model prior that favors the Cauchy model by
+$5 \times 10^8$ relative to the Gaussian.  The evidence ratio between
+the models for our chosen data set with these priors is
+%
+\be \frac{p(\textnormal{Gaussian} | d)}{p(\textnormal{Cauchy}|d)}
+\equiv r = 1.15,
+\ee
+%
+yielding a theoretical maximum acceptance rate of inter-model jumps of
+$(1+1/r)/2 = 0.93$.
+
+We obtain $10^4$ individual model MCMC samples within each model, and
+use the kD-tree interpolation method described above to propose
+inter-model jumps in an RJMCMC.  The acceptance rate of inter-model
+jumps is approximately 0.8.  To explore how the efficiency of the
+method degrades as the interpolation becomes less accurate, we
+artificially truncated the kD tree with higher and higher numbers of
+points in each box (this can be accomplished during the neighborhood
+search phase by stopping the search for a box when one is found
+containing the desired number of points).  For each truncation, we
+performed an RJMCMC with the resulting interpolated jump proposal.
+The acceptance rate versus truncation number appear in Figure
+\ref{fig:acceptRate}.  The more points in each box of the tree when
+the search is truncated, the lower the acceptance probability; when
+points are drawn from the top level of the tree, the acceptance
+probability asymptotes to the naive draw from the prior ($\sim 5\%$).
 
@@
   \end{center}
-  \caption{\label{fig:acceptRate} The inter-model acceptance rate as a
-    function of boxing number for the RJMCMC described in the text.}
+  \caption{\label{fig:acceptRate} The inter-model acceptance rate
+    versus the number of points per box when the kD-tree neighborhood
+    search is truncated.  As the number of points per box increases,
+    and the interpolation becomes less accurate, the acceptance rate
+    falls, asymptoting to the rate for naive draws from the uniform
+    prior (about $5\%$).}
 \end{figure}
+
+The relative error on the determination of the evidence ratio scales
+with $1/\sqrt{N_{\textnormal{transitions}}}$, where
+$N_{\textnormal{transitions}}$ is the number of inter-model
+transitions in the RJMCMC.  Thus, as the acceptance rate of
+inter-model jumps goes down, the RJMCMC must run longer to achieve a
+desired accuracy in the evidence ratio.  By boosting the acceptance
+rate of inter-model jumps significantly, the interpolation method
+described above can dramatically improve the runtime of an RJMCMC.
 

### 2011-03-17 paper.tex
Fixed tree space
Minor changes to IV.

@@
 
-We obtain $10^4$ individual model MCMC samples within each model, and
-use the kD-tree interpolation method described above to propose
+We obtain $10^4$ single-model MCMC samples by independently running MCMC within each model, and use the kD-tree interpolation method described above to propose
 inter-model jumps in an RJMCMC.  The acceptance rate of inter-model
@@
 search phase by stopping the search for a box when one is found
-containing the desired number of points).  For each truncation, we
+containing the desired number of points).  For each truncation choice, we
 performed an RJMCMC with the resulting interpolated jump proposal.
-The acceptance rate versus truncation number appear in Figure
-\ref{fig:acceptRate}.  The more points in each box of the tree when
+The acceptance rate is plotted against the number of single-model MCMC points per box (kD-tree leaf) in Figure \ref{fig:acceptRate}.  The more points in each leaf of the tree when
 the search is truncated, the lower the acceptance probability; when
@@
     falls, asymptoting to the rate for naive draws from the uniform
-    prior (about $5\%$).}
+    prior (about $5\%$ for this data set).}
 \end{figure}
 
-The relative error on the determination of the evidence ratio scales
+The relative error on the determination of the Bayes factor (evidence ratio) scales
 with $1/\sqrt{N_{\textnormal{transitions}}}$, where
@@
 inter-model jumps goes down, the RJMCMC must run longer to achieve a
-desired accuracy in the evidence ratio.  By boosting the acceptance
+desired accuracy in the Bayes factor.  By boosting the acceptance
 rate of inter-model jumps significantly, the interpolation method

### 2011-03-17 paper.tex
Edits throughout.

@@
 the odds ratio for the models.
-%, the ratio of Bayesian probabilities for the two models given the observation data.
 
@@
 
-%\ilya{Something about how nested sampling distributes points according to the prior, not the posterior, and therefore can miss modes that are disfavored by prior.
-Also, it does not exploit any information about the posterior to make the evidence calculation more efficient.}
-
-Markov Chain Monte Carlo (MCMC) methods attempt to
-stochastically produce parameter samples with density proportional to
-the posterior probability distribution.
-In MCMC techniques, the primary target is an accurate estimate of the posterior distribution.
-(We note that an alternative stochastic method for exploring a model parameter space, nested sampling \cite{Skilling:2004, Skilling:2006,Feroz:2009}, focuses on evidence computation rather than sampling the posterior probability density functions.)  \will{Will, are you happy with nested sampling in a parenthetical expression, without further comments?}
-It is not straightforward to compute the model evidence from the MCMC
-samples.  The most direct way to estimate the evidence for a model
-from MCMC samples is to compute the harmonic-mean estimator, but
-this estimator of the evidence suffers from infinite
-variance\cite{NewtonRaftery:1994,Chib:1995,vanHaasteren:2009}.  Various
-methods have been proposed to stabilize the harmonic-mean estimator \will{Citation}, but all
-require some assumptions about the shape of the posterior distribution
-that may not be satisfied in practice.
-MCMC implementations with parallel tempering \cite{EarlDeem:2005} allow for evidence computation via thermodynamic integration, but these can be computationally costly.
+Markov Chain Monte Carlo (MCMC) methods attempt to stochastically
+produce parameter samples with density proportional to the posterior
+probability distribution.  In MCMC techniques, the primary target is
+an accurate estimate of the posterior distribution.  (We note that an
+alternative stochastic method for exploring a model parameter space,
+nested sampling \cite{Skilling:2004, Skilling:2006,Feroz:2009},
+focuses on evidence computation rather than sampling the posterior
+probability density functions.)  It is not straightforward to compute
+the model evidence from MCMC samples.  The most direct way to estimate
+the evidence for a model from MCMC samples is to compute the
+harmonic-mean estimator, but this estimator of the evidence suffers
+from infinite
+variance\cite{NewtonRaftery:1994,Chib:1995,vanHaasteren:2009}.  MCMC
+implementations with parallel tempering \cite{EarlDeem:2005} allow for
+evidence computation via thermodynamic integration, but these can be
+computationally costly.
 
 Reference \cite{Weinberg2009} gives a method for directly computing
-the evidence integral from existing MCMC samples by using a kD-tree data structure to decompose a
-parameter space into boxes containing the MCMC sample points.  The
-integral is approximated as a sum over box volumes.  This method is
-promising, but it is not clear in general what statistical and
-systematic errors it introduces and how these are affected by the
-shape of the posterior distribution from which the MCMC samples.
+the evidence integral from existing MCMC samples by using a kD-tree
+data structure to decompose a parameter space into boxes containing
+the MCMC sample points.  The integral is approximated as a sum over
+box volumes.  This method is promising, but it is not clear in general
+what statistical and systematic errors it introduces and how these are
+affected by the shape of the posterior distribution from which the
+MCMC samples.
 
@@
 chain of samples must be Markovian, only the current sample is
-available to the algorithm.  Each time an RJMCMC transitions between
-models, the information about the choices of parameter values in the
-previous model is lost; subsequent jumps into that model must ``start
-fresh,'' and are correspondingly unlikely to be accepted, delaying
-convergence of the RJMCMC sample chain.
+available to the algorithm as it is choosing the next sample.  Each
+time an RJMCMC transitions between models, the information about the
+choices of parameter values in the previous model is lost; subsequent
+jumps into that model must ``start fresh,'' and are correspondingly
+unlikely to be accepted, delaying convergence of the RJMCMC sample
+chain.
 
@@
 
-Reference \cite{Littenberg2009} proposed an alternate method for producing
-inter-model jumps in an RJMCMC that relies on a box decomposition of
-parameter space, but uses fixed-sized boxes.  The box size is
-fixed, so the method cannot adapt to the local structure of the
-posterior, and becomes asymptotically inefficient for high-dimensional parameter spaces or highly peaked posteriors.  Meanwhile, the approximation to the
-posterior distribution produced by the kD-tree is a constant-in-box
-interpolation of the posterior, similar in spirit to the phase-space
-density interpolants produced from N-body positions and momenta in
-Reference \cite{Ascasibar2005}.  The kD-tree interpolation is effective
-in parameter spaces of arbitrary dimensionality, and is quite space-efficient,
-requiring $\order{N \log N}$ storage space and $\order{\log N}$ time
-to produce each proposed jump, where $N$ is the number of single-model
-MCMC samples used to construct the interpolation.
+Reference \cite{Littenberg2009} proposed an alternate method for
+producing inter-model jumps in an RJMCMC that relies on a box
+decomposition of parameter space, but uses fixed-sized boxes.  The box
+size is fixed, so the method cannot adapt to the local structure of
+the posterior, and becomes asymptotically inefficient for
+high-dimensional parameter spaces or highly peaked posteriors.
+Meanwhile, the approximation to the posterior distribution produced by
+the kD-tree is a constant-in-box interpolation of the posterior,
+similar in spirit to the phase-space density interpolants produced
+from N-body positions and momenta in Reference \cite{Ascasibar2005}.
+The kD-tree interpolation is effective in parameter spaces of
+arbitrary dimensionality, and is quite space-efficient, requiring
+$\order{N}$ storage space and $\order{\log N}$ time to produce each
+proposed jump, where $N$ is the number of single-model MCMC samples
+used to construct the interpolation.
 
@@
 a technique for Bayesian model comparison.  Below, we give a very
-brief introduction to Bayesian analysis; describe a standard MCMC; and
+brief introduction to Bayesian analysis, describe a standard MCMC, and
 introduce RJMCMC.
@@
 techniques focused directly on evidence computation (e.g., nested
-sampling \cite{Skilling:2004,Skilling:2006} and its variant MultiNest \cite{Feroz:2009}).  Although nested sampling can be used to compute the posterior PDFs within each model along
-with the evidences for the various models, perhaps the most common
-technique for computing posterior PDFs in the context of a model is
-the Markov chain Monte Carlo, which we now describe.
+sampling \cite{Skilling:2004,Skilling:2006} and its variant MultiNest
+\cite{Feroz:2009}).  Although nested sampling can be used to compute
+the posterior PDFs within each model along with the evidences for the
+various models, the most common technique for computing posterior PDFs
+in the context of a model is the Markov chain Monte Carlo, which we
+now describe.
 
@@
 Sec.~\ref{sec:kDTree}), are described in \cite{Weinberg2009}.  These
-techniques are promising, but in some cases suffer \cite{Farr2010} \will{Is this really the reference you meant?}
-from large variance and bias as well.  An alternative approach to
-model selection among a set of models is based on performing an MCMC
-in a ``super-model'' that encompasses all of the models under
+techniques are promising, but in some cases suffer \cite{Farr2010}
+from large variance and bias.  An alternative approach to model
+selection among a set of models is based on performing an MCMC in a
+``super-model'' that encompasses all of the models under
 consideration; this is known the the Reversible Jump Markov chain
@@
 proposal covers the entire model parameter space.  
-%The solution to this problem was to add a uniform distribution to the proposal PDF, in the form of one ``fictitious'' point per brick.
-The extra uniform distribution from the fictitious points can represent a significant cost, since inter-model jumps proposed from the uniform distribution are unlikely to be accepted.
-However, the bricks in this
-algorithm do not adapt to the local structure of the PDF.  One must
-either use small bricks to capture the local structure of the PDF,
-placing many bricks in regions without MCMC samples (which can increase memory management and access costs, even for sparse matrices), or use large
-bricks, missing the local structure of the PDF in exchange for fewer
-empty bricks. \will{Are you happy with this bit of editing?}
+However, the bricks in this algorithm do not adapt to the local
+structure of the PDF.  One must either use small bricks to capture the
+local structure of the PDF, placing many bricks in regions without
+MCMC samples (which can increase memory management and access costs),
+or use large bricks, missing the local structure of the PDF in
+exchange for fewer empty bricks.
 
@@
 inter-model jumps goes down, the RJMCMC must run longer to achieve a
-desired accuracy in the Bayes factor.  By boosting the acceptance
-rate of inter-model jumps significantly, the interpolation method
-described above can dramatically improve the runtime of an RJMCMC.
+desired accuracy in the evidence ratio.  By boosting the acceptance
+rate of inter-model jumps, the interpolation method described above
+can dramatically improve the runtime of an RJMCMC.
 
@@
 
-As discussed in Section \ref{sec:mcmc}, finding an efficient jump proposal distribution can be a challenge even for a single-model MCMC.
-This is particularly difficult when the parameter space is multi-modal, in which case the Markov chain can get stuck on individual high-likelihood islands in the parameter space.
-Traversals of low-likelihood oceans by a series of short jumps are extremely unlikely, and random long jumps are almost always rejected, making sampling very inefficient.
-Ideally, an analytical understanding of the model that would give a glimpse of the structure of the archipelago of islands would allow for the creation of jump proposal distributions that combine short jumps designed to explore individual islands with directed long jumps that are likely to end up on neighboring islands.
-However, it is not always possible to gain such understanding short of performing the MCMC itself.
-One solution lies in running multiple chains that explore the parameter space, perhaps getting stuck on local maxima, and then combining the points and interpolating with a kD tree to create a jump proposal distribution.
-Although the chains in the first iteration will not have converged, and the relative sizes of the modes will not be accurately known, this will ensure that all of the modes (islands) are at least included in the jump proposal distribution for a second stage, greatly improving its efficiency.
-Other modifications to this technique, which we are currently investigating, include multiple iterative stages, with the interpolated samples from each previous stage used as a jump proposal distribution for the next stage, until the posterior PDF is read off from the final stage; and the use of higher temperatures in the early-stage chains to improve sampling.
+As discussed in Section \ref{sec:mcmc}, finding an efficient jump
+proposal distribution can be a challenge even for a single-model MCMC.
+This is particularly difficult when the parameter space is
+multi-modal, in which case the Markov chain can get stuck on
+individual high-posterior islands in the parameter space.  Traversals
+of low-posterior oceans by a series of short jumps are extremely
+unlikely, and random long jumps are almost always rejected, making
+sampling very inefficient.  Essentially, such isolated islands in
+parameter space behave as separate models, and weighting them by the
+number of samples in each peak then corresponds to model selection.
+Ideally, an analytical understanding of the model would allow for the
+creation of jump proposal distributions that combine short jumps
+designed to explore individual islands with directed long jumps that
+are likely to end up on neighboring islands.  However, it is not
+always possible to gain such understanding short of performing the
+MCMC itself.  One solution lies in running multiple chains that
+explore the parameter space, perhaps getting stuck on local maxima,
+and then combining the points and interpolating with a kD tree to
+create a jump proposal distribution.  Although the chains in the first
+iteration will not have transitioned between the peaks, and the
+samples will therefore not weight the separate peaks appropriately,
+this will ensure that all of the modes (islands) are included in the
+jump proposal distribution for a second stage.  This second stage can
+employ an interpolated jump proposal exactly as in the model selection
+discussed above.  Other modifications to this technique, which we are
+currently investigating, include multiple iterative stages, with the
+interpolated samples from each previous stage used as a jump proposal
+distribution for the next stage, until the posterior PDF is read off
+from the final stage; and the use of higher temperatures in the
+early-stage chains to improve sampling.
 
-The possibility of MCMC chains becoming even temporarily stuck on
-local maxima also affects convergence.  In the absence of a sufficient
-number of jump proposals between the local maxima, the MCMC may not
-accurately represent their relative sizes.  For example, this can
-happen with antipodal sky location degeneracies for networks of three
-ground-based gravitational-wave interferometers or for low-frequency
-gravitational wave signals in the proposed space-borne instrument LISA
-\cite{LISA}.  Although PDFs produced by MCMC chains will generally
-show both locations on the sky, the chain length may not be
-sufficiently long in practice to include enough jumps between the two
-high-likelihood antipodal locations.  In general, the relative weight
-of two separated peaks is determined with a fractional error that
-scales as $1/\sqrt{N_{\textnormal{transitions}}}$, where
+An example of a multi-modal parameter space where chains can get stuck
+on high-posterior islands comes from antipodal sky location
+degeneracies for networks of three ground-based gravitational-wave
+interferometers or for low-frequency gravitational wave signals in the
+proposed space-borne instrument LISA \cite{LISA}.  Although PDFs
+produced by MCMC chains will generally show both locations on the sky,
+the chain length may not be sufficiently long in practice to include
+enough jumps between the two high-likelihood antipodal locations.  In
+general, the relative weight of two separated peaks is determined with
+a fractional error that scales as
+$1/\sqrt{N_{\textnormal{transitions}}}$, where
 $N_{\textnormal{transitions}}$ is the number of transitions between
@@
 convergence of such PDFs can be enhanced using the kD tree
-interpolated proposal.  We can test the convergence of an MCMC with
-local jumps by running a followup MCMC with a jump proposal
-distribution from an interpolation of the first MCMC's PDF.  This
-follow-up MCMC would make large global jump proposals rather than
-small local ones, meaning that it could rapidly accumulate sufficient
-points to determine whether the original run spent correct fractional
-amounts of time in the various modes.
+interpolated proposal.
 

### 2011-03-17 paper.tex
Added conclusion.

--- 2011-03-17 paper.tex
@@
 
-\section{Conclusion/Discussion}
+\section{Conclusion}
 \label{sec:conclusion}
+
+The need to compare evidences for multiple models arises in a large variety of physical and astronomical contexts.
+In this paper, we described a new technique that allows for efficient evidence computations via a Reversible-Jump Markov chain Monte Carlo.
+This technique solves the usual problem of finding good inter-model jump proposals in an RJMCMC by using a kD-tree to quickly and accurately interpolate an approximate posterior PDF from a single-model MCMC run, and then using this interpolated PDF to propose efficient inter-model jumps.
+
+We demonstrated the efficiency of this technique on a toy model-comparison problem described in Section \ref{sec:efficiency}.
+We also successfully applied this technique to the problem of selecting the best model for the observed distribution of black-hole X-ray binaries, as described in Section \ref{sec:examples} and Ref.~\cite{Farr2010}.
+In addition to model comparison, the PDF interpolation described here can be useful in multi-step MCMCs to improve the sampling efficiency by selecting better jump proposal distributions or to test MCMC convergence.
+
+We have made our implementation of the technique described in this paper publicly available online at \will{URL}, and welcome readers to take advantage of this toolkit.
+
+\section*{Acknowledgments}
+
+We are grateful to Neil Cornish for interesting discussions.
+IM acknowledges support from the NSF Astronomy and Astrophysics Postdoctoral Fellowship, award AST-0901985.
+WF acknowledges support from \will{Add your acknowledgements}
 

### 2011-03-17 paper.tex
Change to V.

@@
 the peaks.  By increasing the number of proposed transitions, the
-convergence of such PDFs can be enhanced using the kD tree
-interpolated proposal.
+convergence of such PDFs can be enhanced using the kD-tree
+interpolated proposal.  As described above, one could test the convergence of the PDF 
+by running a followup MCMC with a jump proposal
+distribution from an interpolation of the first original PDF.  This
+follow-up MCMC would attempt large global jump proposals rather than
+small local ones, meaning that it could rapidly accumulate sufficient
+points to determine whether the original run spent the correct fractional
+amounts of time in the various modes.
 

### 2011-03-17 paper.tex
GitHub address and acknowledgements for WMF.

@@
 
-We have made our implementation of the technique described in this paper publicly available online at \will{URL}, and welcome readers to take advantage of this toolkit.
+We have made our implementation of the technique described in this
+paper publicly available online at
+\url{http://github.com/farr/mcmc-ocaml}, and welcome readers to take
+advantage of this toolkit.
 
@@
 
-We are grateful to Neil Cornish for interesting discussions.
-IM acknowledges support from the NSF Astronomy and Astrophysics Postdoctoral Fellowship, award AST-0901985.
-WF acknowledges support from \will{Add your acknowledgements}
+We are grateful to Neil Cornish for interesting discussions.  IM
+acknowledges support from the NSF Astronomy and Astrophysics
+Postdoctoral Fellowship, award AST-0901985.  WF acknowledges support
+from NSF grants CAREER AST-0449558 and AST0908930.
 

### 2011-03-17 paper.tex
Corrected grants for WMF.
Fixed WMF affiliation.

@@
 
-\affiliation{Northwestern University Center for Interdisciplinary
-  Research and Education in Astrophysics}
+\affiliation{Center for Interdisciplinary Exploration and Research in Astrophysics (CIERA) \\ Department of Physics and Astronomy, Northwestern University}
 
@@
 Postdoctoral Fellowship, award AST-0901985.  WF acknowledges support
-from NSF grants CAREER AST-0449558 and AST0908930.
+from NSF grant AST0908930 and NASA grant NNX09AJ56G.
 

### 2011-04-01 paper.tex
Changes suggested by Neil.
Made sure to note that Littenberg and Cornish suggested a similar
method first.

Footnoted that our kD tree is unusual in that it places box boundaries
between the points, while a standard kD tree places boundaries on the
points.

@@
 unlikely to be accepted, delaying convergence of the RJMCMC sample
-chain.
+chain.  One of the first papers to address this issue was Reference
+\cite{Littenberg2009}, where the authors propose a method for
+producing inter-model jumps in an RJMCMC that relies on interpolating
+single-model posterior distributions using a box decomposition of
+parameter space.
 
-Here we introduce a technique for improving the acceptance ratio of
-inter-model jumps in an RJMCMC, leading to dramatically improved
-convergence of RJMCMC sample chains.  The technique uses a kD-tree
-data structure to construct an approximation to each model's posterior
-parameter distribution.  We draw jump proposals into the model from this approximation to its posterior.  Because jumps are proposed
-preferentially to locations favored by the single-model posterior, the
-RJMCMC compares ``good'' locations in parameter space across all the
-models, and convergence is generally rapid.
-We have successfully applied this RJMCMC technique to a 10-way model selection among alternative mass distribution models for black-hole X-ray binaries \cite{Farr2010}.
+Here we introduce an alternative technique for improving the
+acceptance ratio of inter-model jumps in an RJMCMC, leading to
+dramatically improved convergence of RJMCMC sample chains.  The
+technique uses a kD-tree data structure to construct an approximation
+to each model's posterior parameter distribution.  We draw jump
+proposals into the model from this approximation to its posterior.
+Because jumps are proposed preferentially to locations favored by the
+single-model posterior, the RJMCMC compares ``good'' locations in
+parameter space across all the models, and convergence is generally
+rapid.  We have successfully applied this RJMCMC technique to a 10-way
+model selection among alternative mass distribution models for
+black-hole X-ray binaries \cite{Farr2010}.
 
-Reference \cite{Littenberg2009} proposed an alternate method for
-producing inter-model jumps in an RJMCMC that relies on a box
-decomposition of parameter space, but uses fixed-sized boxes.  The box
-size is fixed, so the method cannot adapt to the local structure of
-the posterior, and becomes asymptotically inefficient for
-high-dimensional parameter spaces or highly peaked posteriors.
-Meanwhile, the approximation to the posterior distribution produced by
-the kD-tree is a constant-in-box interpolation of the posterior,
-similar in spirit to the phase-space density interpolants produced
-from N-body positions and momenta in Reference \cite{Ascasibar2005}.
-The kD-tree interpolation is effective in parameter spaces of
-arbitrary dimensionality, and is quite space-efficient, requiring
-$\order{N}$ storage space and $\order{\log N}$ time to produce each
-proposed jump, where $N$ is the number of single-model MCMC samples
-used to construct the interpolation.
+The method of Reference \cite{Littenberg2009} for producing
+inter-model jumps in an RJMCMC relies on a box decomposition of
+parameter space, using fixed-sized boxes.  The method cannot adapt to
+the local structure of the posterior, and becomes asymptotically
+inefficient for high-dimensional parameter spaces or highly peaked
+posteriors.  Meanwhile, the approximation to the posterior
+distribution produced by the kD-tree is a constant-in-box
+interpolation of the posterior, similar in spirit to the phase-space
+density interpolants produced from N-body positions and momenta in
+Reference \cite{Ascasibar2005}.  The kD-tree interpolation is
+effective in parameter spaces of arbitrary dimensionality, and is
+quite space-efficient, requiring $\order{N}$ storage space and
+$\order{\log N}$ time to produce each proposed jump, where $N$ is the
+number of single-model MCMC samples used to construct the
+interpolation.
 
@@
 points into a kD-tree, begin by placing them in a rectangular box that
-contains all of parameter space.  Then proceed recursively:
+contains all of parameter space.  Then proceed recursively%
+\footnote{The kD-tree datastructure defined here places box boundaries
+  between the points.  An alternate definition common in the
+  literature places box boundaries on the median point, but such a
+  definition is inconvenient for our purposes.}: %
 \begin{enumerate}

### 2011-04-05 paper.tex
Wording change on Littenberg and Cornish intro reference.

@@
-\IfFileExists{revtex4-1.cls}{\documentclass[prd, twocolumn, lengthcheck,
+\IfFileExists{revtex4-1.cls}{\documentclass[prd, reprint,
 superscriptaddress, showpacs, letterpaper, nofootinbib]{revtex4-1}}{ 
-\IfFileExists{revtex4.cls}{\documentclass[prd, preprint,
+\IfFileExists{revtex4.cls}{\documentclass[prd,
 superscriptaddress, showpacs, letterpaper, nofootinbib]{revtex4}}{}
@@
 
+\pacs{02.70.Tt, 95.75.Pq, 07.05.Kf}
+
 \maketitle
@@
 unlikely to be accepted, delaying convergence of the RJMCMC sample
-chain.  One of the first papers to address this issue was Reference
-\cite{Littenberg2009}, where the authors propose a method for
-producing inter-model jumps in an RJMCMC that relies on interpolating
-single-model posterior distributions using a box decomposition of
-parameter space.
+chain.  Reference \cite{Littenberg2009} addressed this issue by
+proposing a new method for producing inter-model jumps in an RJMCMC
+that relies on interpolating single-model posterior distributions
+using a box decomposition of parameter space.
 

### 2011-04-05 paper.tex
As submitted to PRD.
* kdtree converted to EPS format.
* Reference abbreviations standardized.

@@
-\IfFileExists{revtex4-1.cls}{\documentclass[prd, reprint,
+\IfFileExists{revtex4-1.cls}{\documentclass[aps, prd, reprint,
 superscriptaddress, showpacs, letterpaper, nofootinbib]{revtex4-1}}{ 
@@
 
-\def\aj{Astronomical Journal}                 % Astronomical Journal
-\def\apj{Astrophysical Journal}                % Astrophysical Journal
-\def\apjl{Astrophysical Journal}             % Astrophysical Journal, Letters
+\def\aj{Astron. J.}                 % Astronomical Journal
+\def\apj{Astrophys. J.}                % Astrophysical Journal
+\def\apjl{Astrophys. J. Lett.}             % Astrophysical Journal, Letters
 \def\pasj{PASJ}
-\def\apjs{ApJS}              % Astrophysical Journal, Supplement
-\def\mnras{MNRAS}            % Monthly Notices of the RAS
-\def\prd{Phys.~Rev.~D}       % Physical Review D
-\def\prl{Phys.~Rev.~Lett}    % Physical Review Letters
-\def\cqg{Class.~Quant.~Grav.~}%Classical and Quantum Gravity
-\def\araa{ARA\&A}             % Annual Review of Astron and Astrophys
+\def\apjs{Astrophys. J., Suppl. Ser.}              % Astrophysical Journal, Supplement
+\def\mnras{Mon. Not. R. Astron. Soc.}            % Monthly Notices of the RAS
+\def\prd{Phys. Rev. D}       % Physical Review D
+\def\prl{Phys. Rev. Lett.}    % Physical Review Letters
+\def\cqg{Class. Quant. Grav.}%Classical and Quantum Gravity
+\def\araa{Annu. Rev. Astron. Astrophys.}             % Annual Review of Astron and Astrophys
 \def\nat{Nature}              % Nature
-\def\aap{A\&A}                % Astronomy and Astrophysics
+\def\aap{Astron. Astrophys.}                % Astronomy and
+                                % Astrophysics
+\def\jasa{J. Am. Stat. Assoc.}
+\def\pccp{Phys. Chem. Chem. Phys.}
+\def\jrssb{J. R. Stat. Soc. B}
+\def\aipcs{AIP Conf. Ser.}
 
@@
   \begin{center}
-    \includegraphics[width=0.8\columnwidth]{kdtree.png}
+    \includegraphics[width=0.8\columnwidth]{kdtree}
   \end{center}

### 2011-04-07 paper.tex
Re-tooled for CQG.

@@
-\IfFileExists{revtex4-1.cls}{\documentclass[aps, prd, reprint,
-superscriptaddress, showpacs, letterpaper, nofootinbib]{revtex4-1}}{ 
-\IfFileExists{revtex4.cls}{\documentclass[prd,
-superscriptaddress, showpacs, letterpaper, nofootinbib]{revtex4}}{}
-}
+\documentclass{iopart}
 
-\usepackage{amsmath}
 \usepackage{hyperref}
@@
 \usepackage{graphicx}
+\usepackage{harvard}
 
@@
 \def\aipcs{AIP Conf. Ser.}
+\def\jcp{J. Chem. Phys.}
 
-\IfFileExists{apsrev4-1.bst}{\bibliographystyle{apsrev4-1}}
-\IfFileExists{apsrev4.bst}{\bibliographystyle{apsrev4}}
+\bibliographystyle{jphysicsB}
 
 \begin{document}
-\title{An Efficient Interpolation Technique for Jump Proposals in
-  Reversible-Jump Markov Chain Monte Carlo Calculations}
+\title[Efficient Jump Proposals in RJMCMC]{An Efficient Interpolation
+  Technique for Jump Proposals in Reversible-Jump Markov Chain Monte
+  Carlo Calculations}
 
@@
 
-\author{Will M. Farr}
-\email{w-farr@northwestern.edu}
+\author{W M Farr$^1$ and I Mandel$^2$}
 
-\affiliation{Center for Interdisciplinary Exploration and Research in Astrophysics (CIERA) \\ Department of Physics and Astronomy, Northwestern University}
+\address{$^1$ Center for Interdisciplinary Exploration and Research in Astrophysics (CIERA) \\ Department of Physics and Astronomy, Northwestern University}
 
-\author{Ilya Mandel}
-\email{ilyamandel@chgk.info}
+\address{$^2$ MIT Kavli Institute; and University of Birmingham, UK}
 
-\affiliation{MIT Kavli Institute; and University of Birmingham, UK}
+\eads{\mailto{w-farr@northwestern.edu}, \mailto{ilyamandel@chgk.info}}
 
@@
   data set is an important challenge in many areas of physics and
-  astronomy.
-  %ranging from gravitational-wave data analysis to
-  %measurements of black-hole masses in X-ray binaries.
-  Reversible-jump Markov chain Monte Carlo (RJMCMC) is an extremely
-  powerful technique for performing Bayesian model selection, but it
-  suffers from a fundamental difficulty: it requires jumps between
-  model parameter spaces, but cannot retain a memory of the favored
-  locations in more than one parameter space at a time.  Thus, a naive
-  jump between parameter spaces is unlikely to be accepted in the MCMC
-  algorithm and convergence is correspondingly slow.  Here we
-  demonstrate an interpolation technique that uses samples from
-  single-model MCMCs to propose inter-model jumps from an
+  astronomy.  Reversible-jump Markov chain Monte Carlo (RJMCMC) is an
+  extremely powerful technique for performing Bayesian model
+  selection, but it suffers from a fundamental difficulty: it requires
+  jumps between model parameter spaces, but cannot retain a memory of
+  the favored locations in more than one parameter space at a time.
+  Thus, a naive jump between parameter spaces is unlikely to be
+  accepted in the MCMC algorithm and convergence is correspondingly
+  slow.  Here we demonstrate an interpolation technique that uses
+  samples from single-model MCMCs to propose inter-model jumps from an
   approximation to the single-model posterior of the target parameter
@@
 \pacs{02.70.Tt, 95.75.Pq, 07.05.Kf}
+
+\submitto{\CQG}
 
@@
 alternative stochastic method for exploring a model parameter space,
-nested sampling \cite{Skilling:2004, Skilling:2006,Feroz:2009},
-focuses on evidence computation rather than sampling the posterior
-probability density functions.)  It is not straightforward to compute
-the model evidence from MCMC samples.  The most direct way to estimate
-the evidence for a model from MCMC samples is to compute the
-harmonic-mean estimator, but this estimator of the evidence suffers
-from infinite
-variance\cite{NewtonRaftery:1994,Chib:1995,vanHaasteren:2009}.  MCMC
+nested sampling \cite{Skilling:2004,Skilling:2006,Feroz:2009}, focuses
+on evidence computation rather than sampling the posterior probability
+density functions.)  It is not straightforward to compute the model
+evidence from MCMC samples.  The most direct way to estimate the
+evidence for a model from MCMC samples is to compute the harmonic-mean
+estimator, but this estimator of the evidence suffers from infinite
+variance \cite{NewtonRaftery:1994,Chib:1995,vanHaasteren:2009}.  MCMC
 implementations with parallel tempering \cite{EarlDeem:2005} allow for
@@
 
-Reference \cite{Weinberg2009} gives a method for directly computing
+\cite{Weinberg2009} gives a method for directly computing
 the evidence integral from existing MCMC samples by using a kD-tree
@@
 circumstance, the Reversible Jump MCMC technique first introduced in
-Reference \cite{Green1995} is one of the most reliable and accurate
-ways to compare the models.  Reversible Jump MCMC (RJMCMC), described
-more fully in Section \ref{sec:reversible-jump}, performs a standard
-MCMC in a superspace that is a direct sum of all the model parameter
+\cite{Green1995} is one of the most reliable and accurate ways to
+compare the models.  Reversible Jump MCMC (RJMCMC), described more
+fully in Section \ref{sec:reversible-jump}, performs a standard MCMC
+in a superspace that is a direct sum of all the model parameter
 spaces.  Such an MCMC involves both intra- and inter-model jumps; the
@@
 unlikely to be accepted, delaying convergence of the RJMCMC sample
-chain.  Reference \cite{Littenberg2009} addressed this issue by
-proposing a new method for producing inter-model jumps in an RJMCMC
-that relies on interpolating single-model posterior distributions
-using a box decomposition of parameter space.
+chain.  \cite{Littenberg2009} addressed this issue by proposing a new
+method for producing inter-model jumps in an RJMCMC that relies on
+interpolating single-model posterior distributions using a box
+decomposition of parameter space.
 
@@
 
-The method of Reference \cite{Littenberg2009} for producing
-inter-model jumps in an RJMCMC relies on a box decomposition of
-parameter space, using fixed-sized boxes.  The method cannot adapt to
-the local structure of the posterior, and becomes asymptotically
-inefficient for high-dimensional parameter spaces or highly peaked
-posteriors.  Meanwhile, the approximation to the posterior
-distribution produced by the kD-tree is a constant-in-box
-interpolation of the posterior, similar in spirit to the phase-space
-density interpolants produced from N-body positions and momenta in
-Reference \cite{Ascasibar2005}.  The kD-tree interpolation is
-effective in parameter spaces of arbitrary dimensionality, and is
-quite space-efficient, requiring $\order{N}$ storage space and
-$\order{\log N}$ time to produce each proposed jump, where $N$ is the
-number of single-model MCMC samples used to construct the
-interpolation.
+The method of \cite{Littenberg2009} for producing inter-model jumps in
+an RJMCMC relies on a box decomposition of parameter space, using
+fixed-sized boxes.  The method cannot adapt to the local structure of
+the posterior, and becomes asymptotically inefficient for
+high-dimensional parameter spaces or highly peaked posteriors.
+Meanwhile, the approximation to the posterior distribution produced by
+the kD-tree is a constant-in-box interpolation of the posterior,
+similar in spirit to the phase-space density interpolants produced
+from N-body positions and momenta in \cite{Ascasibar2005}.  The
+kD-tree interpolation is effective in parameter spaces of arbitrary
+dimensionality, and is quite space-efficient, requiring $\order{N}$
+storage space and $\order{\log N}$ time to produce each proposed jump,
+where $N$ is the number of single-model MCMC samples used to construct
+the interpolation.
 
@@
 particular model, $M_i$, given the data, $p(M_i|d)$.  Using Bayes'
-rule, we see that this involves the evidence (Eq.~\eqref{evidence}):
+rule, we see that this involves the evidence \eref{evidence}:
 \be
@@
 attempts to directly compute the evidence by performing the
-integration in Eq.~\eqref{evidence} are generally very difficult in a
+integration in \eref{evidence} are generally very difficult in a
 multi-dimensional, multi-modal parameter space when the likelihood has
@@
 One way to produce such a sequence of samples is via the
-Metropolis-Hastings algorithm, first proposed in Reference
-\cite{Metropolis:1953}, and later generalized by Hastings \cite{Hastings:1970}:
+Metropolis-Hastings algorithm, first proposed in
+\cite{Metropolis:1953}, and later generalized in \cite{Hastings:1970}:
 \begin{enumerate}
@@
 Gravitational-wave astronomy provides the setting for another ongoing
-study.  Interesting triggers from the LIGO \cite{InitLIGO} and Virgo \cite{Virgo}
-gravitational-wave detector pipeline that searches for compact binary
-coalescences are followed up with Bayesian parameter estimation tools (see, e.g., \cite{Rover:2006ni, vanderSluys:2008b, VeitchVecchio:2009}).
-In general, several possible models for the data are considered,
+study.  Interesting triggers from the LIGO \cite{InitLIGO} and Virgo
+\cite{Virgo} gravitational-wave detector pipeline that searches for
+compact binary coalescences are followed up with Bayesian parameter
+estimation tools (see, e.g.,
+\cite{Rover:2006ni,vanderSluys:2008b,VeitchVecchio:2009}).  In
+general, several possible models for the data are considered,
 including a pure noise model, noise superimposed on a non-spinning
@@
 components.  The models can include a large number of parameters (up
-to 15 for the model with two spinning components \cite{Raymond:2010}), and RJMCMC with
-interpolation via kD trees is again successful at computing the model
-evidences.
+to 15 for the model with two spinning components \cite{Raymond:2010}),
+and RJMCMC with interpolation via kD trees is again successful at
+computing the model evidences.
 
@@
 
+\section*{References}
+
 \bibliography{paper}

### 2013-11-19 paper.tex
Re-named Figures to FigureNN_name
Dan's edits (11/8/2012 email: Paper (First Draft))

@@
 \newcommand{\vtheta}{\vec{\theta}}
+\newcommand{\Nbox}{N_\mathrm{box}}
+\newcommand{\Vbox}{V_\mathrm{box}}
 
@@
 
-\author{W M Farr$^1$ and I Mandel$^2$}
+\author{W M Farr$^{1,3}$, I Mandel$^{2,3}$ and D Stevens$^{1,4}$}
 
-\address{$^1$ Center for Interdisciplinary Exploration and Research in Astrophysics (CIERA) \\ Department of Physics and Astronomy, Northwestern University}
+\address{$^1$ Center for Interdisciplinary Exploration and Research in
+  Astrophysics (CIERA), Department of Physics and Astronomy,
+  Northwestern University, Evanston IL USA}
 
-\address{$^2$ MIT Kavli Institute; and University of Birmingham, UK}
+\address{$^2$ MIT Kavli Institute, Cambridge MA USA}
 
-\eads{\mailto{w-farr@northwestern.edu}, \mailto{ilyamandel@chgk.info}}
+\address{$^3$ School of Physics and Astronomy, University of
+  Birmingham, Birmingham UK}
+
+\address{$^4$ Department of Astronomy, The Ohio State
+  University, Columbus OH USA}
+
+\eads{\mailto{w.farr@bham.ac.uk}, \mailto{ilyamandel@chgk.info} \mailto{dan.stevens@u.northwestern.edu}}
 
@@
   space.  The interpolation technique, based on a kD-tree data
-  structure, is adaptive and efficient in arbitrary dimensions.  We
+  structure, is adaptive and efficient in modest dimensionality.  We
   show that our technique leads to dramatically improved convergence
@@
   the literature to improve the convergence of RJMCMCs.  We also
-  discuss the use of the same interpolation technique in two other
-  contexts: as a convergence test for a single-model MCMC and as a way
-  to construct efficient ``global'' proposal distributions for
+  demonstrate the use of the same interpolation technique as a way to
+  construct efficient ``global'' proposal distributions for
   single-model MCMCs without prior knowledge of the structure of the
-  posterior distribution.
+  posterior distribution, and discuss improvements that permit the
+  method to be used in higher-dimensional spaces efficiently.
 \end{abstract}
@@
 the unnormalized posterior probability distribution over the model
-parameter space, representing the probability of obtaining the data set within that model.
-Models with larger evidence are preferred; the ratio of the evidences
-of two models is the Bayes factor between them.  The product of the
-Bayes factor and the ratio of prior probabilities for the two models yields
-the odds ratio for the models.
+parameter space, representing the probability of obtaining the data
+set within that model.  Models with larger evidence are preferred; the
+ratio of the evidences of two models is the Bayes factor between them.
+The product of the Bayes factor and the ratio of prior probabilities
+for the two models yields the odds ratio for the models.
 
@@
 alternative stochastic method for exploring a model parameter space,
-nested sampling \cite{Skilling:2004,Skilling:2006,Feroz:2009}, focuses
-on evidence computation rather than sampling the posterior probability
-density functions.)  It is not straightforward to compute the model
-evidence from MCMC samples.  The most direct way to estimate the
-evidence for a model from MCMC samples is to compute the harmonic-mean
-estimator, but this estimator of the evidence suffers from infinite
-variance \cite{NewtonRaftery:1994,Chib:1995,vanHaasteren:2009}.  MCMC
-implementations with parallel tempering \cite{EarlDeem:2005} allow for
-evidence computation via thermodynamic integration, but these can be
-computationally costly.
+nested sampling \cite{Skilling:2004,Skilling:2006,Feroz:2009},
+focuses on evidence computation rather than sampling the posterior
+probability density functions.)  It is not straightforward to compute
+the model evidence from MCMC samples.  The most direct way to estimate
+the evidence for a model from MCMC samples is to compute the
+harmonic-mean estimator, but this estimator of the evidence suffers
+from infinite variance
+\cite{NewtonRaftery:1994,Chib:1995,vanHaasteren:2009}.  MCMC
+implementations with parallel tempering \cite{EarlDeem:2005} allow
+for evidence computation via thermodynamic integration, but these can
+be computationally costly.
 
@@
 model selection among alternative mass distribution models for
-black-hole X-ray binaries \cite{Farr2010}.
+black-hole X-ray binaries \cite{Farr2010}.  We also provide an example
+using this method as an ``asymptotically Markovian''
+\cite{terBraak2008} jump proposal in the context of a single-model,
+nine-dimensional MCMC in Section \ref{sec:higherdim}.
 
@@
 from N-body positions and momenta in \cite{Ascasibar2005}.  The
-kD-tree interpolation is effective in parameter spaces of arbitrary
+kD-tree interpolation is effective in parameter spaces of modest
 dimensionality, and is quite space-efficient, requiring $\order{N}$
@@
 naive jump proposal in an RJMCMC.  In Section \ref{sec:kDTree} we
-introduce the kD-tree data structure used to decompose the parameter space
-into boxes for interpolation.  In Section \ref{sec:efficiency} we
-demonstrate the efficiency gains that are achieved from use of the
-interpolated jump proposal.  In Section \ref{sec:examples} we give
-examples of some other uses of the interpolated jump proposal, and suggest its utility in the context of a
-single-model MCMC.  Finally, in Section \ref{sec:conclusion}
-we offer a summary and some concluding remarks on the method.
+introduce the kD-tree data structure used to decompose the parameter
+space into boxes for interpolation.  In Section \ref{sec:efficiency}
+we demonstrate the efficiency gains that are achieved from use of the
+interpolated jump proposal.  In Section \ref{sec:higherdim} we give
+examples of some other uses of the interpolated jump proposal that
+suggest its utility in the context of a single-model MCMC.  Finally,
+in Section \ref{sec:conclusion} we offer a summary and some concluding
+remarks on the method.
 
@@
 Metropolis-Hastings algorithm, first proposed in
-\cite{Metropolis:1953}, and later generalized in \cite{Hastings:1970}:
+\cite{Metropolis:1953}, and later generalized in
+\cite{Hastings:1970}:
 \begin{enumerate}
@@
 PDFs are found, creating a Catch-22; one possibility for resolving
-this infinite loop is described in Section \ref{sec:examples}.
+this infinite loop is described in Section \ref{sec:higherdim}.
 
@@
 effective sample size that is too small to accurately represent the
-relative sizes of the modes in the PDF (however, see Section
-\ref{sec:examples} for one intriguing suggestion for remedying this
-issue).
+relative sizes of the modes in the PDF.
 
@@
 Here we propose to use a decomposition of the parameter space into
-neighborhoods based on a data structure called a kD-tree (see, e.g.\
-\cite{Berg2008} or \cite{Gaede1998}).  The decomposition is more
-adaptive than the cells of Ref.~\cite{Littenberg2009}, and more
+neighborhoods based on a data structure called a kD-tree (see,
+e.g.\ \cite{Berg2008} or \cite{Gaede1998}).  The decomposition is more
+adaptive than the boxes of Ref.~\cite{Littenberg2009}, and more
 efficient in high-dimensional spaces than the Voronoi decomposition.
@@
   \begin{center}
-    \includegraphics[width=0.8\columnwidth]{kdtree}
+    \includegraphics[width=0.8\columnwidth]{Figure1_kdtree}
   \end{center}
@@
 
-We obtain $10^4$ single-model MCMC samples by independently running MCMC within each model, and use the kD-tree interpolation method described above to propose
-inter-model jumps in an RJMCMC.  The acceptance rate of inter-model
-jumps is approximately 0.8.  To explore how the efficiency of the
-method degrades as the interpolation becomes less accurate, we
-artificially truncated the kD tree with higher and higher numbers of
-points in each box (this can be accomplished during the neighborhood
-search phase by stopping the search for a box when one is found
-containing the desired number of points).  For each truncation choice, we
-performed an RJMCMC with the resulting interpolated jump proposal.
-The acceptance rate is plotted against the number of single-model MCMC points per box (kD-tree leaf) in Figure \ref{fig:acceptRate}.  The more points in each leaf of the tree when
-the search is truncated, the lower the acceptance probability; when
-points are drawn from the top level of the tree, the acceptance
+We obtain $10^4$ single-model MCMC samples by independently running
+MCMC within each model, and use the kD-tree interpolation method
+described above to propose inter-model jumps in an RJMCMC.  The
+acceptance rate of inter-model jumps is approximately 0.8.  To explore
+how the efficiency of the method degrades as the interpolation becomes
+less accurate, we artificially truncated the kD tree with higher and
+higher numbers of points in each box (this can be accomplished during
+the neighborhood search phase by stopping the search for a box when
+one is found containing the desired number of points).  For each
+truncation choice, we performed an RJMCMC with the resulting
+interpolated jump proposal.  The acceptance rate is plotted against
+the number of single-model MCMC points per box (kD-tree leaf) in
+Figure \ref{fig:acceptRate}.  The more points in each leaf of the tree
+when the search is truncated, the lower the acceptance probability;
+when points are drawn from the top level of the tree, the acceptance
 probability asymptotes to the naive draw from the prior ($\sim 5\%$).
@@
   \begin{center}
-    \includegraphics[width=0.8\columnwidth]{acceptRate}
+    \includegraphics[width=0.8\columnwidth]{Figure2_acceptRate}
   \end{center}
@@
 
-\section{Examples and other uses} 
-\label{sec:examples}
+\section{kD-Interpolated Jump Proposal in Higher-Dimensional
+  Single-Model MCMCs}
+\label{sec:higherdim}
 
-RJMCMC with efficient PDF interpolation via kD trees can be used in a large variety of problems in physics and astronomy, whenever Bayesian analysis is used to perform model selection.
-Below, we describe several scenarios in which we have successfully applied this technique.
-Moreover, PDF interpolation with a kD tree can be extremely useful in other contexts, beyond a reversible-jump MCMC.
-We suggest two examples below: generating efficient jump proposal distributions in a single-model MCMC and convergence tests.
+In the model selection example from Section \ref{sec:efficiency}, the
+models have two-dimensional parameter spaces; in \cite{Farr2010} the
+highest-dimensional model had five dimensions.  As the number of
+dimensions increases the interpolation becomes more difficult for two
+reasons.  First, the number of sample points required for a given
+number of subdivisions in each dimension grows exponentially with
+dimension, so for reasonable numbers of points a higher-dimensional kD
+tree will have fewer subdivisions along each dimension.  Second, as
+the dimensionality increases the volume ``at the corners'' of each kD
+tree box becomes more significant; this problem is even more
+pronounced when the distribution of points does not align with the
+coordinate axes along which the tree subdivides.  In this section, we
+discuss practical solutions to this issue in the context of a
+single-model MCMC, concluding with an example of a successful
+application to gravitational-wave parameter estimation.
 
-We have successfully employed the RJMCMC technique described above
-when evaluating several alternative models for the distribution of the
-masses of black holes in X-ray binaries \cite{Farr2010}.  We
-performed a Bayesian analysis of the mass distribution of stellar-mass
-black holes using the observed masses of 15 low-mass X-ray binary
-systems undergoing Roche lobe overflow and five high-mass, wind-fed
-X-ray binary systems.  We considered ten different mass distribution
-models: Gaussian, double Gaussian, power law, exponential decay, log
-normal, and 1-, 2-, 3-, 4-, and 5-bin histograms.  Each model was
-described by between two and six parameters.  Model selection using
-RJMCMC with kD trees allowed us to determine that the mass
-distribution of the low-mass systems is best fit by a power-law, while
-the distribution of the combined sample is best fit by the exponential
-model.  Based on the model selection, we were able to determine which
-models would provide the best information about astrophysically
-relevant parameters, such as the minimum black hole mass.  We were
-also able to conclude that the low-mass subsample is not consistent
-with being drawn from the distribution of the combined population
-\cite{Farr2010}.
+To address the arbitrary orientation of points, the modified jump
+proposal does not descend to the leaf-nodes of the tree, but instead
+stops descending whenever the number of points in the box falls below
+a threshold, $\Nbox$.  We then find the principal axes of the points
+contained in this box and draw from a modified box which is aligned
+with these axes and whose center is the mean of the points contained
+within the rectangular box. The new point is chosen uniformly from
+within the bounds of the modified box, subject to the restriction that
+it also lie within the original, coordinate-aligned box.  Figure
+\ref{fig:PCC} illustrates this two-box method.
 
-Gravitational-wave astronomy provides the setting for another ongoing
-study.  Interesting triggers from the LIGO \cite{InitLIGO} and Virgo
-\cite{Virgo} gravitational-wave detector pipeline that searches for
-compact binary coalescences are followed up with Bayesian parameter
-estimation tools (see, e.g.,
-\cite{Rover:2006ni,vanderSluys:2008b,VeitchVecchio:2009}).  In
-general, several possible models for the data are considered,
-including a pure noise model, noise superimposed on a non-spinning
-gravitational-wave signal, or noise superimposed on a gravitational
-wave signal that includes significant spin or spins in the binary
-components.  The models can include a large number of parameters (up
-to 15 for the model with two spinning components \cite{Raymond:2010}),
-and RJMCMC with interpolation via kD trees is again successful at
-computing the model evidences.
+\begin{figure}
+  \includegraphics[width=0.8\columnwidth]{PCC}
+  \caption{\label{fig:PCC} A two-dimensional illustration of the two
+    boxes involved in the modified kD proposal.  The larger box
+    aligned with the axes is the normal kD-tree box containing the
+    given points, while the tighter box is the box from which the
+    modified kd-interpolated jump proposal draws a point. }
+\end{figure}
 
-As discussed in Section \ref{sec:mcmc}, finding an efficient jump
-proposal distribution can be a challenge even for a single-model MCMC.
-This is particularly difficult when the parameter space is
-multi-modal, in which case the Markov chain can get stuck on
-individual high-posterior islands in the parameter space.  Traversals
-of low-posterior oceans by a series of short jumps are extremely
-unlikely, and random long jumps are almost always rejected, making
-sampling very inefficient.  Essentially, such isolated islands in
-parameter space behave as separate models, and weighting them by the
-number of samples in each peak then corresponds to model selection.
-Ideally, an analytical understanding of the model would allow for the
-creation of jump proposal distributions that combine short jumps
-designed to explore individual islands with directed long jumps that
-are likely to end up on neighboring islands.  However, it is not
-always possible to gain such understanding short of performing the
-MCMC itself.  One solution lies in running multiple chains that
-explore the parameter space, perhaps getting stuck on local maxima,
-and then combining the points and interpolating with a kD tree to
-create a jump proposal distribution.  Although the chains in the first
-iteration will not have transitioned between the peaks, and the
-samples will therefore not weight the separate peaks appropriately,
-this will ensure that all of the modes (islands) are included in the
-jump proposal distribution for a second stage.  This second stage can
-employ an interpolated jump proposal exactly as in the model selection
-discussed above.  Other modifications to this technique, which we are
-currently investigating, include multiple iterative stages, with the
-interpolated samples from each previous stage used as a jump proposal
-distribution for the next stage, until the posterior PDF is read off
-from the final stage; and the use of higher temperatures in the
-early-stage chains to improve sampling.
+The forward jump probability for this jump proposal is given by
+%
+\bel{modforward} P(\vtheta_i \rightarrow \vtheta_{i+1}) =
+\frac{\Nbox}{N \left(V \cap \Vbox \right)} , \ee
+%
+where N is the total number of points in the kD-tree, $\Nbox$ is the
+number of points in the box containing $\vtheta_{i+1}$, $V$ is the
+volume of the original box containing $\vtheta_{i+1}$ and $\Vbox$ is
+the volume of the principal-component box containing $\vtheta_{i+1}$.
 
-An example of a multi-modal parameter space where chains can get stuck
-on high-posterior islands comes from antipodal sky location
-degeneracies for networks of three ground-based gravitational-wave
-interferometers or for low-frequency gravitational wave signals in the
-proposed space-borne instrument LISA \cite{LISA}.  Although PDFs
-produced by MCMC chains will generally show both locations on the sky,
-the chain length may not be sufficiently long in practice to include
-enough jumps between the two high-likelihood antipodal locations.  In
-general, the relative weight of two separated peaks is determined with
-a fractional error that scales as
-$1/\sqrt{N_{\textnormal{transitions}}}$, where
-$N_{\textnormal{transitions}}$ is the number of transitions between
-the peaks.  By increasing the number of proposed transitions, the
-convergence of such PDFs can be enhanced using the kD-tree
-interpolated proposal.  As described above, one could test the convergence of the PDF 
-by running a followup MCMC with a jump proposal
-distribution from an interpolation of the first original PDF.  This
-follow-up MCMC would attempt large global jump proposals rather than
-small local ones, meaning that it could rapidly accumulate sufficient
-points to determine whether the original run spent the correct fractional
-amounts of time in the various modes.
+The backward jump probability is computed in a similar fashion by
+finding the principal-component box that would contain the current set
+of parameters;.
+%
+\bel{modbackward} P(\vtheta_{i+1} \rightarrow \vtheta_{i}) =
+\frac{\Nbox'}{N\left(V' \cap \Vbox'\right)} , \ee
+%
+where $\Nbox'$ is the number of points in the original box containing
+$\vtheta_i$, $V'$ is the volume of the kD box containing $\vtheta_i$,
+and $\Vbox'$ is the volume of the principal-axes box containing
+$\vtheta_i$.  Note, however, that the current parameters may lie in a
+region of the kD box that does not intersect the principal axes box;
+in this case, the backward jump probability is zero, and the algorithm
+rejects the proposed step.  When used as the only jump proposal for an
+RJMCMC, it is important that the kD tree proposal be capable of
+proposing points in all allowed regions of parameter space; on the
+other hand, when the kD proposal is used as one of several proposals
+in the context of an MCMC, this is no longer a requirement.
+
+It can also be advantageous, particularly in high dimensional spaces,
+to accumulate more points into the kD tree as the MCMC explores more
+of the parameter space.  Because this introduces a dependence on the
+previous states of the MCMC, the jump proposal is no longer
+Markovian.  However, it remains \emph{asymptotically} Markovian, which
+ensures that the equilibrium distribution of the MCMC is correct.  To
+efficiently insert points into the tree as the chain accumulates them,
+the algorithm from Section \ref{sec:kDTree} must be modified:
+\begin{enumerate}
+\item Instead of partitioning the point set at its median, we now
+  partition the bounding box at its geometric center along a
+  particular dimension, which cycles as we descend the tree levels.
+  Note that this allows for empty boxes if all the points cluster to
+  one side of the domain.
+\item When inserting a point, we descend the tree to a leaf node,
+  which now can be either empty or contain one point.  If it is empty,
+  we insert the point at this node.  If it contains one point, we
+  subdivide the appropriate dimension, and place both points into the
+  two sub-boxes.  If both land in the same sub-box, we continue
+  subdividing, until each box in the sub-sub-sub... tree contains one
+  or zero points.
+\end{enumerate}
+Insertion of a point is $\order{\log N}$, since it involves a constant
+amount of work per tree level, and there are $\log N$ tree levels.
+Neighborhood lookup proceeds as before, and is therefore also
+$\order{\log N}$.
+
+We have implemented this modified kD tree proposal as one of many jump
+proposals in the \texttt{LALInferenceMCMC} sampler
+\cite{Raymond2009,Raymond:2010,Raymond2012}.
+\texttt{LALInferenceMCMC} is a MCMC code, based on the LIGO algorithms
+library, designed to sample the posterior on parameters of
+compact-object binaries (masses, orbital orientation, distance, sky
+position, etc) implied by their gravitational wave signals in
+Earth-based gravitational wave detectors like LIGO.  The simplest such
+signal has a nine-dimensional parameter space, and the posterior often
+includes near-degeneracies and multiple modes that make convergence of
+the MCMC chain very slow with traditional proposals.  Figure
+\ref{fig:accratio} shows the acceptance ratio for the kD tree jump
+proposal in such a simulation.  In spite of the very low acceptance
+rate of the proposal, applying the kD tree proposal to one out of
+twenty proposed jumps improved the convergence time of the simulation
+by a factor of two compared to the standard suite of proposals because
+it is particularly efficient at producing ``mode-hopping'' jumps,
+which are difficult to produce with the other proposals in
+\texttt{LALInferenceMCMC}.
+
+\begin{figure}
+  \begin{center}
+    \includegraphics[width=0.8\columnwidth]{linearlog3}
+  \end{center}
+  \caption{\label{fig:accratio} The cumulative parameter acceptance
+    ratio for the modified kD-interpolated jump proposal used in the
+    \texttt{LALInferenceMCMC} code as a function of the number of
+    steps (in hundreds). In spite of the small acceptance ratio, the
+    kD tree proposal, applied to one in twenty proposed jumps,
+    improved the convergence time of the sampler by a factor of two
+    compared to using only the standard suite of proposals.}
+\end{figure}
 
@@
 
-The need to compare evidences for multiple models arises in a large variety of physical and astronomical contexts.
-In this paper, we described a new technique that allows for efficient evidence computations via a Reversible-Jump Markov chain Monte Carlo.
-This technique solves the usual problem of finding good inter-model jump proposals in an RJMCMC by using a kD-tree to quickly and accurately interpolate an approximate posterior PDF from a single-model MCMC run, and then using this interpolated PDF to propose efficient inter-model jumps.
+The need to compare evidences for multiple models arises in a large
+variety of physical and astronomical contexts.  In this paper, we
+described a new technique that allows for efficient evidence
+computations via a Reversible-Jump Markov chain Monte Carlo.  This
+technique solves the usual problem of finding good inter-model jump
+proposals in an RJMCMC by using a kD-tree to quickly and accurately
+interpolate an approximate posterior PDF from a single-model MCMC run,
+and then using this interpolated PDF to propose efficient inter-model
+jumps.
 
-We demonstrated the efficiency of this technique on a toy model-comparison problem described in Section \ref{sec:efficiency}.
-We also successfully applied this technique to the problem of selecting the best model for the observed distribution of black-hole X-ray binaries, as described in Section \ref{sec:examples} and Ref.~\cite{Farr2010}.
-In addition to model comparison, the PDF interpolation described here can be useful in multi-step MCMCs to improve the sampling efficiency by selecting better jump proposal distributions or to test MCMC convergence.
+We demonstrated the efficiency of this technique on a toy
+model-comparison problem described in Section \ref{sec:efficiency}.
+We also successfully applied this technique to the problem of
+selecting the best model for the observed distribution of black-hole
+X-ray binaries, as described in Ref.~\cite{Farr2010}.  In addition to
+model comparison, the PDF interpolation described here can be useful
+in multi-step MCMCs to improve the sampling efficiency by selecting
+better jump proposal distributions or to test MCMC convergence.
 
@@
 paper publicly available online at
-\url{http://github.com/farr/mcmc-ocaml}, and welcome readers to take
-advantage of this toolkit.
+\url{http://github.com/farr/mcmc-ocaml}, and also in the
+\texttt{LALInferenceMCMC} sampler, at
+\url{https://www.lsc-group.phys.uwm.edu/daswg/projects/lalsuite.html}.
+We welcome readers to take advantage of this toolkit.
 
@@
 Postdoctoral Fellowship, award AST-0901985.  WF acknowledges support
-from NSF grant AST0908930 and NASA grant NNX09AJ56G.
+from NSF grant AST0908930 and NASA grant NNX09AJ56G. DS acknowledges
+support from a Northwestern University Summer Undergraduate Research Grant.
 

### 2013-12-01 paper.tex
Done editing through II.

@@
 \documentclass{iopart}
+%\documentclass{article}
 
@@
 \usepackage{harvard}
+\usepackage{ulem}
+%\usepackage{natbib}
 
@@
 
-\newcommand{\ilya}[1]{{\color{red} \bf Ilya: #1}}
+\newcommand{\ilya}[1]{{\color{red} \bf #1}}
 \newcommand{\will}[1]{{\color{blue} \bf Will: #1}}
@@
 
-\eads{\mailto{w.farr@bham.ac.uk}, \mailto{ilyamandel@chgk.info} \mailto{dan.stevens@u.northwestern.edu}}
+\eads{\mailto{w.farr@bham.ac.uk}, \mailto{imandel@star.sr.bham.ac.uk} \mailto{dan.stevens@u.northwestern.edu}}
 
@@
   jumps between model parameter spaces, but cannot retain a memory of
-  the favored locations in more than one parameter space at a time.
+  the favored locations in more than one parameter space at a time. \ilya{[Not quite true -- see 2.3 -- maybe we can find a better way of saying this in the abstract that doesn't require caveats?]}
   Thus, a naive jump between parameter spaces is unlikely to be
@@
 In a Bayesian context, model selection involves computing the evidence
-for each model given the data.  The model evidence is an integral of
+for \sout{each model given the data} \ilya{the data given each model}.  The model evidence is an integral of
 the unnormalized posterior probability distribution over the model
@@
 directly.  For many problems or models of interest, however, the
-dimensionality of parameter space is too large to make this approach
+dimensionality of the parameter space is too large to make this approach
 practical, and stochastic sampling must be used.
 
-Markov Chain Monte Carlo (MCMC) methods attempt to stochastically
+Markov chain Monte Carlo (MCMC) methods attempt to stochastically
 produce parameter samples with density proportional to the posterior
@@
 
-\cite{Weinberg2009} gives a method for directly computing
+\cite{Weinberg2009} \ilya{[Not ideal reference style, but without citet, I have no idea how to fix this ;) ]}
+gives a method for directly computing
 the evidence integral from existing MCMC samples by using a kD-tree
@@
 what statistical and systematic errors it introduces and how these are
-affected by the shape of the posterior distribution from which the
+affected by the shape of the posterior distribution which the
 MCMC samples.
@@
 unlikely to be accepted, delaying convergence of the RJMCMC sample
-chain.  \cite{Littenberg2009} addressed this issue by proposing a new
+chain \ilya{(see \Sref{RJMCMC} for a caveat)}.  \cite{Littenberg2009} addressed this issue by proposing a new
 method for producing inter-model jumps in an RJMCMC that relies on
@@
 p_{\textnormal{accept}} \equiv \min\Bigl(1,
-\frac{p(\vtheta^p|d)}{p(\vtheta^{(j)}|d)} \frac{Q(\vtheta^p \to
+\frac{p(\vtheta^p|d\ilya{,M})}{p(\vtheta^{(j)}|d\ilya{,M})} \frac{Q(\vtheta^p \to
   \vtheta^{(j)})}{Q(\vtheta^{(j)} \to \vtheta^p)} \Bigr).  
@@
 
-\subsection{RJMCMC}
+\subsection{RJMCMC}\label{RJMCMC}
 
@@
 Sec.~\ref{sec:kDTree}), are described in \cite{Weinberg2009}.  These
-techniques are promising, but in some cases suffer \cite{Farr2010}
-from large variance and bias.  An alternative approach to model
+techniques are promising, but in some cases suffer 
+from large variance and bias \cite{Farr2010}.  An alternative approach to model
 selection among a set of models is based on performing an MCMC in a
@@
 (intermodel jumps).  The resulting chain samples from the posterior
-$p(M_i,\vtheta_i|d)$.  As in a usual MCMC, the PDF on the model as a
+$p(M_i,\ilya{\{\vtheta_i\}}|d)$.  As in a usual MCMC, the PDF on the model as a
 parameter, with other parameters ignored, is obtained by marginalizing
@@
 %
-p(M_i|d) = \int d\vtheta_i \frac{p(d|M_i, \vtheta_i) p(M_i, \vtheta_i)}{p(d)}
+p(M_i|d) = \int d\vtheta_i \frac{\ilya{L}(d|M_i, \vtheta_i) \ilya{p(\vtheta_i|M_i) p(M_i)}}{p(d)}
 \approx \frac{N_i}{N},
@@
 proposed into a model from outside can not access information from
-earlier in the chain which may identify a posterior peak.
+earlier in the chain which may identify a posterior peak. \ilya{(It is possible to overcome this constraint
+by storing a union of $\{\vtheta_i\}$ as the MCMC state vector, with the likelihood a function only of $\vtheta_i$ that correspond to the current $M_i$.
+In this case, intermodel jump proposals would change only $M_i$.
+However, if the chain finds a high-likelihood region in one model space faster than in another, a jump to the other model will take a very long time to be accepted -- again rendering RJMCMC inefficient.)}
 
@@
 single-model MCMC must be interpolated to construct the desired jump
-proposal.  The novel strategy we propose for efficiently interpolating
+proposal.  The strategy we propose for efficiently interpolating
 a discretely sampled PDF is described in the next section.
@@
 \bibliography{paper}
+%\bibliographystyle{unsrt}
 

### 2013-12-01 paper.tex
Finsihed editing.

@@
 ``burn-in'' points.
+\ilya{[Up to this paragraph, we mostly use "samples", not "samples"; for the rest of the paper, we mostly use "points", with occasional sliding back to "samples".
+I don't have a preference, but it might be better to use uniform terminology.]}
 
@@
 a local ``neighborhood'' to each point in the chain of MCMC samples.
-We choose these neighborhoods to be non-overlapping and fill parameter
-space.  To draw a proposed jump, we select a point uniformly from the
+We choose these neighborhoods to be non-overlapping and to fill the parameter
+space.  The size of a
+neighborhood is inversely proportional to the local point density.
+The proposed jumps are drawn from a piecewise-constant (constant on
+each neighborhood) interpolation of the PDF.  
+To draw a proposed jump, we select a point uniformly from the
 MCMC samples, find its associated neighborhood, and then draw the
@@
 this procedure produces proposed jumps that are approximately
-distributed according to the posterior PDF.  The size of a
-neighborhood is inversely proportional to the local point density.
-The proposed jumps are drawn from a piecewise-constant (constant on
-each neighborhood) interpolation of the PDF.  There are various
+distributed according to the posterior PDF.  
+\ilya{[I moved sentences around to improve flow without actually changing the text -- hope that's OK.]}
+
+There are various
 techniques that could be used to construct the set of neighborhoods
 associated with each point.
-
-Ref.~\cite{Littenberg2009} decomposed the parameter space into
-constant-volume ``bricks'' whose size was set by the typical size of
-the peaks of the PDF.  Each sample was associated with the brick that
-contained it, and the probability of proposing a jump into a particular brick is thus proportional to the
-number of samples within that brick.  Additionally, an extra uniform jump proposal was added to allow for jumps into bricks that did not contain any points, so that the jump 
+\cite{Littenberg2009} decompose the parameter space into
+constant-volume ``bricks'' whose size is set by the typical size of
+the peaks of the PDF.  Each sample is associated with the brick that
+contains it, and the probability of proposing a jump into a particular brick is thus proportional to the
+number of samples within that brick.  Additionally, an extra uniform jump proposal is added to allow for jumps into bricks that do not contain any points, so that the jump 
 proposal covers the entire model parameter space.  
@@
 in a sense, maximally adaptive, in contrast to the approach of
-Ref.~\cite{Littenberg2009}, which is minimally adaptive.
-Unfortunately, defining the Voronoi regions requires a metric on
+\cite{Littenberg2009}, which is minimally adaptive.
+Unfortunately, defining the Voronoi decomposition requires a metric on
 parameter space, which may be difficult or impossible to define.
@@
 e.g.\ \cite{Berg2008} or \cite{Gaede1998}).  The decomposition is more
-adaptive than the boxes of Ref.~\cite{Littenberg2009}, and more
+adaptive than the boxes of \cite{Littenberg2009}, and more
 efficient in high-dimensional spaces than the Voronoi decomposition.
@@
 contains all of parameter space.  Then proceed recursively%
-\footnote{The kD-tree datastructure defined here places box boundaries
+\footnote{The kD-tree data structure defined here places box boundaries
   between the points.  An alternate definition common in the
@@
 half in Step 2.  For $n$ points, this can be accomplished in
-$\order{n}$ time (see, e.g., Ref.~\cite{Press2007}).  If there are $N$
+$\order{n}$ time (see, e.g., \cite{Press2007}).  If there are $N$
 points in total, there are $\order{\log N}$ levels in the tree; at
@@
 volume.  The Cauchy distribution is much broader than the Gaussian (it
-has no finite moments), so with uniform model priors, the posterior
+has no finite moments), so with \ilya{equal} model priors, the posterior
 probability for the Gaussian model over the Cauchy model is extremely
@@
   \end{center}
-  \caption{\label{fig:acceptRate} The inter-model acceptance rate
+  \caption{\label{fig:acceptRate} The inter-model \ilya{jump proposal} acceptance rate
     versus the number of points per box when the kD-tree neighborhood
@@
 number of subdivisions in each dimension grows exponentially with
-dimension, so for reasonable numbers of points a higher-dimensional kD
-tree will have fewer subdivisions along each dimension.  Second, as
-the dimensionality increases the volume ``at the corners'' of each kD
+dimension \ilya{number}, so for reasonable numbers of points a high-dimensional kD
+tree will have few subdivisions along each dimension.  Second, as
+the dimensionality increases the volume ``at the \ilya{edges /} corners'' of each kD
 tree box becomes more significant; this problem is even more
@@
 
-To address the arbitrary orientation of points, the modified jump
+\ilya{[There is at least one paragraph missing here.
+What's this "single-model MCMC" and why does it require PDF interpolation?
+Are you running several MCMCs in succession as increasingly good approximations to the desired posterior?
+Are you using higher-temperature chains to provide a jump proposal guess for lower temperatures?
+More details needed.]}
+
+To address the arbitrary orientation of points, \ilya{[Points don't have an orientation -- they are points!  Did you mean something like "To address significant covariance between samples"?]} 
+ the modified jump
 proposal does not descend to the leaf-nodes of the tree, but instead
 stops descending whenever the number of points in the box falls below
-a threshold, $\Nbox$.  We then find the principal axes of the points
+a threshold, $\Nbox$.
+\ilya{[Is $\Nbox$ the threshold, or the actual number of points in the box -- it's defined to be the latter below \Eref{modforward}?]}  We then find the principal axes of the points
 contained in this box and draw from a modified box which is aligned
 with these axes and whose center is the mean of the points contained
-within the rectangular box. The new point is chosen uniformly from
+within the rectangular box.
+\ilya{[Is the box drawn so that all $\Nbox$ samples fall inside it?
+I think so, but it should be stated explicitly.]}  The new point \ilya{[The proposed jump?
+Maybe this would be a good distinction between sample and point -- use point for proposed jumps and samples for the PDF samples being interpolated; if this was done consistently, this would be more clear...]} is chosen uniformly from
 within the bounds of the modified box, subject to the restriction that
@@
 volume of the original box containing $\vtheta_{i+1}$ and $\Vbox$ is
-the volume of the principal-component box containing $\vtheta_{i+1}$.
+the volume of the principal-component \ilya{[First time that "principal-component" is used to identify the diagonal sub-box...]} box containing $\vtheta_{i+1}$.
 
@@
 finding the principal-component box that would contain the current set
-of parameters;.
+of parameters: \ilya{[Probably not necessary -- we don't explicitly provide backward jump proposals in earlier sections]}
 %
@@
 
-It can also be advantageous, particularly in high dimensional spaces,
+It can also be advantageous, particularly in high-dimensional spaces,
 to accumulate more points into the kD tree as the MCMC explores more
-of the parameter space.  Because this introduces a dependence on the
+of the parameter space.  \ilya{[See above -- it's not clear what exactly we are doing here in general.]}
+Because this introduces a dependence on the
 previous states of the MCMC, the jump proposal is no longer
-Markovian.  However, it remains \emph{asymptotically} Markovian, which
+Markovian.
+However, it remains \emph{asymptotically} Markovian \ilya{[And being asymptotically Markovian (whatever that actually means -- at least a reference?
+-- is provably sufficient to guarantee that we sample from the equilibrium distribution even though we nominally violate detailed balance?
+(again, a reference?)]} , which
 ensures that the equilibrium distribution of the MCMC is correct.  To
@@
 amount of work per tree level, and there are $\log N$ tree levels.
+\ilya{[No -- the number of tree levels can be arbitrarily large if the original samples are clustered into a very small region of the prior volume...]}
 Neighborhood lookup proceeds as before, and is therefore also
-$\order{\log N}$.
+$\order{\log N}$.  \ilya{[Perhaps $\order{\log N + \Nbox}$, since must recompile the inner bounding box every time -- and, depending on problem, $\Nbox$ could be $\gg \log N$?]}
 
@@
 proposals in the \texttt{LALInferenceMCMC} sampler
-\cite{Raymond2009,Raymond:2010,Raymond2012}.
-\texttt{LALInferenceMCMC} is a MCMC code, based on the LIGO algorithms
-library, designed to sample the posterior on parameters of
+\cite{vanderSluys:2008a,Raymond:2010,Raymond2012}.
+\texttt{LALInferenceMCMC} is a MCMC code \cite{lalinf_method}, based on the LIGO algorithms
+library, designed to sample the posterior on parameters of \ilya{merging}
 compact-object binaries (masses, orbital orientation, distance, sky
-position, etc) implied by their gravitational wave signals in
+position, etc) \ilya{encoded in} their gravitational wave \ilya{signatures} in
 Earth-based gravitational wave detectors like LIGO.  The simplest such
@@
 the MCMC chain very slow with traditional proposals.  Figure
-\ref{fig:accratio} shows the acceptance ratio for the kD tree jump
+\ref{fig:accratio} shows the acceptance ratio for the kD-tree jump
 proposal in such a simulation.  In spite of the very low acceptance
-rate of the proposal, applying the kD tree proposal to one out of
+rate of the proposal, applying the kD-tree proposal to one out of
 twenty proposed jumps improved the convergence time of the simulation
@@
 \texttt{LALInferenceMCMC}.
+\ilya{[Again, what exactly is reported in this figure?
+Results with a jump proposal based on an interpolated PDF (of what)?
+A proposal that's being updated on the fly as more samples come in?]}
 
@@
     improved the convergence time of the sampler by a factor of two
-    compared to using only the standard suite of proposals.}
+    compared to using only the standard suite of proposals.
+\ilya{[I see structure in this figure (a high acceptance probability, followed by a dip, followed by an asymptote).
+Is this a fluke, or do you always see this -- and, if so, why?
+BTW, what's the autocorrelation length (just to give me a sense of scale, so I am not reading too much into meaningless fluctuations within an AC length)?]}}
 \end{figure}
@@
 variety of physical and astronomical contexts.  In this paper, we
-described a new technique that allows for efficient evidence
+described a technique that allows for efficient evidence
 computations via a Reversible-Jump Markov chain Monte Carlo.  This
 technique solves the usual problem of finding good inter-model jump
-proposals in an RJMCMC by using a kD-tree to quickly and accurately
+proposals in an RJMCMC by using a kD tree to quickly and accurately
 interpolate an approximate posterior PDF from a single-model MCMC run,
-and then using this interpolated PDF to propose efficient inter-model
-jumps.
+and then proposing efficient inter-model
+jumps \ilya{from} this interpolated PDF.
 
@@
 selecting the best model for the observed distribution of black-hole
-X-ray binaries, as described in Ref.~\cite{Farr2010}.  In addition to
+X-ray binary \ilya{masses}, as described in \cite{Farr2010}.  In addition to
 model comparison, the PDF interpolation described here can be useful
-in multi-step MCMCs to improve the sampling efficiency by selecting
-better jump proposal distributions or to test MCMC convergence.
+in multi-step MCMCs to \sout{improve sampling efficiency by selecting
+better jump proposal distributions} \ilya{propose jumps that can sample the parameter space more efficiently} or to test MCMC convergence.
+\ilya{[What are multi-step MCMCs (first mention)?
+How do you use interpolation to test convergence?
+You can't just introduce new ideas or new terminology in the penultimate paragraph of the paper with zero explanation.]}
 

### 2013-12-01 paper.tex
Minor tweaks.

@@
 ``burn-in'' points.
-\ilya{[Up to this paragraph, we mostly use "samples", not "samples"; for the rest of the paper, we mostly use "points", with occasional sliding back to "samples".
+\ilya{[Up to this paragraph, we mostly use "samples", not "points"; for the rest of the paper, we mostly use "points", with occasional sliding back to "samples".
 I don't have a preference, but it might be better to use uniform terminology.]}
@@
 Markovian.
-However, it remains \emph{asymptotically} Markovian \ilya{[And being asymptotically Markovian (whatever that actually means -- at least a reference?
--- is provably sufficient to guarantee that we sample from the equilibrium distribution even though we nominally violate detailed balance?
+However, it remains \emph{asymptotically} Markovian \ilya{[And being asymptotically Markovian (whatever that actually means -- at least a reference?) -- is provably sufficient to guarantee that we sample from the equilibrium distribution even though we nominally violate detailed balance?
 (again, a reference?)]} , which
@@
 How do you use interpolation to test convergence?
-You can't just introduce new ideas or new terminology in the penultimate paragraph of the paper with zero explanation.]}
+I think it's bad form to introduce new ideas or new terminology in the penultimate paragraph of the paper with zero explanation.]}
 
@@
 from NSF grant AST0908930 and NASA grant NNX09AJ56G. DS acknowledges
-support from a Northwestern University Summer Undergraduate Research Grant.
+support from a Northwestern University Summer Undergraduate Research Grant.  \ilya{This work has been partially supported by the UK Science and Technology Facilities Council.}
 

### 2013-12-05 paper.tex
Adding Dan's commits.

@@
 \newcommand{\will}[1]{{\color{blue} \bf Will: #1}}
+\newcommand{\dan}[1]{{\color{magenta} \bf #1}}
 
@@
 
-\eads{\mailto{w.farr@bham.ac.uk}, \mailto{imandel@star.sr.bham.ac.uk} \mailto{dan.stevens@u.northwestern.edu}}
+\eads{\mailto{w.farr@bham.ac.uk}, \mailto{imandel@star.sr.bham.ac.uk} \mailto{stevens@astronomy.ohio-state.edu}}
 
@@
 storage space and $\order{\log N}$ time to produce each proposed jump,
-where $N$ is the number of single-model MCMC samples used to construct
+where $N$ is the number of single-model MCMC \dan{(i.e. an MCMC over a parameter space of only one model)} samples used to construct
 the interpolation.
@@
 MCMCs to generate efficient jump proposal distributions for our
-reversible jump MCMC.  (Single-model MCMCs can take small local jumps
+reversible jump MCMC (Single-model MCMCs can take small local jumps
 within their model, meaning that they are much less likely than an
-RJMCMC to lose a high-posterior mode once it has been located.)  The
+RJMCMC to lose a high-posterior mode once it has been located).  The
 ideal jump proposal distribution for the parameters within a model
@@
 dimensions increases the interpolation becomes more difficult for two
-reasons.  First, the number of sample points required for a given
+reasons.  First, the number of samples required for a given
 number of subdivisions in each dimension grows exponentially with
-dimension \ilya{number}, so for reasonable numbers of points a high-dimensional kD
-tree will have few subdivisions along each dimension.  Second, as
+the number of dimensions; hence, for a reasonable number of samples, a high-dimensional kD tree will have few subdivisions along each dimension.
+\dan{cleaned the syntax up a little in the preceding.} Second, as
 the dimensionality increases the volume ``at the \ilya{edges /} corners'' of each kD
@@
 coordinate axes along which the tree subdivides.  In this section, we
-discuss practical solutions to this issue in the context of a
-single-model MCMC, concluding with an example of a successful
-application to gravitational-wave parameter estimation.
+discuss practical solutions to this issue, \dan{illustrating our solutions with a single-model MCMC. We conclude with an example of a successful
+application to gravitational-wave parameter estimation.}
 
@@
 Are you using higher-temperature chains to provide a jump proposal guess for lower temperatures?
-More details needed.]}
+More details needed.]} \dan{[I added a brief explanation of a single-model MCMC at its earliest mention in the introduction.
+Discussion of the single-model MCMC and the application of our kD-interpolation proposal seems like it would be better discussed after we introduce our proposal, no?
+Will, do you remember how we obtained the original samples that we fed into the tree?]}
 
-To address the arbitrary orientation of points, \ilya{[Points don't have an orientation -- they are points!  Did you mean something like "To address significant covariance between samples"?]} 
- the modified jump
-proposal does not descend to the leaf-nodes of the tree, but instead
-stops descending whenever the number of points in the box falls below
-a threshold, $\Nbox$.
-\ilya{[Is $\Nbox$ the threshold, or the actual number of points in the box -- it's defined to be the latter below \Eref{modforward}?]}  We then find the principal axes of the points
-contained in this box and draw from a modified box which is aligned
-with these axes and whose center is the mean of the points contained
-within the rectangular box.
-\ilya{[Is the box drawn so that all $\Nbox$ samples fall inside it?
-I think so, but it should be stated explicitly.]}  The new point \ilya{[The proposed jump?
-Maybe this would be a good distinction between sample and point -- use point for proposed jumps and samples for the PDF samples being interpolated; if this was done consistently, this would be more clear...]} is chosen uniformly from
-within the bounds of the modified box, subject to the restriction that
-it also lie within the original, coordinate-aligned box.  Figure
-\ref{fig:PCC} illustrates this two-box method.
+\dan{A naive jump proposal based on the kD tree structure would select one sample stored in the tree at random and perform the binary tree search given in Section \ref{sec:kDTree} to find that sample's neighborhood.
+A new point is then proposed uniformly from within this box.
+A better proposal however, would account for covariances between parameters in the parameter space; the naive proposal mentioned above, drawing uniformly from within a box, would yield a low acceptance ratio if parameters are heavily correlated.
+To address significant covariance between samples, the modified jump proposal does not descend to the leaf-nodes of the tree, but instead stops descending whenever the number of samples in the current box, $N_{box}$, falls below some threshold, $N_{crit}$.
+We then find the principal axes of the samples
+contained in this box by calculating the eigenvectors of the covariance matrix of these samples.
+We use these eignevectors to draw a new box.
+This box -- henceforth called a covariance cell -- contains all $N_{box}$ samples from the kD tree box, is centered on the mean of these samples, and has edges that are aligned with the principal axes.
+Furthermore, this box is drawn around the samples as tightly as possible; in other words, for each edge, there exists a point in the covariance cell that lies along that edge.
+Figure
+\ref{fig:PCC} illustrates these two boxes.}
 
@@
 
-The forward jump probability for this jump proposal is given by
+\dan{The jump proposal chooses a stored sample at random and finds the largest kD tree box containing it and fewer than $N_{crit}$ total samples.
+It the draws the covariance cell and proposes a new set of parameters uniformly from the intersection of the covariance cell with the original kD tree box.}
+The jump probability for this jump proposal is given by
 %
 \bel{modforward} P(\vtheta_i \rightarrow \vtheta_{i+1}) =
-\frac{\Nbox}{N \left(V \cap \Vbox \right)} , \ee
+\frac{\Nbox}{N V} , \ee
 %
-where N is the total number of points in the kD-tree, $\Nbox$ is the
-number of points in the box containing $\vtheta_{i+1}$, $V$ is the
-volume of the original box containing $\vtheta_{i+1}$ and $\Vbox$ is
-the volume of the principal-component \ilya{[First time that "principal-component" is used to identify the diagonal sub-box...]} box containing $\vtheta_{i+1}$.
+\dan{where N is the total number of points in the kD-tree, $V$ is the
+volume of the intersection between the kD tree box and the correlation cell containing $\vtheta_{i+1}$. When used as the only jump proposal for an
+RJMCMC, it is important that this proposal be capable of
+proposing points in all allowed regions of parameter space; on the
+other hand, when the kD proposal is used in a suite of other jump proposals, this is no longer a requirement.}
 
-The backward jump probability is computed in a similar fashion by
-finding the principal-component box that would contain the current set
-of parameters: \ilya{[Probably not necessary -- we don't explicitly provide backward jump proposals in earlier sections]}
+%The backward jump probability is computed in a similar fashion by
+%finding the principal-component box that would contain the current set
+%of parameters: \ilya{[Probably not necessary -- we don't explicitly provide backward jump proposals in earlier sections]}
 %
-\bel{modbackward} P(\vtheta_{i+1} \rightarrow \vtheta_{i}) =
-\frac{\Nbox'}{N\left(V' \cap \Vbox'\right)} , \ee
+%\bel{modbackward} P(\vtheta_{i+1} \rightarrow \vtheta_{i}) =
+%\frac{\Nbox'}{N\left(V' \cap \Vbox'\right)} , \ee
 %
-where $\Nbox'$ is the number of points in the original box containing
-$\vtheta_i$, $V'$ is the volume of the kD box containing $\vtheta_i$,
-and $\Vbox'$ is the volume of the principal-axes box containing
-$\vtheta_i$.  Note, however, that the current parameters may lie in a
-region of the kD box that does not intersect the principal axes box;
-in this case, the backward jump probability is zero, and the algorithm
-rejects the proposed step.  When used as the only jump proposal for an
-RJMCMC, it is important that the kD tree proposal be capable of
-proposing points in all allowed regions of parameter space; on the
-other hand, when the kD proposal is used as one of several proposals
-in the context of an MCMC, this is no longer a requirement.
+%where $\Nbox'$ is the number of points in the original box containing
+%$\vtheta_i$, $V'$ is the volume of the kD box containing $\vtheta_i$,
+%and $\Vbox'$ is the volume of the principal-axes box containing
+%$\vtheta_i$.  Note, however, that the current parameters may lie in a
+%region of the kD box that does not intersect the principal axes box;
+%in this case, the backward jump probability is zero, and the algorithm
+%rejects the proposed step.
 
-It can also be advantageous, particularly in high-dimensional spaces,
-to accumulate more points into the kD tree as the MCMC explores more
-of the parameter space.  \ilya{[See above -- it's not clear what exactly we are doing here in general.]}
-Because this introduces a dependence on the
-previous states of the MCMC, the jump proposal is no longer
-Markovian.
+\dan{This technique trivially violates detailed balance since the jump proposal distribution depends on the past history (the number of samples in the kD tree box and in the whole tree).
+One way to address this problem requires continuously accumulating samples into the kD tree as the MCMC explores more
+of the parameter space.
+As the number of samples drawn from the posterior PDF increases, the neighborhoods around each sample decrease with increasing sample density and the calculated covariances between samples in the kD tree better approximate the true covariances between parameters in the equilibrium posterior PDF.
+ence, for large $N$, the change in $P(\vtheta^{(j)} \rightarrow \vtheta^{(j+1)})$ approaches zero and the jump proposal does a progressively better job of sampling the equilibrium posterior [someone confirm or deny!]}  \ilya{[See above -- it's not clear what exactly we are doing here in general.]}
+%Because this introduces a dependence on the
+%previous states of the MCMC, the jump proposal is no longer
+%Markovian.
 However, it remains \emph{asymptotically} Markovian \ilya{[And being asymptotically Markovian (whatever that actually means -- at least a reference?) -- is provably sufficient to guarantee that we sample from the equilibrium distribution even though we nominally violate detailed balance?
 (again, a reference?)]} , which
-ensures that the equilibrium distribution of the MCMC is correct.  To
-efficiently insert points into the tree as the chain accumulates them,
+%ensures that the equilibrium distribution of the MCMC is correct.
+To efficiently insert points into the tree as the chain accumulates them,
 the algorithm from Section \ref{sec:kDTree} must be modified:
@@
 amount of work per tree level, and there are $\log N$ tree levels.
-\ilya{[No -- the number of tree levels can be arbitrarily large if the original samples are clustered into a very small region of the prior volume...]}
+\ilya{[No -- the number of tree levels can be arbitrarily large if the original samples are clustered into a very small region of the prior volume...]} \dan{[Is this true?
+Couldn't it be arbitrarily small, depending on how far into the tree you want to burrow (i.e.
+depending on $N_{crit}$)?]}
 Neighborhood lookup proceeds as before, and is therefore also
-$\order{\log N}$.  \ilya{[Perhaps $\order{\log N + \Nbox}$, since must recompile the inner bounding box every time -- and, depending on problem, $\Nbox$ could be $\gg \log N$?]}
+$\order{\log N}$.
+\ilya{[Perhaps $\order{\log N + \Nbox}$, since must recompile the inner bounding box every time -- and, depending on problem, $\Nbox$ could be $\gg \log N$?]} \dan{[Need only recompute the covariance cell if a new point was accepted into the corresponding kD tree cell between calls to that kD tree cell.
+I'm not sure how that would factor in here, but I feel like it should.] }
 
@@
 \ref{fig:accratio} shows the acceptance ratio for the kD-tree jump
-proposal in such a simulation.  In spite of the very low acceptance
+proposal in such a simulation.
+\dan{For this case, the kD-tree jump proposal is one of several jump proposals used during the MCMC, and the kD tree itself is updated with accepted samples whenever its jump proposal is called.}
+In spite of the very low acceptance
 rate of the proposal, applying the kD-tree proposal to one out of
-twenty proposed jumps improved the convergence time of the simulation
+\dan{20} proposed jumps improved the convergence time of the simulation
 by a factor of two compared to the standard suite of proposals because
@@
   \caption{\label{fig:accratio} The cumulative parameter acceptance
-    ratio for the modified kD-interpolated jump proposal used in the
+    ratio for the modified kD-\dan{tree}jump proposal used in the
     \texttt{LALInferenceMCMC} code as a function of the number of
     steps (in hundreds). In spite of the small acceptance ratio, the
-    kD tree proposal, applied to one in twenty proposed jumps,
+    kD tree proposal, applied to one in \dan{20} proposed jumps,
     improved the convergence time of the sampler by a factor of two
     compared to using only the standard suite of proposals.
-\ilya{[I see structure in this figure (a high acceptance probability, followed by a dip, followed by an asymptote).
+\dan{[Taking a leap here, someone please confirm/deny/add qualifiers] The initial decrease in acceptances arises from insufficient population of the kD tree with too few samples, while the leveling-off later in the MCMC indicates convergence.} \ilya{[I see structure in this figure (a high acceptance probability, followed by a dip, followed by an asymptote).
 Is this a fluke, or do you always see this -- and, if so, why?
-BTW, what's the autocorrelation length (just to give me a sense of scale, so I am not reading too much into meaningless fluctuations within an AC length)?]}}
+BTW, what's the autocorrelation length (just to give me a sense of scale, so I am not reading too much into meaningless fluctuations within an AC length)?]} \dan{[Don't recall the autocorrelation length.
+Will?]}}
 \end{figure}

### 2015-01-13 paper.tex
Partial commit of edits through section IV.

@@
   selection, but it suffers from a fundamental difficulty: it requires
-  jumps between model parameter spaces, but cannot retain a memory of
-  the favored locations in more than one parameter space at a time. \ilya{[Not quite true -- see 2.3 -- maybe we can find a better way of saying this in the abstract that doesn't require caveats?]}
-  Thus, a naive jump between parameter spaces is unlikely to be
+  jumps between model parameter spaces, but cannot efficiently explore both parameter spaces at once.  Thus, a naive jump between parameter spaces is unlikely to be
   accepted in the MCMC algorithm and convergence is correspondingly
@@
 In a Bayesian context, model selection involves computing the evidence
-for \sout{each model given the data} \ilya{the data given each model}.  The model evidence is an integral of
+for the data given each model.  The model evidence is an integral of
 the unnormalized posterior probability distribution over the model
@@
 
-\cite{Weinberg2009} \ilya{[Not ideal reference style, but without citet, I have no idea how to fix this ;) ]}
-gives a method for directly computing
+\cite{Weinberg2009} gives a method for directly computing
 the evidence integral from existing MCMC samples by using a kD-tree
@@
 unlikely to be accepted, delaying convergence of the RJMCMC sample
-chain \ilya{(see \Sref{RJMCMC} for a caveat)}.  \cite{Littenberg2009} addressed this issue by proposing a new
+chain (see \Sref{RJMCMC} for a caveat).  
+\cite{Littenberg2009} addressed this issue by proposing a new
 method for producing inter-model jumps in an RJMCMC that relies on
@@
 storage space and $\order{\log N}$ time to produce each proposed jump,
-where $N$ is the number of single-model MCMC \dan{(i.e. an MCMC over a parameter space of only one model)} samples used to construct
-the interpolation.
+where $N$ is the number of samples in an MCMC over the parameter space of one model (`single-model MCMC') used to construct the interpolation.
 
@@
 p_{\textnormal{accept}} \equiv \min\Bigl(1,
-\frac{p(\vtheta^p|d\ilya{,M})}{p(\vtheta^{(j)}|d\ilya{,M})} \frac{Q(\vtheta^p \to
+\frac{p(\vtheta^p|d, M)}{p(\vtheta^{(j)}|d, M)} \frac{Q(\vtheta^p \to
   \vtheta^{(j)})}{Q(\vtheta^{(j)} \to \vtheta^p)} \Bigr).  
@@
 in the parameter space.  Because jumps are frequently local, we will
-generally want to ignore the early points in a finite-size chain to
+generally want to ignore the early samples in a finite-size chain to
 avoid biases in the recovered posterior PDF due to the choice of the
-initial location.  The points thus discarded are referred to as
-``burn-in'' points.
-\ilya{[Up to this paragraph, we mostly use "samples", not "points"; for the rest of the paper, we mostly use "points", with occasional sliding back to "samples".
-I don't have a preference, but it might be better to use uniform terminology.]}
+initial location.  The samples thus discarded are referred to as
+``burn-in'' samples.
 
@@
 (intermodel jumps).  The resulting chain samples from the posterior
-$p(M_i,\ilya{\{\vtheta_i\}}|d)$.  As in a usual MCMC, the PDF on the model as a
+$p(M_i, \{\vtheta_i\}|d)$.  As in a usual MCMC, the PDF on the model as a
 parameter, with other parameters ignored, is obtained by marginalizing
@@
 %
-p(M_i|d) = \int d\vtheta_i \frac{\ilya{L}(d|M_i, \vtheta_i) \ilya{p(\vtheta_i|M_i) p(M_i)}}{p(d)}
+p(M_i|d) = \int d\vtheta_i \frac{L(d|M_i, \vtheta_i) p(\vtheta_i|M_i) p(M_i)}{p(d)}
 \approx \frac{N_i}{N},
@@
 proposed into a model from outside can not access information from
-earlier in the chain which may identify a posterior peak. \ilya{(It is possible to overcome this constraint
-by storing a union of $\{\vtheta_i\}$ as the MCMC state vector, with the likelihood a function only of $\vtheta_i$ that correspond to the current $M_i$.
-In this case, intermodel jump proposals would change only $M_i$.
-However, if the chain finds a high-likelihood region in one model space faster than in another, a jump to the other model will take a very long time to be accepted -- again rendering RJMCMC inefficient.)}
+earlier in the chain which may identify a posterior peak.
+It is, in principle, possible to overcome this constraint by storing a union of $\{\vtheta_i\}$ as the MCMC state vector, with the likelihood a function only of parameters $\vtheta_i$ that correspond to the current model $M_i$.
+In this case, intermodel jump proposals would change only the model $M_i$.
+However, if the chain finds a high-likelihood region in one model space faster than in another, a jump to the other model will take a very long time to be accepted -- again rendering RJMCMC inefficient.
 
@@
 
-%\ilya{I think this section could just be phrased in terms of interpolation rather than jump proposal distributions; i.e., "the problem of finding the local density" instead of "the problem of drawing a proposed jump", etc.}
-
 The problem of drawing a proposed jump from an interpolation of
-single-model MCMC points can be thought of as the problem of assigning
-a local ``neighborhood'' to each point in the chain of MCMC samples.
+single-model MCMC samples can be thought of as the problem of assigning
+a local ``neighborhood'' to each sample in the MCMC chain.
 We choose these neighborhoods to be non-overlapping and to fill the parameter
 space.  The size of a
-neighborhood is inversely proportional to the local point density.
+neighborhood is inversely proportional to the local sample density.
 The proposed jumps are drawn from a piecewise-constant (constant on
 each neighborhood) interpolation of the PDF.  
-To draw a proposed jump, we select a point uniformly from the
+To draw a proposed jump, we select a sample uniformly from the
 MCMC samples, find its associated neighborhood, and then draw the
-proposed jump uniformly from the neighborhood.  Since the MCMC points
+proposed jump uniformly from the neighborhood.  Since the MCMC samples
 are distributed according to the posterior PDF for the single model,
@@
 distributed according to the posterior PDF.
-\ilya{[I moved sentences around to improve flow without actually changing the text -- hope that's OK.]}
 
@@
 techniques that could be used to construct the set of neighborhoods
-associated with each point.
+associated with each sample.
 \cite{Littenberg2009} decompose the parameter space into
@@
 contains it, and the probability of proposing a jump into a particular brick is thus proportional to the
-number of samples within that brick.  Additionally, an extra uniform jump proposal is added to allow for jumps into bricks that do not contain any points, so that the jump 
+number of samples within that brick.  Additionally, an extra uniform jump proposal is added to allow for jumps into bricks that do not contain any samples, so that the jump 
 proposal covers the entire model parameter space.  
@@
 A kD-tree is a binary, space-partitioning tree.  To partition a set of
-points into a kD-tree, begin by placing them in a rectangular box that
+samples into a kD-tree, begin by placing them in a rectangular box that
 contains all of parameter space.  Then proceed recursively%
 \footnote{The kD-tree data structure defined here places box boundaries
-  between the points.  An alternate definition common in the
-  literature places box boundaries on the median point, but such a
+  between the samples.  An alternate definition common in the
+  literature places box boundaries on the median sample, but such a
   definition is inconvenient for our purposes.}: %
 \begin{enumerate}
-\item If the given box contains exactly one point, stop; this is a
+\item If the given box contains exactly one sample, stop; this is a
   leaf of the tree.  Otherwise:
-\item Choose a dimension along which to divide the points.  Divide the
-  points in half along this dimension (or nearly in half, if the
-  number of points is odd), forming two sub-boxes.  The ``left''
-  sub-box contains the half (or nearly half) of the points that have
+\item Choose a dimension along which to divide the samples.  Divide the
+  samples in half along this dimension (or nearly in half, if the
+  number of samples is odd), forming two sub-boxes.  The ``left''
+  sub-box contains the half (or nearly half) of the samples that have
   small coordinates along the chosen dimension; the ``right'' sub-box
-  contains the half (or nearly half) of the points that have large
+  contains the half (or nearly half) of the samples that have large
   coordinates along the chosen dimension.
@@
 The key algorithmic step in the production of a kD-tree is finding the
-median point along a given dimension in order to divide the points in
-half in Step 2.  For $n$ points, this can be accomplished in
+median sample along a given dimension in order to divide the samples in
+half in Step 2.  For $n$ samples, this can be accomplished in
 $\order{n}$ time (see, e.g., \cite{Press2007}).  If there are $N$
-points in total, there are $\order{\log N}$ levels in the tree; at
-each level, $\order{N}$ points must be processed once in the
+samples in total, there are $\order{\log N}$ levels in the tree; at
+each level, $\order{N}$ samples must be processed once in the
 median-finding algorithm.  Tree construction thus costs $\order{N \log
   N}$ in time, and the tree consumes $\order{N}$ space.  As an
-illustration, box boundaries for a kD-tree constructed around a point
+illustration, box boundaries for a kD-tree constructed around a sample
 set that is normally distributed around the origin in two dimensions
@@
   \caption{\label{fig:kD-tree} The neighborhoods from a kD-tree
-    constructed around a set of points that are normally distributed
-    about the origin in two dimensions.  As the points become denser
+    constructed around a set of samples that are normally distributed
+    about the origin in two dimensions.  As the samples become denser
     around the origin, the typical neighborhood gets smaller.  The
     interpolated PDF within a box of volume $V_i$ is $1/(N V_i)$,
-    where $N$ is the total number of points (which is also the number
+    where $N$ is the total number of samples (which is also the number
     of boxes).}
@@
 volume.  The Cauchy distribution is much broader than the Gaussian (it
-has no finite moments), so with \ilya{equal} model priors, the posterior
+has no finite moments), so with equal model priors, the posterior
 probability for the Gaussian model over the Cauchy model is extremely
@@
 less accurate, we artificially truncated the kD tree with higher and
-higher numbers of points in each box (this can be accomplished during
+higher numbers of samples in each box (this can be accomplished during
 the neighborhood search phase by stopping the search for a box when
-one is found containing the desired number of points).  For each
+one is found containing the desired number of samples).  For each
 truncation choice, we performed an RJMCMC with the resulting
 interpolated jump proposal.  The acceptance rate is plotted against
-the number of single-model MCMC points per box (kD-tree leaf) in
-Figure \ref{fig:acceptRate}.  The more points in each leaf of the tree
+the number of single-model MCMC samples per box (kD-tree leaf) in
+Figure \ref{fig:acceptRate}.  The more samples in each leaf of the tree
 when the search is truncated, the lower the acceptance probability;
@@
   \end{center}
-  \caption{\label{fig:acceptRate} The inter-model \ilya{jump proposal} acceptance rate
-    versus the number of points per box when the kD-tree neighborhood
-    search is truncated.  As the number of points per box increases,
+  \caption{\label{fig:acceptRate} The inter-model jump acceptance rate
+    versus the number of samples per box when the kD-tree neighborhood
+    search is truncated.  As the number of samples per box increases,
     and the interpolation becomes less accurate, the acceptance rate
@@
 tree box becomes more significant; this problem is even more
-pronounced when the distribution of points does not align with the
+pronounced when the distribution of samples does not align with the
 coordinate axes along which the tree subdivides.  In this section, we
@@
     aligned with the axes is the normal kD-tree box containing the
-    given points, while the tighter box is the box from which the
+    given samples, while the tighter box is the box from which the
     modified kd-interpolated jump proposal draws a point. }
@@
 %
-\dan{where N is the total number of points in the kD-tree, $V$ is the
+\dan{where N is the total number of samples in the kD-tree, $V$ is the
 volume of the intersection between the kD tree box and the correlation cell containing $\vtheta_{i+1}$. When used as the only jump proposal for an
@@
 %ensures that the equilibrium distribution of the MCMC is correct.
-To efficiently insert points into the tree as the chain accumulates them,
+To efficiently insert samples into the tree as the chain accumulates them,
 the algorithm from Section \ref{sec:kDTree} must be modified:
 \begin{enumerate}
-\item Instead of partitioning the point set at its median, we now
+\item Instead of partitioning the sample set at its median, we now
   partition the bounding box at its geometric center along a
   particular dimension, which cycles as we descend the tree levels.
-  Note that this allows for empty boxes if all the points cluster to
+  Note that this allows for empty boxes if all the samples cluster to
   one side of the domain.
-\item When inserting a point, we descend the tree to a leaf node,
-  which now can be either empty or contain one point.  If it is empty,
-  we insert the point at this node.  If it contains one point, we
-  subdivide the appropriate dimension, and place both points into the
+\item When inserting a sample, we descend the tree to a leaf node,
+  which now can be either empty or contain one sample.  If it is empty,
+  we insert the sample at this node.  If it contains one sample, we
+  subdivide the appropriate dimension, and place both samples into the
   two sub-boxes.  If both land in the same sub-box, we continue
   subdividing, until each box in the sub-sub-sub... tree contains one
-  or zero points.
+  or zero samples.
 \end{enumerate}
-Insertion of a point is $\order{\log N}$, since it involves a constant
+Insertion of a sample is $\order{\log N}$, since it involves a constant
 amount of work per tree level, and there are $\log N$ tree levels.
@@
 $\order{\log N}$.
-\ilya{[Perhaps $\order{\log N + \Nbox}$, since must recompile the inner bounding box every time -- and, depending on problem, $\Nbox$ could be $\gg \log N$?]} \dan{[Need only recompute the covariance cell if a new point was accepted into the corresponding kD tree cell between calls to that kD tree cell.
+\ilya{[Perhaps $\order{\log N + \Nbox}$, since must recompile the inner bounding box every time -- and, depending on problem, $\Nbox$ could be $\gg \log N$?]} \dan{[Need only recompute the covariance cell if a new sample was accepted into the corresponding kD tree cell between calls to that kD tree cell.
 I'm not sure how that would factor in here, but I feel like it should.] }
@@
 from NSF grant AST0908930 and NASA grant NNX09AJ56G. DS acknowledges
-support from a Northwestern University Summer Undergraduate Research Grant.  \ilya{This work has been partially supported by the UK Science and Technology Facilities Council.}
+support from a Northwestern University Summer Undergraduate Research Grant.  This work has been partially supported by the UK Science and Technology Facilities Council.
 

### 2015-01-13 paper.tex
Finished edits.

@@
 number of subdivisions in each dimension grows exponentially with
-the number of dimensions; hence, for a reasonable number of samples, a high-dimensional kD tree will have few subdivisions along each dimension.
-\dan{cleaned the syntax up a little in the preceding.} Second, as
-the dimensionality increases the volume ``at the \ilya{edges /} corners'' of each kD
-tree box becomes more significant; this problem is even more
+the number of dimensions; hence, for a reasonable number of samples, a high-dimensional kD tree will have few subdivisions along each dimension.  Second, as
+the dimensionality increases, the fraction of the volume at the ``edges'' and ``corners'' of each kD-tree box becomes more significant, and the placement of the sample becomes increasingly unrepresentative of the typical density in the box.
+This problem is even more
 pronounced when the distribution of samples does not align with the
-coordinate axes along which the tree subdivides.  In this section, we
-discuss practical solutions to this issue, \dan{illustrating our solutions with a single-model MCMC. We conclude with an example of a successful
-application to gravitational-wave parameter estimation.}
+coordinate axes along which the tree subdivides.
 
-\ilya{[There is at least one paragraph missing here.
-What's this "single-model MCMC" and why does it require PDF interpolation?
-Are you running several MCMCs in succession as increasingly good approximations to the desired posterior?
-Are you using higher-temperature chains to provide a jump proposal guess for lower temperatures?
-More details needed.]} \dan{[I added a brief explanation of a single-model MCMC at its earliest mention in the introduction.
-Discussion of the single-model MCMC and the application of our kD-interpolation proposal seems like it would be better discussed after we introduce our proposal, no?
-Will, do you remember how we obtained the original samples that we fed into the tree?]}
+In this section, we propose a practical solution that allows us to incorporate larger-scale features in the sample density distribution.
+We illustrate our solution with an single-model MCMC that updates the jump proposal distribution on-the-fly by using a kD tree to interpolate the density of existing samples.
+We conclude with an example of a successful application to gravitational-wave parameter estimation.
 
-\dan{A naive jump proposal based on the kD tree structure would select one sample stored in the tree at random and perform the binary tree search given in Section \ref{sec:kDTree} to find that sample's neighborhood.
-A new point is then proposed uniformly from within this box.
-A better proposal however, would account for covariances between parameters in the parameter space; the naive proposal mentioned above, drawing uniformly from within a box, would yield a low acceptance ratio if parameters are heavily correlated.
-To address significant covariance between samples, the modified jump proposal does not descend to the leaf-nodes of the tree, but instead stops descending whenever the number of samples in the current box, $N_{box}$, falls below some threshold, $N_{crit}$.
+The kD-tree based interpolated jump proposal described in Section \ref{sec:kDTree} selects one sample from the tree at random and proposes a uniform point from the box containing that sample.
+This ignores the density of other samples in the neighborhood of this box, which contains additional information about the true density within the box, which is likely non-uniform.
+A better proposal, therefore, would account for larger-scale sample covariances in parameter space.
+To address significant covariance between samples, which would correspond to strong density gradients and a very non-uniform density profile within a rectangular box, the modified jump proposal does not descend to the leaf-nodes of the tree, but instead stops descending whenever the number of samples in the current box, $N_{box}$, falls below some threshold, $N_{crit}$.
 We then find the principal axes of the samples
 contained in this box by calculating the eigenvectors of the covariance matrix of these samples.
-We use these eignevectors to draw a new box.
-This box -- henceforth called a covariance cell -- contains all $N_{box}$ samples from the kD tree box, is centered on the mean of these samples, and has edges that are aligned with the principal axes.
+We use these eigenvectors to draw a new box.
+This box -- henceforth called a covariance cell -- contains all $N_{box}$ samples from the kD-tree box, is centered on the mean of these samples, and has edges that are aligned with the principal axes.
 Furthermore, this box is drawn around the samples as tightly as possible; in other words, for each edge, there exists a point in the covariance cell that lies along that edge.
-Figure
-\ref{fig:PCC} illustrates these two boxes.}
+Figure \ref{fig:PCC} illustrates these two boxes.
 
@@
     aligned with the axes is the normal kD-tree box containing the
-    given samples, while the tighter box is the box from which the
-    modified kd-interpolated jump proposal draws a point. }
+    given samples, while the tighter box is the ``covariance cell'' aligned with the eigenvectors of the sample distribution.
+The modified kd-interpolated jump proposal draws a point from the intersection of the two boxes.
+}
 \end{figure}
 
-\dan{The jump proposal chooses a stored sample at random and finds the largest kD tree box containing it and fewer than $N_{crit}$ total samples.
-It the draws the covariance cell and proposes a new set of parameters uniformly from the intersection of the covariance cell with the original kD tree box.}
+The jump proposal chooses a stored sample at random and finds the largest kD-tree box containing it and fewer than $N_{crit}$ total samples.
+It then draws the covariance cell and proposes a new point uniformly from the intersection of the covariance cell with the original kD-tree box.
 The jump probability for this jump proposal is given by
@@
 %
-\dan{where N is the total number of samples in the kD-tree, $V$ is the
-volume of the intersection between the kD tree box and the correlation cell containing $\vtheta_{i+1}$. When used as the only jump proposal for an
+where $N$ is the total number of samples in the kD tree, and $V$ is the
+volume of the intersection between the kD-tree box and the correlation cell containing $\vtheta_{i+1}$. When used as the only jump proposal for an
 RJMCMC, it is important that this proposal be capable of
 proposing points in all allowed regions of parameter space; on the
-other hand, when the kD proposal is used in a suite of other jump proposals, this is no longer a requirement.}
+other hand, when the kD proposal is used in a suite of other jump proposals, this is no longer a requirement.
 
@@
 
-\dan{This technique trivially violates detailed balance since the jump proposal distribution depends on the past history (the number of samples in the kD tree box and in the whole tree).
+This technique trivially violates detailed balance since the jump proposal distribution depends on the past history.
 One way to address this problem requires continuously accumulating samples into the kD tree as the MCMC explores more
@@
 As the number of samples drawn from the posterior PDF increases, the neighborhoods around each sample decrease with increasing sample density and the calculated covariances between samples in the kD tree better approximate the true covariances between parameters in the equilibrium posterior PDF.
-ence, for large $N$, the change in $P(\vtheta^{(j)} \rightarrow \vtheta^{(j+1)})$ approaches zero and the jump proposal does a progressively better job of sampling the equilibrium posterior [someone confirm or deny!]}  \ilya{[See above -- it's not clear what exactly we are doing here in general.]}
+Hence, for large $N$, the change in $P(\vtheta^{(j)} \rightarrow \vtheta^{(j+1)})$ approaches zero and the jump proposal does a progressively better job of sampling the equilibrium posterior while becoming asymptotically Markovian.
+Conservatively, the entire set of samples obtained while the jump proposal is being dynamically updated could be discarded as burn-in, and future analysis could use the static jump proposal distribution interpolated from samples accumulated during the extended burn-in phase.
 %Because this introduces a dependence on the
@@
 %ensures that the equilibrium distribution of the MCMC is correct.
+
 To efficiently insert samples into the tree as the chain accumulates them,
@@
 \end{enumerate}
-Insertion of a sample is $\order{\log N}$, since it involves a constant
-amount of work per tree level, and there are $\log N$ tree levels.
+%Insertion of a sample is $\order{\log N}$, since it involves a constant amount of work per tree level, and there are $\log N$ tree levels.
 \ilya{[No -- the number of tree levels can be arbitrarily large if the original samples are clustered into a very small region of the prior volume...]} \dan{[Is this true?
 Couldn't it be arbitrarily small, depending on how far into the tree you want to burrow (i.e.
-depending on $N_{crit}$)?]}
-Neighborhood lookup proceeds as before, and is therefore also
-$\order{\log N}$.
+depending on $N_{crit}$)?]} Neighborhood lookup proceeds as before, and is therefore also $\order{\log N}$.
 \ilya{[Perhaps $\order{\log N + \Nbox}$, since must recompile the inner bounding box every time -- and, depending on problem, $\Nbox$ could be $\gg \log N$?]} \dan{[Need only recompute the covariance cell if a new sample was accepted into the corresponding kD tree cell between calls to that kD tree cell.
@@
 proposals in the \texttt{LALInferenceMCMC} sampler
-\cite{vanderSluys:2008a,Raymond:2010,Raymond2012}.
-\texttt{LALInferenceMCMC} is a MCMC code \cite{lalinf_method}, based on the LIGO algorithms
-library, designed to sample the posterior on parameters of \ilya{merging}
-compact-object binaries (masses, orbital orientation, distance, sky
-position, etc) \ilya{encoded in} their gravitational wave \ilya{signatures} in
-Earth-based gravitational wave detectors like LIGO.  The simplest such
+\cite{vanderSluys:2008a,Raymond:2010,Raymond2012,Veitch:2014}.
+\texttt{LALInferenceMCMC} is a MCMC code, based on the LIGO algorithms
+library ({\tt http://www.lsc-group.phys.uwm.edu/lal}), designed to sample the posterior on parameters of merging
+compact-object binaries (masses, sky location, orbital orientation, distance, etc.) 
+encoded in their gravitational-wave signatures as observed by the ground-based
+gravitational-wave detectors LIGO \cite{AdvLIGO} and Virgo \cite{AdvVirgo}.  The simplest such
 signal has a nine-dimensional parameter space, and the posterior often
 includes near-degeneracies and multiple modes that make convergence of
-the MCMC chain very slow with traditional proposals.  Figure
-\ref{fig:accratio} shows the acceptance ratio for the kD-tree jump
-proposal in such a simulation.
-\dan{For this case, the kD-tree jump proposal is one of several jump proposals used during the MCMC, and the kD tree itself is updated with accepted samples whenever its jump proposal is called.}
+the MCMC chain very slow with traditional proposals \cite{S6PE}.
+
+Figure \ref{fig:accratio} shows the acceptance ratio for the kD-tree jump
+proposal applied as part of a \texttt{LALInferenceMCMC} analysis.
+In this example, the kD-tree jump proposal is one of several jump proposals used during the MCMC, and the kD tree itself is updated with accepted samples whenever its jump proposal is called.
 In spite of the very low acceptance
 rate of the proposal, applying the kD-tree proposal to one out of
-\dan{20} proposed jumps improved the convergence time of the simulation
+20 proposed jumps improved the convergence time of the simulation
 by a factor of two compared to the standard suite of proposals because
@@
 \texttt{LALInferenceMCMC}.
-\ilya{[Again, what exactly is reported in this figure?
-Results with a jump proposal based on an interpolated PDF (of what)?
-A proposal that's being updated on the fly as more samples come in?]}
 
@@
   \caption{\label{fig:accratio} The cumulative parameter acceptance
-    ratio for the modified kD-\dan{tree}jump proposal used in the
+    ratio for the modified kD-tree jump proposal used in the
     \texttt{LALInferenceMCMC} code as a function of the number of
     steps (in hundreds). In spite of the small acceptance ratio, the
-    kD tree proposal, applied to one in \dan{20} proposed jumps,
+    kD tree proposal, applied to one in 20 proposed jumps,
     improved the convergence time of the sampler by a factor of two
     compared to using only the standard suite of proposals. 
-\dan{[Taking a leap here, someone please confirm/deny/add qualifiers] The initial decrease in acceptances arises from insufficient population of the kD tree with too few samples, while the leveling-off later in the MCMC indicates convergence.} \ilya{[I see structure in this figure (a high acceptance probability, followed by a dip, followed by an asymptote).
+    The acceptance rate asymptotes to the steady-state solution once sufficient samples have been accumulated in the kD tree to allow the sample density to be accurately interpolated.
+%\dan{[Taking a leap here, someone please confirm/deny/add qualifiers] The initial decrease in acceptances arises from insufficient population of the kD tree with too few samples, while the leveling-off later in the MCMC indicates convergence.} \ilya{[I see structure in this figure (a high acceptance probability, followed by a dip, followed by an asymptote).
 Is this a fluke, or do you always see this -- and, if so, why?
 BTW, what's the autocorrelation length (just to give me a sense of scale, so I am not reading too much into meaningless fluctuations within an AC length)?]} \dan{[Don't recall the autocorrelation length.
-Will?]}}
+Will?]}
+}
 \end{figure}
@@
 and then proposing efficient inter-model
-jumps \ilya{from} this interpolated PDF.
+jumps from this interpolated PDF.
 
@@
 selecting the best model for the observed distribution of black-hole
-X-ray binary \ilya{masses}, as described in \cite{Farr2010}.  In addition to
+X-ray binary masses, as described in \cite{Farr2010}.  In addition to
 model comparison, the PDF interpolation described here can be useful
-in multi-step MCMCs to \sout{improve sampling efficiency by selecting
-better jump proposal distributions} \ilya{propose jumps that can sample the parameter space more efficiently} or to test MCMC convergence.
-\ilya{[What are multi-step MCMCs (first mention)?
-How do you use interpolation to test convergence?
-I think it's bad form to introduce new ideas or new terminology in the penultimate paragraph of the paper with zero explanation.]}
+in single-model MCMCs to inform the jump proposal distribution on the fly in order to propose jumps that can efficiently sample the parameter space (see Section \ref{sec:higherdim}), or to test MCMC convergence.
 

### 2015-01-15 paper.tex
Reformatted to RSOS.

@@
-\documentclass{iopart}
+\documentclass{rsos}
 %\documentclass{article}
@@
 
-\author{W M Farr$^{1,3}$, I Mandel$^{2,3}$ and D Stevens$^{1,4}$}
+\author{W M Farr$^{1,3}$, I Mandel$^{2,3,4}$ and D Stevens$^{1,5}$}
 
-\address{$^1$ Center for Interdisciplinary Exploration and Research in
+\address{\small{$^1$ Center for Interdisciplinary Exploration and Research in
   Astrophysics (CIERA), Department of Physics and Astronomy,
-  Northwestern University, Evanston IL USA}
+  Northwestern University, Evanston IL USA\\
+$^2$ MIT Kavli Institute, Cambridge MA USA\\
+$^3$ School of Physics and Astronomy, University of
+  Birmingham, Birmingham UK\\
+$^4$ Monash Center for Astrophysics and School of Physics and Astronomy, Monash University, Clayton, VIC Australia\\
+$^5$ Department of Astronomy, The Ohio State
+  University, Columbus OH USA}}
 
-\address{$^2$ MIT Kavli Institute, Cambridge MA USA}
+\subject{Astrostatistics, Data Analysis, Markov chain Monte Carlo}
 
-\address{$^3$ School of Physics and Astronomy, University of
-  Birmingham, Birmingham UK}
+\keywords{MCMC, RJMCMC}
+%\eads{\mailto{w.farr@bham.ac.uk}, \mailto{imandel@star.sr.bham.ac.uk} \mailto{stevens@astronomy.ohio-state.edu}}
 
-\address{$^4$ Department of Astronomy, The Ohio State
-  University, Columbus OH USA}
-
-\eads{\mailto{w.farr@bham.ac.uk}, \mailto{imandel@star.sr.bham.ac.uk} \mailto{stevens@astronomy.ohio-state.edu}}
+\corres{Ilya Mandel\\
+\email{imandel@star.sr.bham.ac.uk}}
 
@@
 
-\pacs{02.70.Tt, 95.75.Pq, 07.05.Kf}
+%\pacs{02.70.Tt, 95.75.Pq, 07.05.Kf}
 
-\submitto{\CQG}
+%\submitto{\CQG}
 
@@
 unlikely to be accepted, delaying convergence of the RJMCMC sample
-chain (see \Sref{RJMCMC} for a caveat).  
+chain (see Section \ref{RJMCMC} for a caveat).  
 \cite{Littenberg2009} addressed this issue by proposing a new
@@
 particular model, $M_i$, given the data, $p(M_i|d)$.  Using Bayes'
-rule, we see that this involves the evidence \eref{evidence}:
+rule, we see that this involves the evidence, Eq.~(\ref{evidence}):
 \be
@@
 attempts to directly compute the evidence by performing the
-integration in \eref{evidence} are generally very difficult in a
+integration in Eq.~(\ref{evidence}) are generally very difficult in a
 multi-dimensional, multi-modal parameter space when the likelihood has
@@
 from NSF grant AST0908930 and NASA grant NNX09AJ56G. DS acknowledges
-support from a Northwestern University Summer Undergraduate Research Grant.  This work has been partially supported by the UK Science and Technology Facilities Council.
+support from a Northwestern University Summer Undergraduate Research Grant.
+IM is grateful for the hospitality offered by Monash University and, particularly, Yuri Levin.
+This work has been partially supported by the UK Science and Technology Facilities Council.
 
-\section*{References}
+\section*{}
 

### 2015-03-22 Response.txt
Starting work on response.

Dear Editor:

We are grateful to the referees of our paper, "An Efficient Interpolation Technique for Jump Proposals in Reversible-Jump Markov Chain Monte Carlo Calculations" (Farr, Mandel, Stevens) for their many helpful comments and suggestions.  We detail our responses to the referees below.  The re-submitted manuscript carries these changes in addition to a few minor revisions, such as an update of the reference list.

Sincerely yours,
Will Farr, Ilya Mandel, Daniel Stevens

----

Report 1:

> I am familiar with an earlier draft of this paper from internal LIGO review, and I'm pleased to see that it > is finally being submitted for publication.

...

> One minor quibble that I have is that the language in the introduction makes it seem that the underlying > idea is different. Both methods "construct an approximation to each
> modelÕs posterior parameter distribution". The difference is that the kDE tree approach is far superior to > the original sparse histogram approach.

We reformulated the sentence in question as follows to remove the ambiguity and make it clear that we are proposing an improved interpolation technique for constructing an approximate posterior, not a fundamentally different idea:

"...Here we introduce an alternative technique based on a kD-tree data  structure to construct an approximation to each model's posterior parameter distribution.  This improved interpolation method leads to dramatically faster convergence of RJMCMC sample chains."

> Other minor point is that the reference to parallel tempering does not point to the original paper on the > topic (Swendsen and Wang 1986).

We included the missing reference.

----

Report 2:

>>>This paper addresses an interesting and important issue in Bayesian model selection: Developing
efficient jump proposals in reversible-jump Markov-chain Monte Carlo (RJMCMC)
algorithms to transition between the models under consideration (high rejection rates for
inter-model proposals represent the limiting factor in many applications). The approach
proposed here|to exploit sampling of the posterior probability density (PPD) in the individual
model spaces, characterized as a kD tree structure|seems like a very promising idea
which could justify publication.

We thank the referee for this assessment.

>>> However, there are a number of significant issues that must be addressed before publication is recommended.

We address these issues below, and in the text where appropriate.

>>> 1. The history of this work is unclear and concerning. The  first paragraph of page 4 says
"We have successfully applied this RJMCMC technique to a 10-way model selection
among alternative mass distribution models for black-hole X-ray binaries (Farr et al.
2010)". This reference is again cited on page 13 as a successful application of the
method developed in the present manuscript. But this raises an obvious question:
Has the method described in the present manuscript already been published by the
authors? If so, publication of the present manuscript does not seem justified since the
it considers only a trivial toy problem which adds nothing of significance. Further,
the reference section indicates that Farr et al. 2010 is in the submission stage to the
Astrophysical Journal. What's the story here|is this a typo, or a badly out-of-date
reference, or has the paper actually been in-review for five years? Is the method
described in the present manuscript the same as in the 2010 paper (sounds like it is)?
This must be sorted out, and unless the present manuscript contains new concepts
compared to earlier work, publication is not justified.

The current paper does not repeat earlier publications, and includes key new material that has not been previously published.  The Farr et al., paper on "The Mass Distribution of Stellar-mass Black Holes" has been published in Astrophys. J. 741, 103, in 2011 (reference updated -- we apologize for the out-of-date reference which added to the confusion).  Farr+, 2011 dealt with measuring the mass distribution of stellar mass black holes in low-mass and high-mass X-ray binaries.  That analysis made use of an early implementation of the method described in this paper, and, consequently, a very brief summary of the method was included in Farr+, 2011.  However, the focus of Farr+, 2011, was on the astrophysical results rather than the method; consequently, the details of the algorithm and the thorough analysis presented here were not included in Farr+, 2011.  Moreover, the present paper includes a number of more recent (and unpublished) improvements to the method, such as the use of "rotated" boxes to improve the interpolation quality in higher-dimensional spaces (see Fig. 3 and associated text).  Finally, the use of the kD-interpolation scheme for adaptive sampling improvements in a single-model MCMC (Section 5 of the present paper) is entirely new.  Figure 4 illustrates the acceptance ratio of this single-model MCMC as applied to the very computationally challenging problem of parameter estimation on gravitational-wave parameter estimation (contrary to the referee's assertion that only toy models have been used for illustration).

>>> 2. The explanation of the inter-model proposal and acceptance criterion is Section 3c is
inadequate. While well-known material on Bayesian analysis (Section 3a) and MCMC
(Section 3b) is presented at length with equations, the new RJMCMC approach is
skimmed over in a single sentence (no equations) as \We perform MCMC in the supermodel
parameter space just like a regular MCMC; we propose jumps to different parameters
within a model (intra-model jumps) and jumps to a different model with
different parameters (inter-model jumps)" (page 6)...

>>> 3. The examples given in the manuscript seem simplified to the point of triviality and/or
are poorly explained. The only example given of model selection (the potentially important
contribution of this paper) only involves sampling between 1D Gaussian and
Cauchy distributions (two unknown parameters each). While this does illustrate the
method (barely), it is of no practical relevance and is so much simpler than most realistic
model-selection problems that it is not clear how efficiency and performance
would scale, or how useful the method could be in practice. I assume the authors are
researching real-world model-selection problems of practical interest (like alternative
mass distribution models for black-hole X-ray binaries, mentioned on page 4). Including
a meaningful example such as this would greatly improve the paper and increase
its significance. The other two examples represent only single-model sampling and are
of much less interest, but even so they should be properly explained. In the examples
in Figures 3 and 4, what are the problems, what are the parameters, and what do the
results mean? This current presentation is not satisfying to the reader.

>>> 4. Claims are made of \dramatically improved convergence" (page 3), that \convergence
is generally rapid" (page 4), and \the interpolation method described above can dramatically
improve the runtime of an RJMCMC" (page 10). While I believe the method
developed here could substantially improve efficiency in some cases, this is not actually
demonstrated in the paper. The trivial example in Figure 2 appears to increase
inter-model acceptance by a factor of 8 or so, but no results on RJMCMC sampling
convergence are given. The example considered in Figure 4 claims to have improved
convergence by a factor of two (not dramatic), but there is no discussion of what definition is applied for convergence|and as noted on page 6 (third paragraph), defining
convergence for MCMC sampling is a tricky business. Even applying a reasonable
working definition of convergence, given the random nature of MCMC sampling, averages
over several MCMC runs are often required to meaningfully compare convergence
rates -- was that done here? While the claims of \dramatic" improvements in convergence may be true in some cases, they don't appear to be properly established in the paper.

>>>  5a. Notation: While notation differs between disciplines, I am used to boldface (rather
than over-arrows) being used to represent vectors in published work -- the authors
could consider this. Further, why is the set of parameters written as a vector,
but the set of data not written as a vector (d)?

We write parameters as a vector because we occasionally consider individual parameters or subsets of the parameter set; no such confusion is likely for the data.

>>> 5b. Page 5, rst line after Section 2b heading: Should be j = 1; : : : not i = 1; : : :.
(c) Page 5, third line after Section 2b heading: Should be \set of parameters" not
\set of parameter".
(d) Figure 1 would be clearer if parameter axes were added indicating the origin, with
axes scales in terms of standard deviations in x and y.
(e) The x axis of Figure 2 is labeled Nboxing while it should be Nbox.
(f) Figure 3 caption: kd should be kD.
(g) Equation (5.1): P should be Q (this is the proposal distribution). Also, V might
be better written as Vbox to indicate that it is the volume of the current box, like
Nbox indicates the number of samples in the current box (V is not a constant but
varies with boxes).
(h) Page 11: Inconsistent use of math or Roman font for some subscripts; e.g., Nbox
and Nbox, and Ncrit. When the subscript is a label rather than an index, Roman
font is preferred.
(i) Page 11, after Eq. (5.1): Please give an equation for V , the volume of the intersection
of the two (potentially high-dimensional) boxes.
(j) Page 12, first paragraph discusses violations of detailed balance arising from using
the sampling history to define proposals. Such methods are well known, including
the requirements which must be met to avoid biased sampling (so-called diminishing
adaptations). References (such as Brooks et al., 2011 Handbook of Markov
chain Monte Carlo, New York: Springer) should be included here to be clear this
is not a new idea.
(k) Figure 4's caption states \The acceptance probability rate asymptotes to the
steady-state solution once sufficient samples are accumulated for the kD tree
to allow the sample density or be accurately interpolated." This asymptotic
convergence appears to occur by about step 2.0106; hence, all samples collected
prior to this point should be discarded as burn-in or the sampling is potentially
biased. This should be mentioned explicitly in the paper.
(l) Figure 4: Why denote the x axis divisions from 0 to 25 000 and then include
\( X100)" in the axis label? Why not denote the divisions from 0 to 2.5 and
include \(X 10^6)" in the label|this is the common use of exponential notation.

----

### 2015-03-22 paper.tex
Starting work on response.

@@
 \begin{abstract}
-  Selection among alternative theoretical models given an observed
-  data set is an important challenge in many areas of physics and
-  astronomy.  Reversible-jump Markov chain Monte Carlo (RJMCMC) is an
-  extremely powerful technique for performing Bayesian model
-  selection, but it suffers from a fundamental difficulty: it requires
-  jumps between model parameter spaces, but cannot efficiently explore both parameter spaces at once.  Thus, a naive jump between parameter spaces is unlikely to be
-  accepted in the MCMC algorithm and convergence is correspondingly
-  slow.  Here we demonstrate an interpolation technique that uses
-  samples from single-model MCMCs to propose inter-model jumps from an
-  approximation to the single-model posterior of the target parameter
-  space.  The interpolation technique, based on a kD-tree data
-  structure, is adaptive and efficient in modest dimensionality.  We
-  show that our technique leads to dramatically improved convergence
-  over naive jumps in an RJMCMC, and compare it to other proposals in
-  the literature to improve the convergence of RJMCMCs.  We also
-  demonstrate the use of the same interpolation technique as a way to
-  construct efficient ``global'' proposal distributions for
-  single-model MCMCs without prior knowledge of the structure of the
-  posterior distribution, and discuss improvements that permit the
-  method to be used in higher-dimensional spaces efficiently.
+Selection among alternative theoretical models given an observed data set is an important challenge in many areas of physics and astronomy.
+Reversible-jump Markov chain Monte Carlo (RJMCMC) is an extremely powerful technique for performing Bayesian model selection, but it suffers from a fundamental difficulty: it requires jumps between model parameter spaces, but cannot efficiently explore both parameter spaces at once.
+Thus, a naive jump between parameter spaces is unlikely to be accepted in the MCMC algorithm and convergence is correspondingly slow.
+Here we demonstrate an interpolation technique that uses samples from single-model MCMCs to propose inter-model jumps from an approximation to the single-model posterior of the target parameter space.
+The interpolation technique, based on a kD-tree data structure, is adaptive and efficient in modest dimensionality.
+We show that our technique leads to dramatically improved convergence over naive jumps in an RJMCMC, and compare it to other proposals in the literature to improve the convergence of RJMCMCs.
+We also demonstrate the use of the same interpolation technique as a way to construct efficient ``global'' proposal distributions for single-model MCMCs without prior knowledge of the structure of the posterior distribution, and discuss improvements that permit the method to be used in higher-dimensional spaces efficiently.
 \end{abstract}
@@
 \cite{NewtonRaftery:1994,Chib:1995,vanHaasteren:2009}.  MCMC
-implementations with parallel tempering \cite{EarlDeem:2005} allow
+implementations with parallel tempering \cite{SwendsenWang:1986,EarlDeem:2005} allow
 for evidence computation via thermodynamic integration, but these can
@@
 
-Here we introduce an alternative technique for improving the
-acceptance ratio of inter-model jumps in an RJMCMC, leading to
-dramatically improved convergence of RJMCMC sample chains.  The
-technique uses a kD-tree data structure to construct an approximation
-to each model's posterior parameter distribution.  We draw jump
-proposals into the model from this approximation to its posterior.
+Here we introduce an alternative technique based on a kD-tree data 
+structure to construct an approximation
+to each model's posterior parameter distribution.  This improved interpolation method leads to dramatically faster convergence of RJMCMC sample chains.
+We draw jump proposals into the model from this approximation to its posterior.
 Because jumps are proposed preferentially to locations favored by the
@@
 acknowledges support from the NSF Astronomy and Astrophysics
-Postdoctoral Fellowship, award AST-0901985.  WF acknowledges support
+Postdoctoral Fellowship, award AST-0901985, for the early stages of this work.  
+WF acknowledges support
 from NSF grant AST0908930 and NASA grant NNX09AJ56G. DS acknowledges

### 2015-04-06 paper.tex
Implemented Ewan's suggestions.

@@
 the evidence for a model from MCMC samples is to compute the
-harmonic-mean estimator, but this estimator of the evidence suffers
+harmonic-mean estimator, but this estimator of the evidence  can suffer
 from infinite variance 
-\cite{NewtonRaftery:1994,Chib:1995,vanHaasteren:2009}.  MCMC
+\cite{NewtonRaftery:1994,Chib:1995,vanHaasteren:2009,WolpertSchmidler:2012}.  MCMC
 implementations with parallel tempering \cite{SwendsenWang:1986,EarlDeem:2005} allow
-for evidence computation via thermodynamic integration, but these can
+for evidence computation via thermodynamic integration \cite{FrielPettitt:2008}, but these can
 be computationally costly.
@@
 fully in Section \ref{sec:reversible-jump}, performs a standard MCMC
-in a superspace that is a direct sum of all the model parameter
+in space that is an augmented union of all the model parameter
 spaces.  Such an MCMC involves both intra- and inter-model jumps; the
@@
 requirement for an MCMC arises from the fact that the desired
-distribution is the equilibrium distribution.  In other words, if we
-assume that state $(j)$ of the chain is sampled from the desired PDF,
-$p(\vtheta^{(j)})=p(\vtheta^{(j)}|d,M)$ , then the next state $(j+1)$ must
-be sampled from the PDF as well, so that
-$p(\vtheta^{(j+1)})=p(\vtheta^{(j+1)}|d,M)$; this condition is known as
-``detailed balance.''
+distribution is the equilibrium distribution.  Detailed balance requires that
+$p(\vtheta^{(i)}) p(\vtheta^{(i)} \to \vtheta^{(j)}) = p(\vtheta^{(j)}) p(\vtheta^{(j)} \to \vtheta^{(i)})$.
+
+% In other words, if we assume that state $(j)$ of the chain is sampled from the desired PDF, $p(\vtheta^{(j)})=p(\vtheta^{(j)}|d,M)$ , then the next state $(j+1)$ must be sampled from the PDF as well, so that $p(\vtheta^{(j+1)})=p(\vtheta^{(j+1)}|d,M)$; this condition is known as ``detailed balance.''
 
@@
 perform a Monte Carlo evidence integral.  This results in a harmonic
-mean estimator for the evidence, which suffers from a large variance
-and bias \cite{NewtonRaftery:1994,Chib:1995,vanHaasteren:2009}.
+mean estimator for the evidence, which may suffer from infinite variance \cite{NewtonRaftery:1994,Chib:1995,vanHaasteren:2009,WolpertSchmidler:2012}.
 Additional techniques for the direct integration of evidence, also
@@
 
-We are grateful to Neil Cornish for interesting discussions.  IM
+We are grateful to Neil Cornish for interesting discussions.  WF acknowledges support
+from NSF grant AST0908930 and NASA grant NNX09AJ56G.  IM
 acknowledges support from the NSF Astronomy and Astrophysics
 Postdoctoral Fellowship, award AST-0901985, for the early stages of this work.  
-WF acknowledges support
-from NSF grant AST0908930 and NASA grant NNX09AJ56G. DS acknowledges
+ DS acknowledges
 support from a Northwestern University Summer Undergraduate Research Grant.
-IM is grateful for the hospitality offered by Monash University and, particularly, Yuri Levin.
+IM is grateful for the hospitality of the Monash Center for Astrophysics supported by a Monash Research Acceleration Grant (PI Y.~Levin).
 This work has been partially supported by the UK Science and Technology Facilities Council.

### 2015-04-20 Response.txt
Changes to address first reviewer's comments + response.

@@
 
+We have added descriptive text at the bottom of page 5 and top of page 6 that expands on the RJMCMC approach.
+The text now makes clear that the jump proposal ratio is a ratio of densities in the two model spaces, and discusses the special case where the spaces are of equal dimension and the proposal is a diffeomorphism between them (i.e.\ not a probabilistic choice of jump point, but rather a fixed mapping), where the ratio reduces to the Jacobian factor to which the referee refers.
+We have not written the equation as given by the referee,
+
+p_accept = (p(theta')*Q(theta' -> theta))/(p(theta)*Q(theta -> theta')) * |J|
+
+because this form is not correct in our notation.  The factors Q *already incorporate the Jacobian* in our notation.  This point is now emphasised in the text.
+
 >>> 3. The examples given in the manuscript seem simplified to the point of triviality and/or
@@
 
+We have added text to the beginning of Section 4 pointing to the use of this algorithm in real-world applications, such as those of Farr et al (2011).
+We feel that to describe the (lengthy) analysis from that work would needlessly complicate the presentation of the algorithm in this paper, but hope that this emphasises that the algorithm is, indeed, applicable to complicated model selection problems.
+We also emphasise in that text that the algorithm is available as a software library for others interested in adopting it.
+
+We have added text to Figures 3 and 4 describing the gravitational wave data analysis problem and the associated parameter space briefly.
+Again, we feel that a complete description of this problem (available from Veitch et al (2015) and the other references in the text) is beyond the scope of the article, but we understand the need for more context in the explanation of this paper.
+We hope the additional text provides this context.
+
 >>> 4. Claims are made of \dramatically improved convergence" (page 3), that \convergence
@@
 rates -- was that done here? While the claims of \dramatic" improvements in convergence may be true in some cases, they don't appear to be properly established in the paper.
+
+We have removed the (perhaps over-wrought) adjective "dramatic" from our claims of convergence, and included text briefly describing the method of computing convergence times used and referring the reader for details to Veitch et al (2015).
 

### 2015-04-20 paper.tex
Changes to address first reviewer's comments + response.

@@
 The interpolation technique, based on a kD-tree data structure, is adaptive and efficient in modest dimensionality.
-We show that our technique leads to dramatically improved convergence over naive jumps in an RJMCMC, and compare it to other proposals in the literature to improve the convergence of RJMCMCs.
+We show that our technique leads to improved convergence over naive jumps in an RJMCMC, and compare it to other proposals in the literature to improve the convergence of RJMCMCs.
 We also demonstrate the use of the same interpolation technique as a way to construct efficient ``global'' proposal distributions for single-model MCMCs without prior knowledge of the structure of the posterior distribution, and discuss improvements that permit the method to be used in higher-dimensional spaces efficiently.
@@
 structure to construct an approximation
-to each model's posterior parameter distribution.  This improved interpolation method leads to dramatically faster convergence of RJMCMC sample chains.
+to each model's posterior parameter distribution.  This improved interpolation method leads to faster convergence of RJMCMC sample chains.
 We draw jump proposals into the model from this approximation to its posterior.
@@
 jumps) and jumps to a different model with different parameters
-(intermodel jumps).  The resulting chain samples from the posterior
+(intermodel jumps).  The acceptance probability for a proposed jump
+from $\vtheta^{(j)}_i$ in model $M_i$ to $\vtheta^p_j$ in model $M_j$
+becomes
+\begin{equation}
+p_{\textnormal{accept}} \equiv \min\Bigl(1,
+\frac{p(\vtheta^p_j, M_j|d)}{p(\vtheta^{(j)}_i, M_i|d)}
+\frac{Q(\vtheta^p_j ,M_j \to
+  \vtheta^{(j)}_i, M_i)}{Q(\vtheta^{(j)}_i, M_i \to \vtheta^p_j, M_j)} \Bigr).
+\end{equation}
+Here the $Q$ factors incorporate both a discrete probability on the
+model index, reflecting the probabilistic choice of which model to
+jump into, and also a continuous probability density on target model's
+parameter space.  For example,
+$Q(\vtheta^p_j ,M_j \to \vtheta^{(j)}_i, M_i)$ has a factor for the
+probability of proposing a jump to model $i$ when in model $j$, and is
+a density on the parameter space of $M_i$.  These densities cancel the
+corresponding densities in the ratio of posteriors, making the
+acceptance probability a parameterisation-independent scalar.  In the
+common special case where the two parameter spaces have equal
+dimension and the jump proposal is a diffeomorphism, $\phi$, between
+them,
+\begin{equation}
+  \vtheta^p_j = \phi\left( \vtheta^{(j)}_i \right),
+\end{equation}
+and therefore 
+\begin{equation}
+  \vtheta^{(j)}_i = \phi^{-1} \left( \vtheta^p_j \right),
+\end{equation}
+then the jump proposal ratio reduces to the Jacobian of the
+diffeomorphism:
+\begin{equation}
+  \frac{Q(\vtheta^p_j ,M_j \to
+  \vtheta^{(j)}_i, M_i)}{Q(\vtheta^{(j)}_i, M_i \to \vtheta^p_j, M_j)}
+= \left| \frac{\partial \phi}{\partial \vtheta^{(j)}_i} \right|.
+\end{equation}
+
+ The resulting chain samples from the posterior
 $p(M_i, \{\vtheta_i\}|d)$.  As in a usual MCMC, the PDF on the model as a
@@
 In this section, we demonstrate the efficiency of the kD-interpolated
-jump proposal on a toy model-comparison problem.  We draw $N = 100$
-simulated data points from a $N(0,1)$ Gaussian distribution, and then
-ask whether these data are better described by a model where they are
-Gaussian distributed with unknown mean $\mu$ and standard deviation
-$\sigma$
+jump proposal on a toy model-comparison problem.  The same algorithm
+has been used in real-world settings \cite{Farr2010} and, as discussed
+below, is the available in several forms as a software library for
+re-use by others.
+
+In the toy model of this section, we draw $N = 100$ simulated data
+points from a $N(0,1)$ Gaussian distribution, and then ask whether
+these data are better described by a model where they are Gaussian
+distributed with unknown mean $\mu$ and standard deviation $\sigma$
 %
@@
 rate of inter-model jumps, the interpolation method described above
-can dramatically improve the runtime of an RJMCMC.
+can improve the runtime of an RJMCMC.
 
@@
     aligned with the axes is the normal kD-tree box containing the
-    given samples, while the tighter box is the ``covariance cell'' aligned with the eigenvectors of the sample distribution.
-The modified kd-interpolated jump proposal draws a point from the intersection of the two boxes.
-}
+    given samples, while the tighter box is the ``covariance cell''
+    aligned with the eigenvectors of the sample distribution.  The
+    modified kd-interpolated jump proposal draws a point from the
+    intersection of the two boxes. Tightly-correlated posteriors in
+    parameter space such as this are typical of the gravitational-wave
+    parameter estimation problem described in \cite{Veitch:2014} and
+    in the text.  Without the modification to account for the
+    correlated samples, the kD neighbourhood of these points would
+    produce a very inefficient jump proposal, since most of the
+    bounding-box would contain empty (i.e.\ low-posterior) parameter
+    space.}
 \end{figure}
@@
 
-Figure \ref{fig:accratio} shows the acceptance ratio for the kD-tree jump
-proposal applied as part of a \texttt{LALInferenceMCMC} analysis.
-In this example, the kD-tree jump proposal is one of several jump proposals used during the MCMC, and the kD tree itself is updated with accepted samples whenever its jump proposal is called.
-In spite of the very low acceptance
-rate of the proposal, applying the kD-tree proposal to one out of
-20 proposed jumps improved the convergence time of the simulation
-by a factor of two compared to the standard suite of proposals because
-it is particularly efficient at producing ``mode-hopping'' jumps,
-which are difficult to produce with the other proposals in
-\texttt{LALInferenceMCMC}.
+Figure \ref{fig:accratio} shows the acceptance ratio for the kD-tree
+jump proposal applied as part of a \texttt{LALInferenceMCMC} analysis.
+In this example, the kD-tree jump proposal is one of several jump
+proposals used during the MCMC, and the kD tree itself is updated with
+accepted samples whenever its jump proposal is called. In spite of the
+very low acceptance rate of the proposal, applying the kD-tree
+proposal to one out of 20 proposed jumps improved the convergence
+time---defined as the average time to reach a specified number of
+independent samples from the posterior, thinning by the
+autocorrelation length as described in \cite{Veitch:2014}---of the
+simulation by a factor of two compared to the standard suite of
+proposals because it is particularly efficient at producing
+``mode-hopping'' jumps, which are difficult to produce with the other
+proposals in \texttt{LALInferenceMCMC}.
 
@@
   \end{center}
-  \caption{\label{fig:accratio} The cumulative parameter acceptance
+  \caption{\label{fig:accratio} The cumulative acceptance
     ratio for the modified kD-tree jump proposal used in the
     \texttt{LALInferenceMCMC} code as a function of the number of
-    steps (in hundreds). In spite of the small acceptance ratio, the
-    kD tree proposal, applied to one in 20 proposed jumps,
-    improved the convergence time of the sampler by a factor of two
+    steps (in hundreds). The simulation in question was a
+    nine-dimensional analysis of a simulated gravitational-wave signal
+    injected into synthetic data similar to that taken by the LIGO
+    and Virgo detectors \cite{AdvLIGO,AdvVirgo}.  The parameter space
+    includes the masses of the compact objects generating the
+    gravitational-wave, their location on the sky, distance, orbital
+    orientation, the time of signal arrival, and the orbital phase.
+    The posterior in this problem has a number of well-separated modes
+    in the parameter space which are difficult to jump between using
+    traditional jump proposals; in spite of the small acceptance ratio
+    of the kD proposal, when applied to one in twenty jumps proposed
+    in this simulation, it improved the convergence time of the sampler by a factor of two
     compared to using only the standard suite of proposals. 

### 2015-04-20 paper.tex
Final point in reponse to reviewer 1, point 2.

@@
 In order to use the kD-tree interpolation as a jump proposal in an
-MCMC, we must be able to quickly find the neighborhood associated with
-a given point to compute the jump probability (see
-Eq.~\ref{eq:p-accept}).  This can be accomplished in $\order{\log N}$
-time and constant space with the following algorithm, which is a
-standard binary tree search.  Given the point, $\vtheta_i$ and the
-tree, $T$:
+MCMC, we randomly select a point stored in the kD tree with equal
+probability, and propose from the associated neighbourhood.
+Therefore, we must be able to quickly find the neighborhood associated
+with a given point to compute the jump probability (see
+Eq.~\ref{eq:p-accept}).  We introduce an additional parameter into the
+neighbourhood search, $N_\mathrm{boxing}$, which describes the minimum
+number of points in a kD box used as the neighbourhood for a point;
+small $N_\mathrm{boxing}$ increases variance in the proposal but
+resolves finer-scale structure in the posterior.  The neighbourhood
+search can be accomplished in $\order{\log N}$ time and constant space
+with the following algorithm, which is a modified binary tree search.
+Given the randomly-chosen point, $\vtheta_i$, the tree, $T$, and
+$N_\mathrm{boxing}$:
 \begin{enumerate}
-\item If $T$ contains exactly one sample point, then its box is the
-  associated neighborhood.  Otherwise:
+\item If $T$ contains fewer than $2N_\mathrm{boxing}$ points, then
+  its box is the associated neighborhood.  Otherwise:
 \item The tree $T$ has two sub-trees.  If the point $\vtheta_i$ is
@@
 \end{enumerate}
+The returned box is used for the jump proposal by drawing uniformly
+from its interior, so the proposal density is 
+\begin{equation}
+  \label{eq:neighbourhood-jp}
+  Q(\vtheta \to \vtheta^p) = \frac{N_\mathrm{box}}{N V},
+\end{equation}
+where $N$ is the number of points in the tree, $N_\mathrm{box}$ is the
+number of points in the chosen kD box, and $V$ is the coordinate-space
+volume of the box.
 
@@
 It then draws the covariance cell and proposes a new point uniformly from the intersection of the covariance cell with the original kD-tree box.
-The jump probability for this jump proposal is given by
+The jump probability for this jump proposal is still (c.f.\ Eq.\
+\eqref{eq:neighbourhood-jp}) given by
 %

### 2015-04-20 Response.txt
Final point in reponse to reviewer 1, point 2.
Updates to include response to items from 5b...l from report by referee 2.

--- 2015-03-22 Response.txt
@@
 
-> I am familiar with an earlier draft of this paper from internal LIGO review, and I'm pleased to see that it > is finally being submitted for publication.
+>>> I am familiar with an earlier draft of this paper from internal LIGO review, and I'm pleased to see that it is finally being submitted for publication.
 
@@
 
-> One minor quibble that I have is that the language in the introduction makes it seem that the underlying > idea is different. Both methods "construct an approximation to each
-> modelÕs posterior parameter distribution". The difference is that the kDE tree approach is far superior to > the original sparse histogram approach.
+>>> One minor quibble that I have is that the language in the introduction makes it seem that the underlying idea is different.
+Both methods "construct an approximation to each model’s posterior parameter distribution".
+The difference is that the kDE tree approach is far superior to  the original sparse histogram approach.
 
@@
 
-> Other minor point is that the reference to parallel tempering does not point to the original paper on the > topic (Swendsen and Wang 1986).
+>>> Other minor point is that the reference to parallel tempering does not point to the original paper on the topic (Swendsen and Wang 1986).
 
@@
 >>>This paper addresses an interesting and important issue in Bayesian model selection: Developing
-efficient jump proposals in reversible-jump Markov-chain Monte Carlo (RJMCMC)
+efficient jump proposals in reversible-jump Markov-chain Monte Carlo (RJMCMC)
 algorithms to transition between the models under consideration (high rejection rates for
@@
 
-We address these issues below, and in the text where appropriate.
+We address these issues below, and in the text where appropriate.  We are very grateful to the referee for the extremely thorough reading of the text.
 
@@
 (Section 3b) is presented at length with equations, the new RJMCMC approach is
-skimmed over in a single sentence (no equations) as \We perform MCMC in the supermodel
+skimmed over in a single sentence (no equations) as “We perform MCMC in the supermodel
 parameter space just like a regular MCMC; we propose jumps to different parameters
@@
 different parameters (inter-model jumps)" (page 6)...
+
+We have added descriptive text at the bottom of page 5 and top of page 6 that expands on the RJMCMC approach.
+The text now makes clear that the jump proposal ratio is a ratio of densities in the two model spaces, and discusses the special case where the spaces are of equal dimension and the proposal is a diffeomorphism between them (i.e., not a probabilistic choice of jump point, but rather a fixed mapping), where the ratio reduces to the Jacobian factor to which the referee refers.
+We have not written the equation as given by the referee,
+
+p_accept = (p(theta')*Q(theta' -> theta))/(p(theta)*Q(theta -> theta')) * |J|
+
+because this form is not correct in our notation.  The factors Q *already incorporate the Jacobian* in our notation.  This point is now emphasised in the text.
+
+We now give an explicit formula for the kD tree proposal density (Eq.
+3.1 in the updated manuscript) with an expanded discussion of the proposal algorithm at the end of Section 3.
+We have also modified the discussion surrounding Eq.
+5.1 to account for the additional text now in Section 3.
 
@@
 
->>> 4. Claims are made of \dramatically improved convergence" (page 3), that \convergence
-is generally rapid" (page 4), and \the interpolation method described above can dramatically
+We have added text to the beginning of Section 4 pointing to the use of this algorithm in real-world applications, such as those of Farr et al (2011).
+We feel that to describe the (lengthy) analysis from that work would needlessly complicate the presentation of the algorithm in this paper, but hope that this makes it clear that the algorithm is, indeed, applicable to complicated model selection problems.
+We have publicly released the algorithm as a software library for any users interested in adopting it, and emphasize this point in the text.
+
+We have added text to Figures 3 and 4 describing the gravitational-wave data analysis problem and the associated parameter space briefly.
+Again, we feel that a complete description of this problem (available from Veitch et al (2015) and the other references in the text) is beyond the scope of the article, but we appreciate the request for more context in the explanation of this paper.
+We hope the additional text provides this context.
+
+>>> 4. Claims are made of “dramatically improved convergence" (page 3), that “convergence
+is generally rapid" (page 4), and “the interpolation method described above can dramatically
 improve the runtime of an RJMCMC" (page 10). While I believe the method
@@
 convergence are given. The example considered in Figure 4 claims to have improved
-convergence by a factor of two (not dramatic), but there is no discussion of what definition is applied for convergence|and as noted on page 6 (third paragraph), defining
-convergence for MCMC sampling is a tricky business. Even applying a reasonable
+convergence by a factor of two (not dramatic), but there is no discussion of what definition is applied for convergence|and as noted on page 6 (third paragraph), defining convergence for MCMC sampling is a tricky business.
+Even applying a reasonable
 working definition of convergence, given the random nature of MCMC sampling, averages
 over several MCMC runs are often required to meaningfully compare convergence
-rates -- was that done here? While the claims of \dramatic" improvements in convergence may be true in some cases, they don't appear to be properly established in the paper.
+rates -- was that done here? While the claims of “dramatic" improvements in convergence may be true in some cases, they don't appear to be properly established in the paper.
+
+We have removed the (perhaps over-wrought) adjective "dramatic" from our claims of convergence, and included text briefly describing the method of computing convergence times used and referring the reader for details to Veitch et al (2015).
 
@@
 
->>> 5b. Page 5, rst line after Section 2b heading: Should be j = 1; : : : not i = 1; : : :.
-(c) Page 5, third line after Section 2b heading: Should be \set of parameters" not
-\set of parameter".
-(d) Figure 1 would be clearer if parameter axes were added indicating the origin, with
+>>> 5b. Page 5, Frst line after Section 2b heading: Should be j = 1…; not i = 1…;
+
+Fixed.
+
+>>> 5c. Page 5, third line after Section 2b heading: Should be “set of parameters" not
+“set of parameter".
+
+Fixed.
+
+>>> 5d. Figure 1 would be clearer if parameter axes were added indicating the origin, with
 axes scales in terms of standard deviations in x and y.
-(e) The x axis of Figure 2 is labeled Nboxing while it should be Nbox.
-(f) Figure 3 caption: kd should be kD.
-(g) Equation (5.1): P should be Q (this is the proposal distribution). Also, V might
+
+After attempting to follow the referee’s suggestion, we came to the conclusion that it distracted from the clarity of the illustration of kD interpolation.
+
+>>> 5e. The x axis of Figure 2 is labeled Nboxing while it should be Nbox.
+
+Nboxing is the minimum number of samples per box used to evaluate the local density, while Nbox is the actual number of samples (can be different since the total number of samples is not an exact power of 2); see Section 3.
+Nbox is thus specific to a particular point, while Nboxing is a tuning parameter of the algorithm and is the parameter being varied in Fig.
+2.
+
+>>> 5f. Figure 3 caption: kd should be kD.
+
+Fixed.
+
+>>> 5g. Equation (5.1): P should be Q (this is the proposal distribution). Also, V might
 be better written as Vbox to indicate that it is the volume of the current box, like
@@
 varies with boxes).
-(h) Page 11: Inconsistent use of math or Roman font for some subscripts; e.g., Nbox
+
+Fixed here and in subsequent paragraph.  We kept V, since we define it explicitly just after this equation (and it has a slightly different meaning here then in (3.1).
+
+>>> 5h. Page 11: Inconsistent use of math or Roman font for some subscripts; e.g., Nbox
 and Nbox, and Ncrit. When the subscript is a label rather than an index, Roman
 font is preferred.
-(i) Page 11, after Eq. (5.1): Please give an equation for V , the volume of the intersection
+
+Fixed.
+
+>>> 5i. Page 11, after Eq. (5.1): Please give an equation for V , the volume of the intersection
 of the two (potentially high-dimensional) boxes.
-(j) Page 12, first paragraph discusses violations of detailed balance arising from using
+
+We have omitted this as the expression is somewhat tedious without adding any insight.
+
+>>> 5j. Page 12, first paragraph discusses violations of detailed balance arising from using
 the sampling history to define proposals. Such methods are well known, including
@@
 is not a new idea.
-(k) Figure 4's caption states \The acceptance probability rate asymptotes to the
-steady-state solution once sufficient samples are accumulated for the kD tree
+
+We did not wish to imply that this is a new idea; suggested reference added.
+
+>>> 5k. Figure 4's caption states “The acceptance probability rate asymptotes to the
+steady-state solution once sufficient samples are accumulated for the kD tree
 to allow the sample density or be accurately interpolated." This asymptotic
-convergence appears to occur by about step 2.0106; hence, all samples collected
+convergence appears to occur by about step 2.0 * 10^6; hence, all samples collected
 prior to this point should be discarded as burn-in or the sampling is potentially
 biased. This should be mentioned explicitly in the paper.
-(l) Figure 4: Why denote the x axis divisions from 0 to 25 000 and then include
+
+It is now clarified:
+“The acceptance rate asymptotes to the steady-state solution once sufficient samples have been accumulated in the kD tree to allow the sample density to be accurately interpolated; samples collected
+prior to this point should be discarded as burn-in.”
+
+>>> 5l. Figure 4: Why denote the x axis divisions from 0 to 25 000 and then include
 \( X100)" in the axis label? Why not denote the divisions from 0 to 2.5 and
@@
 
+The referee’s suggestion would, indeed, be more standard, but the tools we used to process the data made it more straightforward to generate the plot as given, so we hope the slightly odd axis label is not too jarring.
+
 ----

### 2015-04-20 paper.tex
Updates to include response to items from 5b...l from report by referee 2.

@@
 \newcommand{\Nbox}{N_\mathrm{box}}
+\newcommand{\Ncrit}{N_\mathrm{crit}}
 \newcommand{\Vbox}{V_\mathrm{box}}
@@
 A Markov chain Monte Carlo \cite{Gilks:1996} produces a set of samples
-$\{ \vtheta^{(j)} \, | \, i = 1, \ldots \}$ from the model parameter
+$\{ \vtheta^{(j)} \, | \, j = 1, \ldots \}$ from the model parameter
 space that are sampled according to the posterior, meaning that, in
 the limit that the chain length tends to infinity, the relative
-frequency with which a given set of parameter appears in the chain is
+frequency with which a given set of parameters appears in the chain is
 proportional to the desired posterior, $p(\vtheta|d,M)$.  Therefore,
@@
 A better proposal, therefore, would account for larger-scale sample covariances in parameter space.
-To address significant covariance between samples, which would correspond to strong density gradients and a very non-uniform density profile within a rectangular box, the modified jump proposal does not descend to the leaf-nodes of the tree, but instead stops descending whenever the number of samples in the current box, $N_{box}$, falls below some threshold, $N_{crit}$.
+To address significant covariance between samples, which would correspond to strong density gradients and a very non-uniform density profile within a rectangular box, the modified jump proposal does not descend to the leaf-nodes of the tree, but instead stops descending whenever the number of samples in the current box, $\Nbox$, falls below some threshold, $\Ncrit$.
 We then find the principal axes of the samples
@@
 We use these eigenvectors to draw a new box.
-This box -- henceforth called a covariance cell -- contains all $N_{box}$ samples from the kD-tree box, is centered on the mean of these samples, and has edges that are aligned with the principal axes.
+This box -- henceforth called a covariance cell -- contains all $\Nbox$ samples from the kD-tree box, is centered on the mean of these samples, and has edges that are aligned with the principal axes.
 Furthermore, this box is drawn around the samples as tightly as possible; in other words, for each edge, there exists a point in the covariance cell that lies along that edge.
@@
     aligned with the eigenvectors of the sample distribution.  The
-    modified kd-interpolated jump proposal draws a point from the
+    modified kD-interpolated jump proposal draws a point from the
     intersection of the two boxes. Tightly-correlated posteriors in
     parameter space such as this are typical of the gravitational-wave
-    parameter estimation problem described in \cite{Veitch:2014} and
+    parameter estimation problem described in \protect\cite{Veitch:2014} and
     in the text.  Without the modification to account for the
@@
     produce a very inefficient jump proposal, since most of the
-    bounding-box would contain empty (i.e.\ low-posterior) parameter
+    bounding-box would contain empty (i.e., low-posterior) parameter
     space.}
@@
 
-The jump proposal chooses a stored sample at random and finds the largest kD-tree box containing it and fewer than $N_{crit}$ total samples.
+The jump proposal chooses a stored sample at random and finds the largest kD-tree box containing it and fewer than $\Ncrit$ total samples.
 It then draws the covariance cell and proposes a new point uniformly from the intersection of the covariance cell with the original kD-tree box.
@@
 %
-\bel{modforward} P(\vtheta_i \rightarrow \vtheta_{i+1}) =
+\bel{modforward} Q(\vtheta_i \rightarrow \vtheta_{i+1}) =
 \frac{\Nbox}{N V} , \ee
@@
 This technique trivially violates detailed balance since the jump proposal distribution depends on the past history.
-One way to address this problem requires continuously accumulating samples into the kD tree as the MCMC explores more
+One way to address this problem, through diminishing adapations \cite{Brooks:2011}, requires continuously accumulating samples into the kD tree as the MCMC explores more
 of the parameter space.
 As the number of samples drawn from the posterior PDF increases, the neighborhoods around each sample decrease with increasing sample density and the calculated covariances between samples in the kD tree better approximate the true covariances between parameters in the equilibrium posterior PDF.
-Hence, for large $N$, the change in $P(\vtheta^{(j)} \rightarrow \vtheta^{(j+1)})$ approaches zero and the jump proposal does a progressively better job of sampling the equilibrium posterior while becoming asymptotically Markovian.
+Hence, for large $N$, the change in $Q(\vtheta^{(j)} \rightarrow \vtheta^{(j+1)})$ approaches zero and the jump proposal does a progressively better job of sampling the equilibrium posterior while becoming asymptotically Markovian.
 Conservatively, the entire set of samples obtained while the jump proposal is being dynamically updated could be discarded as burn-in, and future analysis could use the static jump proposal distribution interpolated from samples accumulated during the extended burn-in phase.
@@
     injected into synthetic data similar to that taken by the LIGO
-    and Virgo detectors \cite{AdvLIGO,AdvVirgo}.  The parameter space
+    and Virgo detectors \protect\cite{AdvLIGO,AdvVirgo}.  The parameter space
     includes the masses of the compact objects generating the
@@
     compared to using only the standard suite of proposals. 
-    The acceptance rate asymptotes to the steady-state solution once sufficient samples have been accumulated in the kD tree to allow the sample density to be accurately interpolated.
-%\dan{[Taking a leap here, someone please confirm/deny/add qualifiers] The initial decrease in acceptances arises from insufficient population of the kD tree with too few samples, while the leveling-off later in the MCMC indicates convergence.} \ilya{[I see structure in this figure (a high acceptance probability, followed by a dip, followed by an asymptote).
-Is this a fluke, or do you always see this -- and, if so, why?
-BTW, what's the autocorrelation length (just to give me a sense of scale, so I am not reading too much into meaningless fluctuations within an AC length)?]} \dan{[Don't recall the autocorrelation length.
-Will?]}
+    The acceptance rate asymptotes to the steady-state solution once sufficient samples have been accumulated in the kD tree to allow the sample density to be accurately interpolated; samples collected
+prior to this point should be discarded as burn-in.
   }
